%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
\documentclass[acmsmall,review,anonymous,dvipsnames]{acmart}
\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For double-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For single-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission, w/ required CCS and ACM Reference
%\documentclass[acmsmall]{acmart}\settopmatter{}

\let\Bbbk\relax % polycode defines that again?
%include lhs2TeX.fmt
%include lhs2TeX.sty
%include polycode.fmt
%include ../calculus.fmt


\RequirePackage[utf8]{inputenc}

%% Journal information
%% Supplied to authors by publisher for camera-ready submission;
%% use defaults for review submission.
\acmJournal{PACMPL}
\acmVolume{1}
\acmNumber{CONF} % CONF = POPL or ICFP or OOPSLA
\acmArticle{1}
\acmYear{2018}
\acmMonth{1}
\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}

%% -------------- Copyright information ------------------
%% Supplied to authors (based on authors' rights management selection;
%% see authors.acm.org) by publisher for camera-ready submission;
%% use 'none' for review submission.
\setcopyright{none}
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\copyrightyear{2018}           %% If different from \acmYear

%% ------------- Bibliography style -------------------
\bibliographystyle{ACM-Reference-Format}
\DeclareUnicodeCharacter{1D706}{$\lambda$}
\DeclareUnicodeCharacter{209B}{${}_s$}
%% Citation style
%% Note: author/year citations are required for papers published as an
%% issue of PACMPL.
\citestyle{acmauthoryear}   %% For author/year citations



%% Some recommended packages.
\usepackage{booktabs}   %% For formal tables:
                        %% http://ctan.org/pkg/booktabs
\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
                        %% http://ctan.org/pkg/subcaption

\usepackage{mathpartir}

\usepackage{stmaryrd}
\usepackage{colortbl}
\usepackage{bussproofs}
\usepackage{mathtools}
% \usepackage{amsmath}
\usepackage{verbatim}
\usepackage[capitalize]{cleveref}   % \cref and \Cref
\usepackage{textgreek} % \textalpha etc.

% This makes a tt font that is slightly smaller.
% \usepackage[scaled=0.8]{beramono}

% This incantation makes tt be Latin Modern Mono Light Cond
% which is narrower than normal tt.
%\usepackage{lmodern}
%\DeclareFontFamily{T1}{lmttc}{\hyphenchar \font-1 }
%\DeclareFontShape{T1}{lmttc}{m}{n}
%     {<-> ec-lmtlc10}{}
%\DeclareFontShape{T1}{lmttc}{m}{it}
%     {<->sub*lmttc/m/sl}{}
%\DeclareFontShape{T1}{lmttc}{m}{sl}
%     {<-> ec-lmtlco10}{}
%\renewcommand{\ttdefault}{lmttc}

% ---------------- Boxed figures ------------------
\usepackage{float}
\floatstyle{boxed}
\restylefloat{figure}

% ---------------- Full Verse ----------------
\newif\iffullverse
\fullversefalse
% use \iffullverse \else \fi

% ---------------- Useful macros ----------------
\renewcommand{\arraystretch}{1.1}

% Name of the language
\newcommand{\versecalc}{$\mathcal{VC}$}

% Typesetting rewrite rules
\newcommand{\rulename}[1]{\textsc{\small #1}}
\newcommand{\vb}{\; |vbar| \;}    % Vertical bar
\newcommand{\hole}{\square}
\newcommand{\freevars}[1]{\mathsf{fvs}(#1)}
\newcommand{\boundvars}[1]{\mathsf{bvs}(#1)}
\newcommand{\valvars}[1]{\mathsf{vvs}(#1)}
\newcommand{\disjoint}{\;\#\;}
\newcommand{\hnf}{\mathit{hnf}}  % Head normal form
\newcommand{\subst}[3]{#1\{#3/#2\}}


% Right arrow for rewrites
\newcommand{\movesto}{\longrightarrow}
\newcommand{\movesfrom}{\longleftarrow}
\newcommand{\convertible}{\longleftrightarrow^{\ast}}

% ----------------- Author comments ---------------------
\definecolor{dkcyan}{rgb}{0.1, 0.3, 0.3}
\definecolor{dkgreen}{rgb}{0,0.3,0}
\definecolor{olive}{rgb}{0.5, 0.5, 0.0}
\definecolor{dkblue}{rgb}{0,0.1,0.5}

\iftrue  % iftrue to enable, iffalse to disable
\newcommand{\simon}[1]{{\color{dkblue} [{SLPJ}: #1]}}
\newcommand{\lennart}[1]{{\color{dkcyan} [{LA}: #1]}}
\newcommand{\joachim}[1]{{\color{dkblue} [{JB}: #1]}}
\newcommand{\koen}[1]{{\color{dkred} [{KC}: #1]}}
\newcommand{\olin}[1]{{\color{dkgreen} [{OS}: #1]}}
\else
\newcommand{\simon}[1]{}
\newcommand{\lennart}[1]{}
\newcommand{\joachim}[1]{}
\newcommand{\koen}[1]{}
\newcommand{\olin}[1]{}
\fi
% ---------------- End of comments -----------------





\begin{document}

%% Title information
\title{The essence of functional logic progamming}

                                        %% [Short Title] is optional;
                                        %% when present, will be used in
                                        %% header instead of Full Title.
% \titlenote{with title note}             %% \titlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'
% \subtitle{Subtitle}                     %% \subtitle is optional
% \subtitlenote{with subtitle note}       %% \subtitlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'


%% Author information
%% Contents and number of authors suppressed with 'anonymous'.
%% Each author should be introduced by \author, followed by
%% \authornote (optional), \orcid (optional), \affiliation, and
%% \email.
%% An author may have multiple affiliations and/or emails; repeat the
%% appropriate command.
%% Many elements are not rendered, but should be provided for metadata
%% extraction tools.

%% Author with single affiliation.
\author{Lennart Augustsson}
% \authornote{with author1 note}          %% \authornote is optional;
%                                         %% can be repeated if necessary
\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
  \position{Position1}
  \department{Epic Games}              %% \department is recommended
  \institution{Institution1}            %% \institution is required
  \streetaddress{Street1 Address1}
  \city{City1}
  \state{State1}
  \postcode{Post-Code1}
  \country{Country1}                    %% \country is recommended
}
\email{lennart.augustsson@@epicgames.com}          %% \email is recommended

\author{Simon Peyton Jones}
% \authornote{with author1 note}          %% \authornote is optional;
%                                         %% can be repeated if necessary
\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
  \position{Position1}
  \institution{Epic Games}            %% \institution is required
  \streetaddress{Street1 Address1}
  \city{City1}
  \state{State1}
  \postcode{Post-Code1}
  \country{Country1}                    %% \country is recommended
}
\email{simonpj@@epicgames.com}          %% \email is recommended

\author{Joachim Breitner}
% \authornote{with author1 note}          %% \authornote is optional;
%                                         %% can be repeated if necessary
\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
  \position{Position1}
  \institution{Epic Games}            %% \institution is required
  \streetaddress{Street1 Address1}
  \city{City1}
  \state{State1}
  \postcode{Post-Code1}
  \country{Country1}                    %% \country is recommended
}
\email{joachim.breitner@@epicgames.com}          %% \email is recommended




%% ---------------- Abstract ---------------------
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
  The abstract
\end{abstract}


%% ---------------- Classification ---------------------
%% 2012 ACM Computing Classification System (CSS) concepts
%% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10011007.10011006.10011008</concept_id>
<concept_desc>Software and its engineering~General programming languages</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10003456.10003457.10003521.10003525</concept_id>
<concept_desc>Social and professional topics~History of programming languages</concept_desc>
<concept_significance>300</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Software and its engineering~General programming languages}
\ccsdesc[300]{Social and professional topics~History of programming languages}
%% End of generated code


%% Keywords
%% comma separated list
% \keywords{keyword1, keyword2, keyword3}  %% \keywords are mandatory in final camera-ready submission


%% \maketitle
%% Note: \maketitle command must come after title commands, author
%% commands, abstract environment, Computing Classification System
%% environment and commands, and keywords command.
\maketitle

\textbf{Note} \textit{This is a working draft.  It's not secret, but it's very much
  a work in progress.  So please don't share it; if there is anyone you think would be
  interested/helpful, do suggest them to us.}
\section{Introduction}

Functional logic programming has a long history and a rich literature.
But it is somewhat tricky for programmers to reason about functional logic programs: they must think about
logical variables, narrowing, backtracking, Horn clauses, resolution,
and the like.   This contrasts with functional programming,
where one can say ``just apply rewrite rules, such as beta reduction, let-inlining, and case-of-known-constructor''.
Moreover, these rewrite rules apply even in the presence of free variables; that is, under lambdas.

We therefore seek a completely-precise expression of functional logic programming, as
at term rewriting system, to give us both a formal semantics (via a small-step reductions),
and a powerful set of equvalences that programmers can use to reason about their programs,
and compilers can use to optimise them.

Our main contribution is a new core calculus, the Verse calculus or \versecalc{} for short,
for functional logic programming that
has the following properties:
\begin{itemize}
\item It natively supports \emph{higher-order programming}, just like the lambda calculus.  In contrast,
  the functional-logic literature typically addresses higher order features via defunctionalisation \cite[3.3]{hanus:survey}.
\item It supports \emph{logical variables}. In the expression |def x e|, the variable |x| is brought into scope,
  but gains its value through \emph{equalities} |(e_1 = e_2)| in |e|.  Like any declarative language, however,
  an existential variable always stands for a single value; logical variables are not mutable.
\item It supports so-called \emph{non-deterministic programming}; that is, an expression can return one result, many results, or no result (failure).
  As a result, it natively supports both narrowing and residuation.
\item \versecalc{} has a \emph{small-step term rewriting semantics} (\Cref{sec:rewrite-rules}).
  Just like rewriting for the lambda calculus, this semantics allows programmers to reason, step by step,
  about how their program executes.  This set of rewrite rules is our main contribution.
\item \simon{Something about confluence.}
\end{itemize}
As always with a calculus, the idea is that \versecalc{} distils the essence of functional logic programming.
Each construct does just one thing, and \versecalc{} cannot be made smaller without losing key features.
It should be possible to compile a variety of functional logic languages,
including Curry \cite{hanus:curry-report}, into \versecalc{}.
We are also working on Verse, a new programming language for the metaverse, built directly on \versecalc{}; indeed,
our motivation for developing \versecalc{} is practical rather than theoretical.

There is a rich seam of related work; an excellent starting point is Hanus's survey \cite{hanus:survey}.
We discuss related work in \cref{sec:related}.

\lennart{I think we should mention that this work is related to an actual programming language.
  This will also motivate the name of the calculus.}

\section{The Verse calculus, informally} \label{sec:informal}

\begin{figure}
  $$
  \begin{array}{l}
    \textbf{Syntax} \\
    \begin{array}{lrcl}
    \text{Integer constants} & k \\
    \text{Variables} & \multicolumn{3}{l}{x,y,z,f,g} \\
    \text{Programs} & p  & ::=  & |one e| \quad \text{where} \; \freevars{e} = \emptyset \\
    \text{Expressions} & e,r,s  & ::=  & |v| \vb |apply1N v_1 v_2|
       \vb |e_1 == e_2| \vb |e_1 ; e_2| \vb  |def x e| \\
    && \vb & |e_1 choice e_2| \vb |fail| \vb |one e| \vb |all e| \\
% SPJ Omitting for now
\iffullverse
    && \vb & |succeeds e| \vb |decides e| \\
\fi
    \text{Values} & v & ::= & x  \vb \hnf \\
    \text{Head values} & \hnf & ::= & k \vb op \vb |arr (v_1, dots, v_n)| \vb |lam x e| \\
    \text{Primops} & op  & ::= & |gt| \vb |add|
\iffullverse
    \vb |isInt|
\fi
    \end{array} \\
    \\
    \textbf{Contexts} \\
    \begin{array}{lrcl}
    \text{Choice-free exprs} & ce  & ::=  & |v| \vb |ce_1 == ce_2| \vb |ce_1 ; ce_2| 
    \vb |one e| \vb |all e| \\
\iffullverse
    && \vb & |succeeds e| \vb |decides e| \\
\fi
    \text{Execution contexts} & X & ::= & \hole \vb |X == e| \vb |e == X| \vb |X ; e| \vb |e; X| \\
    \text{Choice contexts} & CX & ::= & \hole \vb |CX == e| \vb |ce == CX| \vb |CX ; e| \vb |ce; CX| \vb |def x CX| \\
    \text{Scope contexts} & SX & ::= & |hole choice e| \vb |e choice hole |
    \vb |one hole| \vb |all hole| \\
\iffullverse
    && \vb & |succeeds hole| \vb |decides hole| \\
\fi
    \text{Value context} & V & ::= & \hole \vb |arr (v_1, dots ,V, dots, v_n)|\\
    \end{array} \\
    \\
    \textbf{Shorthand} \\
    \begin{array}{rcl}
%%      |emptytup| & \text{means} & |arr ()|\\
%%      |tup (v_1, dots, v_n)| & \text{means} & |arr (v_1,dots, v_n)| \quad n \geq 2 \\
      v_1 + v_2 & \text{means} & |applyN add ( v_1, v_2 ) | \\
      v_1 > v_2 & \text{means} & |applyN gt ( v_1, v_2 ) | \\
      |def (x_1,x_2,dots,x_n) e| & \text{means} & |def x_1 (def x_2 (dots ^ def x_n e))| \\
     \end{array} \\
  \\
  \text{The notation $\freevars{e}$ means the free variable of $e$; in \versecalc{}, $\lambda$ and |defKW| are the only binders.} \\
  \text{We say $S_1 \disjoint S_2$ to mean that set $S_1$ is disjoint from $S_2$.}
  \end{array}
  $$
\caption{The Verse calculus: syntax} \label{fig:syntax}
\end{figure}


We begin by presenting the Verse calculus, \versecalc, informally.  We will give its
rewrite rules precisely in \cref{sec:rewrite-rules}.
\versecalc{} is modelled on the lambda calculus, and indeed the lambda calculus is
a sub-language of \versecalc{}.  \emph{Every lambda calculus term is a \versecalc{} term},
and has the same semantics. Just like the lambda calculus, \versecalc{} is untyped;
adding a type system is an excellent goal, but is the subject of another paper.

The syntax of \versecalc{} is given in \cref{fig:syntax}.
It has a very conventional sub-language that is just the lambda-calculus with
some built-in operations and tuples as data constructors:
\begin{itemize}
\item \emph{Head normal forms}.
  A head normal form, |hnf|, is a conventional value: a built-in constant |k| or operator |op|, a tuple,
  also called array, |arr (v_1,dots,v_n)|, or a lambda term |lam x e|.
For the built-in constants and operators we use only integer constants, and two representative
functions |gt| and |add|.
\item \emph{Values.}
In a functional logic language an expression may evaluate to an as-yet-unknown logical variable,
so a \emph{value} |v| is either a head normal form, or a variable |x|.
\item \emph{Expressions} |e| includes values |v|, and applications |apply1N v_1 v_2|.\footnote{
For clarity we with usually write |apply1 v_1 v_2| rather than |apply1N v_1 v_2| when |v_2| is not an array.
}

  Remembering that a value can be a lambda, that gives us the full lambda calculus.
  \item \emph{A program}, |p|, is a closed expression from which we extract one result using |oneKW| (see \cref{sec:one}).
\end{itemize}
Expressions also include two other key collections of constructs:
logical variables and unification (\cref{sec:unify}), and choice (\cref{sec:choice}).
The details of choice and unification, and especially their
interaction, are rather tricky, so this section will do a lot of
arm-waving.  But fear not: \cref{sec:rewrite-rules} will make all this
precise.

\subsection{Logical variables and unification} \label{sec:unify}

The Verse calculus includes first class logical variables and unification:
you can bring a fresh logical variable into scope with |defKW|; equate two expressions with |e_1 = e_2|; and
sequence two expressions with |e_1 ; e_2| (see \cref{fig:syntax}).
As an example, what might be written |dlet x e_1 e_2| in a conventional functional language can be written
|def x (x=e_1; e_2)| in \versecalc{}\footnote{
  Just as with |lam x e|, where the $\lambda$ scopes as far to the right as possible,
  in an expression |def x e| the |defKW| scopes as far to the right as possible,
}.

A program executes by solving its equations, to get values for some of the logical variables.
For example
$$
|def (x,y,z) (x=tup(y,3); x=tup(2,z); y)|
$$
is solved by unifying |x| with |tup(y,3)| and with |tup(2,z)|; that in turn unifies |tup(y,3)| with |tup(2,z)|, which
unifies |y| with |2| and |z| with |3|.  Finally |y| is returned as the result.

High level functional languages usually provide some kind of
pattern-matching; in such a langauge, we might write define |fst| by
|applyN fst (a,b) = a|. Such pattern-matching is typically desugared to
more primitive |case| expressions.  But in \versecalc{} we don't need |case| expressions:
unification does the job.  For example we can define |fst| like this:
$$
|fst = lam pr (def (a,b) (pr = tup (a,b); a))|
$$
The lambda unifies its argument (presumably a pair)
with |(a,b)|, where |a,b| are fresh logical
variables in the body of the lambda; and returns the first of the two, namely |a|.
For convenience, in this presentation we allow ourselves to write a term like
|applyN fst (2,5))|, where we define |fst| separately. Formally, you can imagine
each example |e| being wrapped with a binding for |fst|, thus |def fst (fst = ...); e|;
and similarly for all other library functions.

This way of desugaring pattern matching means that the input
to |fst| may include logical variables. For example:
$$
|def (x,y,z) (x=tup(y,z); apply1 fst x = 2; y)|
$$
Here |apply1 fst x| evaluates to |y|, which we then unify with |2|; the logical
variable |z| never gets a value, but that does not matter.
Another way to say this is that, as usual in logic programming, we may
constraint the \emph{output} of a function (here |apply1 fst x = 2|), and thereby
affect its \emph{input} (here |tup(y,z)|).

Although ``|;|'' is called ``sequencing'', the order of that sequence
is immaterial for equations.
For example\footnote{We informally write |3+y| rather than
  |applyN add (3,y)|, and similarly with other operators.},
consider |def x (x = 3+y; y = 7; x)|.  In \versecalc{} we can only unify |x| with a \emph{value};
we will see why in \cref{sec:call-time-choice}. So the equation |x=3+y| is stuck.
No matter!  We simply leave it and try some other equation. In this case, we can make progress with |y=7|;
and that in turn unlocks |x = 3+y| because now we know that |y| is |7|, so we can evaluate |3+7| to |10|
and unify |x| with that.  The idea of leaving stuck expressions aside, and executing other parts of the
program is called \emph{residuation} \cite{hanus:survey}, and is at the heart of our mantra ``just solve the equations''.

\simon{Why are applications |apply1 v1 v2| not |apply1 e1 e2|?}

\subsection{Choice} \label{sec:choice} \label{sec:call-time-choice}

In conventional functional programming, an expression evaluates to a single value.
In contrast, a \versecalc{} expression (unless it gets stuck) evaluates to a choice of zero, one, or many values.
The expression |fail| yields no values; a head normal form |hnf| yields one value;
and the choice |e_1 choice e_2| yields all the values
yielded by |e_1| and all the values yielded by |e_2|.
Duplicates are not eliminated and, as we shall see in \cref{sec:for}, order is maintained; in short
an expression yields a sequences of values, not a bag, and certainly not a set.

The equations we saw in \cref{sec:unify} can fail, if the arguments
are not equal, yielding no results.  Thus |3=3| succeeds, returns a single result, namely |3|,
while |3=4| fails, returning no results.  In general, we use ``fail''
and ``returns no results'' synonymously.

What if the choice was not at the top-level of an expression?  For example,
what does |tup (3, (7 choice 5))| mean?
In \versecalc{} it does \emph{not} mean a pair with some kind of multi-value in its second component.
Indeed, as you can see from \cref{fig:syntax}, this expression is syntactically ill-formed.
We must instead give a name to that choice, and then we can put it in the pair, thus: |def x (x = (7 choice 5); tup (3,x))|.
This is syntactically legal, but what does it mean?  In \versecalc{} a variable is never bound to a multi-value.
Instead, |x| is successively bound to |7|, and then to |5|, like this:
$$
|def x (x = (7 choice 5); tup (3,x))| \quad = \quad |(def x (x=7; tup (3,x))) choice (def x (x=5; tup (3,x))))|
$$
We duplicate the context surrounding the choice, and ``float the choice outwards''.

\subsection{Mixing choice and unification} \label{sec:mixing}

We saw in \cref{sec:unify} that \emph{equations} are insensitive to sequencing ---
but \emph{choice} is not.
Consider
|def (x,y) (x = (3 choice 4); y = (20 choice 30); tup (x,y))|.  The choices are made \emph{left-to-right}, so that the result is
|((tup (3,20)) choice (tup (3,30)) choice (tup (4,20)) choice (tup (4,30)))|.

So much for choice under unification.  What if we have
unification under choice?  For exapmle:
$$
|def x ((x=3; x+1) choice (x=4; x*2))|
$$
Intuitively, either unify |x| with |3| and return |x+1|, or unify |x| with |4| and return |x*2|.
But so far we have said only ``a program executes by solving its equations'' (\cref{sec:unify}).
Well, we can see two equations here, |(x=3)| and |(x=4)|, which are mutually contradictory,
so clearly we need to refine our notion of ``solving''.
The answer is pretty clear: in a branch of a choice, solve the equations in that branch
to get the value for some logical variables; \emph{and propagate those values to occurrences in
  that branch (only)}. Occurrences of that variable outside the choice are unaffected.  We call
this \emph{local propagation}.
This local-propagation rule would allow us to reason thus:
$$
|def x ((x=3; x+1) choice (x=4; x*2))| \movesto  |def x ((x=3; 4) choice (x=4; 8))|
$$
Are we stuck now?  No, we can float the choice out as
before\footnote{Indeed we could have done so first, had we wished.}, thus
$$
|def x ((x=3; 4) choice (x=4; 8))| \movesto |(def x (x=3; 4)) choice (def x (x=4; 8))| 
$$
and now it is apparent that the sole occurrence of |x| in each |defKW| is the equation
|(x=3)|, or |(x=4)| respectively; so we can drop the |defKW| and the equation, yielding |(4 choice 8)|.
% To take a celebrated example, suppose
% |square = lam x (x*x)|.  What does |apply1 square (3 choice 4)| mean? If we make the choice \emph{before}
% the call, the call will yield |9 choice 16|.  But if we a more call-by-name semantics, we might imagine
% it yields |(3 choice 4) * (3 choice 4)| which in turn yields |(9 choice 12 choice 12 choice 16)|.
% All functional-logic languages, including 

\subsection{Pattern matching and narrowing}

We remarked in \Cref{sec:unify} that we can desugar the pattern matching of a high-level
languag into unification. But what about pattern matching with multiple equations, such as this definition in Haskell:
\begin{code}
  append []      ys = ys
  append (x:xs)  ys = x : append xs ys
\end{code}
If pattern matching on the first equation fails, we want to fall through to the second.
Fortunately, choice allows us to express this idea directly\footnote{
We use the empty tuple |emptytup| to represent the empty list and pairs to represent |cons|;
and we allow ourselves to write |lamhdr (tup(x,y)) body| rather than |lam p (def (x,y) (p = tup (x,y); body)|
  }:
$$
|append =lamhdr (tup (xs,ys))| \begin{array}[t]{l}
  |(xs = emptytup; ys) choice| \\
  |(def (a,as) (xs = tup (a,as); tup (a, applyN append (as, ys))))|
\end{array}
$$
If |xs| is |emptytup|, the left hand choice succeeds, returning |y|; and the right hand choice
fails (by attempting to unify |emptytup| with |tup(a,as)|).
If |xs| is of form |tup (y,ys)|,
the right hand choice succeeds, and we make a recursive call to |append|.
Finally if |xs| is built with head normal forms other than the empty tuple and pairs, both choices fail, and |append| returns no
results.

This approach to pattern matching is akin to \emph{narrowing} \cite{hanus:survey}.
Suppose |single = tup (1, emptytup)|, a singleton list whose only element is 1.
Consider this call:
$$
|def xs (applyN append ( xs, single) = single; xs)|
$$
The call to |append| expands into a choice
$$
  |(xs = emptytup; single)| \quad |choice| \quad
  |(def (a,as) (xs = tup (a,as); tup(a, applyN append (as, single))))|
$$
which amounts to exploring the possiblity that |xs| is headed by |emptytup| or a pair -- the
essence of narrowing.  It should not take long to reassure yourself that the program
evaluates to |emptytup|, effectively running |append| backwards in the classic logic-programming manner.

This example also illustrates that \versecalc{} allows an equality (for |append|)
that is recursive. As in any functional langauge with recursive bindings, you can
go into an infinite loop if you keep fruitlessly inlining the function in its own right hand side.
It is the busines of an \emph{evaluation} strategy to do only rewrites that make
progress towards a solution.

\subsection{Conditionals and |oneKW|} \label{sec:rigid} \label{sec:one} \label{sec:if}

Every source language will provide a conditional, like |if (x=0) then
e_2 else e_3|.  But what is the equality operator in |(x=0)|?  One
possiblity, adopted by Curry, is this: there is one ``='' for
equations (as in \cref{sec:unify}), and another, say ``=='', for
testing equality (returning a boolean with constructors |True| and |False|).  \versecalc{} takes a
different, more minimalist position. In \versecalc{} \emph{there is just one equality operator}, written ``='' just as in
\cref{sec:unify}. The expression |if (x=0) then e_2 else e_3| tries to unify |x| with |0|.  If that succeeds (returns one or more values)
the |if| returns |e_2|; otherwise it returns |e_3|.  There are no data constructors |True| and |False|; instead
failure plays the role of falsity.

But something is terribly wrong here.  Consider |def (x,y) (y = (if (x=0) 3 4); x=7)|.  Presumably this is meant
to set |x| to |7|, test if it is equal to |0| (it is not), and unify |y| with 4.
But what is to stop us instead unifying |x| with |0| (via |(x=0)|), unifying |y| with |3|, and then
failing when we try to unify |x| with |7|?  Not only is that not what we intended, but it also looks very
non-deterministic: the result is affected by the order in which we did unifications!

To address this, we give |if| a special property: in the expression |if e_1 then e_2 else e_3|, unifications inside |e_1| (the condition of the |if|)
can only unify variables |defKW|'d inside |e_1|; variables bound outside |e_1| are ``rigid''.  So in our example, the |x| in |(x=0)| is
rigid and cannot be unified.  Instead the |if| is stuck, and we move on to unify |x=7|.  That unblocks
the |if| and all is well.  This special property is precisely the local propagation rule that we sketched for
choice (\cref{sec:mixing}).

In fact, \versecalc{} distills the three-argument |if| into something simpler, the unary construct |one(e)|.
Its specification is this:
if |e| fails, |one(e)| fails; otherwise |one(e)| returns the first of the values yielded by |e|.
Now, |if e_1 then e_2 else e_3| can (nearly) be re-expressed like this:
$$
|one( (e_1; e_2) choice e_3 )|
$$
If |e_1| fails, the first branch of the choice fails, so we get |e_3|; if |e_1| succeeds, we
get |e_2|, and the outer |oneKW| will select it from the choice.  But this isn't right:
what if |e_2| or |e_3| themselves fail or return multiple results?  Here is a better translation,
which wraps the then and else branches in a thunk:
$$
|apply0 (one( (e_1; (thunk (e_2))) choice (thunk (e_3))))|
$$
The argument of |oneKW| evaluates to either |((thunk (e_2)) choice dots)| or |(thunk (e_3))|
depending on whether |e_1| succeeds or fails, respectively, and |oneKW| then picks that lambda and applies it to |emptytup|.
As a bonus, provided we do no evaluation under a lambda, then |e_2| and |e_3| will
remain un-evaluated until the choice is made, just as we expect.

We use the same local-propagation rule for |oneKW| that we do for choice (\Cref{sec:mixing});
together with the desugaring for |if| into |oneKW|,
we get the ``special property'' of |if| described above.

\subsection{Arrays and |allKW|} \label{sec:array}
The main data structure in \versecalc{} is the array.
An array is a finite sequence of values, |tup (v_1,dots,v_n)|.
It can be used like a function: indexing is simply function application
with the argument being integers from 0 and up.  Indexing out of range is |fail|.
For example, |def a (a = (10,27,32); apply1 a 1)| reduces to |27|.
The reduction rule for indexing in arrays allows a multi-valued index expression.
For instance, |a := (10,27,32); apply1 a (1 `choice` 0 `choice` 1)| will reduce to |(27 `choice` 10 `choice` 27)|.

Arrays can be constructed by collecting all the results from a multi-valued expression, using the |allKW| construct:
if |e| fails the |all e| is the empty array,
otherwise if |e| reduces to |(v_1 `choice` dots `choice` v_n)| then |all e| reduces to
the array |tup (v_1, dots, v_n)|.
\lennart{Somewhere we should mention that |choice| is associative (in some sense) and can be reassociated.}
\label{sec:all}

You might think that array indexing would be stuck until we know the index,
but \versecalc{} uses narrowing to make progress. The expression |(def i (apply1 a i))|
looks stuck because we have no value for |i|, but in fact it rewrites to
$$
|def i ((i=0; 10) choice (i=1; 27) choice (i=2; 32)|
$$
which (as we will see in \Cref{sec:rewrite-rules}) simplifies
to just |(10 `choice` 27 `choice` 32)|.
So |allKW| allows a choice to be reified into an array; and |(def i (apply1 a i))| allow an
array to be turned back into a choice.

Do we even need |oneKW| as a primitive construct, given that we have
|allKW|? Can we not use |apply1 (all e) 0| instead of |one e|? Indeed
they behave the same if |e| fully reduces to finitely many choices of
values. But |allKW| really requires the evaluation of \emph{all}
choices before proceeding, while |oneKW| only needs to evaluate the
first choice. So, supposing that |loop| is a non-terminating function,
|one (1 `choice` apply0 loop)| reduces to 1, while
|apply1 (all (1 `choice` apply0 loop)) 0| loops.


\newcommand\startexample{\vspace{1em}\noindent}

\section{Some Verse examples} \label{sec:verse}
\versecalc{} is a fairly small language, but it is quite expressive.
We will illustrate this with a few examples.

\subsection{Syntactic sugar}

\begin{figure}
  $$
  \begin{array}{l}
    \begin{array}{rcll}
      |x := e; dots| & \equiv & |def x (x = e; dots)| \\
      |apply1 f x := e; dots| & \equiv & |f := lam x e; dots| \\
      |applyN f (x,y) := e; dots| & \equiv & |apply1 f p := (def (x,y) (p=tup (x,y); e)); dots| & \text{|p| fresh} \\
      |apply1 e_1 e_2| & \equiv & |def (x,y) (x = e_1;y=e_2;apply1 x y)| & \text{|x,y| fresh} \\
      |tup (e_1, e_2)| & \equiv & |def (x,y) (x = e_1;y=e_2; tup (x,y))| & \text{|x,y| fresh} \\
      |e_1+e_2| & \equiv & |applyN add (e_1,e_2)| \\
      |e_1>e_2| & \equiv & |applyN gt (e_1,e_2)| \\
      |if e_1 then e_2 else e_3| & \equiv & |t := one( (e_1; thunk e_2) choice (thunk e_3)); apply0 t| & \text{|t,x| fresh} \\
      |for e_1 in e_2| & \equiv & |v := all (e_1; thunk e_2); apply1 map (tup (lam z (apply0 z), v)| & \text{|v,x| fresh} \\
    \end{array} \\
  \end{array}
  $$
\caption{Syntactic sugar}
\label{fig:sugar}
\end{figure}

The {\versecalc} grammar can be a bit unwieldy, with explicit |defKW| and ANF-like restrictions allowing only values,
not arbitrary expressions, in in array literals and function-applications.
To keep the examples readable, we allow ourselves a bit of syntactic sugar, summarized in \cref{fig:sugar}:
\begin{itemize}
\item With |:=| we simultaneously introduce and assign a variable.
\item When defining functions, we allow simple patterns on the left of |:=|
\item In function applications we allow ourselves to write |apply1 e_1 e_2|,
  rather that insisting on both function and argument being a value, and
  likewise for pairs |tup (e_1,e_2)|.
\item We allow the use of |if|, as explained in \cref{sec:if}.
\item We allow the use of |for|, as we will explain in \cref{sec:for}.
\end{itemize}

\begin{figure}
  $$
  \begin{array}{l}
    \begin{array}{l@@{\;}l}
      |apply1 head xs| & :=  |one (def i (apply1 xs i))| \\
      |apply1 tail xs| & :=  |all (def i (i > 0; apply1 xs i))| \\
      |applyN cons (x,xs)| & :=  |all (x `choice` def i (apply1 xs i))| \\
      |applyN append (xs, ys)| & :=  |all (def i (apply1 xs i) `choice` def i (apply1 ys i))| \\
      |applyN flatMap (f,xs)| & := |all (def i (apply1 f (apply1 xs i)))| \\
      |applyN map (f, xs)| & :=  |if x := apply1 head xs then applyN cons (apply1 f x, applyN map (f,apply1 tail xs)) else emptytup| \\
      |applyN filter (p, xs)| & :=  |all (def i (x := apply1 xs i; one (apply1 p x); x)))| \\
      |applyN find (p, xs)| & :=  |one (def i (x := apply1 xs i; one (apply1 p x); x)))| \\
      |applyN every (p, xs)| & :=  |applyN map (p,xs)| \\
    \end{array}
  \end{array}
  $$
\caption{Common list functions} \label{fig:list-functions}
\end{figure}

\subsection{List functions}

We can define the typical list functions one would expect from functional programming
by using the duality between arrays and choices, as seen in \cref{fig:list-functions}.
An array can be turned into choices by indexing, |apply1 a i|, with a logical variable |i|.
Conversely, choices can be turned into an array using |allKW|.
The choice operator, |choice|, serves as both |cons| and |append| for choices
and |oneKW| is like |head| for choices.  With that in mind, here is how to understand
the definitions in \cref{fig:list-functions}:
\begin{itemize}
\item |head|.
To obtain the first element of an array, we turn the array into a sequence of choices and then pick out the first value using |oneKW|. If the array is empty, |head| fails.

\item |tail|.
To get the tail of an array, again we apply the array to a fresh variable, and restrict that variable with |i > 0|.
This way, the expression fails at the first value and only the other values are collected by |allKW|.
This definition of |tail| returns the empty array if |xs| is empty or not even an array.

\item |cons| and |append|.
We can implement |applyN cons (x,xs)| by collecting with |allKW| the choices we get from choosing |x|,
or all the elements of |xs|, and similarly we can define |append|.

\item |flatMap|.
For the function |map|, which applies a function to each element of an array and collects the results,
two possible implementation come to mind, and they differ when the function passed to |map| fails or produces more than a single value.
|flatMap| flattens the choices from iterating through |xs| with
the choices taken by |f|, so |applyN flatMap ( (lam x ((x choice x+1))), tup (10,20)))| $\equiv$ |tup (10,11,20,21)|.

\item |map|.
But often this is not what we want: we may want the effects of |f|, such as failure or returning multiple values,
to affect the whole expression and obtain the cartesian product of these choices.
The expression |applyN map ((lam x ((x `choice` x+1))), tup (10,20))| should reduce to |tup (10,20)`choice` tup(10,21) `choice` tup (11,20) `choice` tup (11,21)|.
We get that behaviour if we simply define |map| much like in any other functional programming language:
if taking the head of the array fails, it must be an empty array, and we return an empty array.
Else we apply |f| to the head, recurse on the |tail| and assemble the result with |cons|.\label{sec:map}

\item |filter|.
How can we filter an array with a predicate |p|?
We turn the array into the sequence of choices, and let those fail where |p| fails before collecting the other values into an array.

\item |find|.
If we use |oneKW| instead of |allKW|, we get the function that checks if |p| holds for some element of the array,
and returns the first matching element, or fails if |p| does not hold for any element.

\item |every|.
And to test if a predicate |p| holds for every element of an array |xs|, we can simply use |map(p,xs)|!
If |p| fails for any value, the whole call to |map(p,xs)| fails.
\end{itemize}

\subsection{Choice comprehensions} \label{sec:for}

Source languages can use this to build more featureful constructs like |for|,
mimicking the list comprehension in languages like Haskell or Python:
The expression |for^(e_1) in e_2| will evaluate |e_2| for each of the choices in |e_1|.
The |e_1| is called the \emph{domain} of the loop.

The scoping is peculiar in that variables bound in |e_1| also scope over |e_2|.
So, e.g., |for^(def x (x = 2 `choice` 3 `choice` 5)) in (x+1)| will reduce to the array |tup (3,4,6)|.
This example can, with the |:=| shorthand, be written |for^(x := 2 `choice` 3 `choice` 5) in (x+1)|.

Like list comprehension it supports filtering, but in Verse this falls out naturally by just using
a possibly failing expression in the domain.
So, |for^(x := 2 `choice` 3 `choice` 5; x > 2) in (x+1)| reduces to the array |tup (4,6)|.

Nested iteration in a |for| works as expected, and requires nothing special.
So, |for^(x:=10 `choice` 20; y := 1 `choice` 2 `choice` 3) in x+y| reduces to |tup (11, 12, 13, 21, 22, 23)|

Just like |if| is defined in terms of the primitive |oneKW| we can define |for|
in terms of the primitive |all e|.
Again, we have to be careful when |e_2| itself fails or produces multiple results;
simply writing |all (e1;e2)| would give the wrong semantics. So we put |e_2| under a lambda, and
apply each element of the array to |emptytup| afterwards, using the |map| function defined above.
The full desugaring for |for^(e_1) in e_2| is thus
\[
|v := all (e_1; thunk e_2); applyN map ((lam z (apply0 z)), v)|
\]
for a fresh variable |v|.

Note how this achieves that peculiar scoping rule: variables defined in |e_1| are in scope in |e_2|.
Any effects (like being multivalued) in |e_2| will not affect the choices defined by |e_1| since they are in a thunk.
So, e.g., |for^(x:=1 `choice` 2) in {3 `choice` x}| will reduce to
|tup (3, 3) `choice` tup (3, 2) `choice` tup (1, 3) `choice` tup (1, 2)|.
At this point it is crucial to use |map|, not |mapFlat|.

Given that array indexing expands into choices, we can iterate over array indices and elements using |for|.
E.g., |for^(def (i, x) (x = apply1 a i)) in x+i| will produce an array with the elements of |a|, increased by their index in |a|.


\subsection{Enumerations}
Making a choice between a sequence of numbers can be expressed like this
$$
\begin{array}{lll}
  |applyN fromTo (lo,hi)| & := & |if lo > hi then fail else lo `choice` applyN fromTo (lo+1,hi)|
\end{array}
$$
so |for^(x:= applyN fromTo (1,5)) in x| will reduce to the array |tup (1,2,3,4,5)|.

\subsection{Handling out-of-bounds errors}
Indexing out-of-bounds of an array results in |fail|.
This means that we cannot avoid handling this case.
$$
|if x:=apply1 a 5 then apply1 print x else apply1 print "out of bounds"|
$$

\subsection{Subset and powerset}
\lennart{Is this example too elementary?}
The subsets (subarrays) of of an array can be expressed with this multi-valued function.
First compute the subsets of all but the first element, and then return these with and without the first element added.
$$
|subset^(a) := if x:=apply1 head a then s:= apply1 subset (apply1 tail a); applyN cons (x, s) `choice` s else emptytup|
$$
And computing the powerset is then simply a matter of collecting all the subsets into an array.
$$
|powerset^(a) := for x := apply1 subset a in x|
$$

\lennart{Add more logic programming examples.}

\section{Rewrite rules} \label{sec:rewrite-rules}

\input{rewrites}

How can we give a precise semantics to a non-strict functional
programming language?  Here are some possibilities:
\begin{itemize}
\item \emph{A denotational semantics} is the classical approach, but it is quite mathematical
  (and hence not a good way to explain to programmer what their program means),
  and it cannot distinguish between call-by-need and call-by-name.
  It is hard to see how to give a (perspicuous) denotational semantics to a functional logic language,
  because of the logical variables.
\item \emph{A big-step operational semantics} typically involves explaining how a (heap,expression) starting point evaluates to a (heap,value) pair; Launchbury's natural semantics for lazy evaluation \cite{launchbury:natural} is the classic paper.  The heap, threaded through the semantics, accounts for updating thunks as they are evaluated.
\item \emph{A small-step operational semantics}. Despite its ``operational semantics'' title, the big-step approach
  does not convey accurate operational intutition, because it goes all the way to a value
  in one step. So-called ``small-step'' operational semantics are therefore widely used;
  they typically describe how a (heap, expression, stack) configuration evolves, one small
  step at a time (e.g. \cite{pj:stg-machine}).
  The difficulty is that the description is now so low-level that it is again hard to explain to programmers.
\item \emph{A rewrite semantics} steers between these two extremes.
  For example, Ariola \emph{et al}'s ``A call by need lambda calculus'' \cite{ariola:call-by-need} shows how
  to give the semantics of a call-by-need language as a set of rewrite rules. The great
  advantage of this approach is that it is readily explicable to programmmers.
  Indeed teachers almost always explain the execution of Haskell or ML programs as a succession of rewrites
  of the program (e.g. inline this call, simplify this |case| expression, etc).
\end{itemize}
Up to this point there has been no satisfying rewrite semantics for functional logic languages
(see \cref{sec:related} for previous work).
Our main contribution is to fill this gap with a rewrite semantics for \versecalc{}, that
has the following properties:
\begin{itemize}
\item The semantics is expressed as a set of rewrite rules.
\item Any rule can be applied, in either direction,
  anywhere in the program term (including under lambdas) to obtain an equivalent program.
\item The rules are oriented, with the intent that using them left-to-right makes progress.
\item Even given this orientation, the rules do not say in which rule should be
  applied where; that is the task of a separate \emph{evaluation stragegy}.
\item The rules can be applied by programmers, to reason about what their program does;
  and by compilers, to transform (and hopefully optimise) the program.
\item There is no ``magical rewriting'' (\cref{sec:magical}): all the variables on the
  right hand side of a rule are bound on the left.
\end{itemize}
The rewrite rules are given in Figures \ref{fig:rules1} and \ref{fig:rules2},
and are described next.

\subsection{Basic rewrites}

Looking at \cref{fig:general-axioms}, rules \rulename{p-add} and \rulename{p-gt}
should be familiar: they simply rewrite applications of |add| and |gt| to integer
constants.  For example $|applyN add (3,4)| \movesto |7|$.  The only unusual point is
that |gt| fails if $k_1 \leq k_2$ (rather than returning |False| as is more conventional),
and returns |k_1| otherwise.

% Rule \rulename{p-map} implements a map operation over arrays: It applies the function |f| to each value
% in turn, and collects the results into a array. If the function applications produce a choice
% of values, this will, as it should, produce a choice of arrays:
% \[
% |applyN map ((lam x (x choice  x+1)), (10,20))| \movesto^\ast |(10,20) choice (10,21) choice (11,20) choice (11,21)|
% \]

Sequencing, is very simple (rule \rulename{seq}):
|(e_1;e_2)|, simply reduces |e_1| to a value, and then discards it, returning
|e_2|.  The reason for evaluating |e_1| is for its unification effects.
Our insistence on clear sequencing also makes it possible to extend \versecalc{}
with more general effects, but that is beyond the scope of this paper.

\subsection{Function application} \label{sec:beta}

Beta-reduction is performed quite conventionally by \rulename{app-beta}; the
only unusual feature is that we use a |defKW| to bind |x| on the right-hand side,
and an equality |(x=v)| to bind |x| to the argument.
The rule may appear to use call-by-value, because the argument is a value |v|,
but remember that values include variables, and that variable may be bound to an
as-yet-unevaluated expression.  For example:
$$
|def y (y = 3+4; apply1 (lam x (x+1)) y)|
\; \movesto \;
|def y (y = 3+4; def x (x=y; x+1))|
$$
Finally, the side condition $x \not\in \freevars{v}$ in \rulename{app-beta}
ensures that the |defKW x|
does not capture any variables free in |v|.  If |x| appears free in |v|, just use
\textalpha-conversion to rename |x| to $x' \not\in \freevars{v}$.

In \versecalc{}, arrays behave like (finite) functions, in which application is
indexing.  Rule \rulename{app-arr} describes how array application works.
Notice that \rulename{app-arr} does not require the argument to be an integer $k$; instead
the rule works by narrowing.  So the expression |def x (apply1 (tup(2,3,2,7,9)) x = 2; x)|
does not suspend awaiting a value for |x|; instead it explores all the
alternatives, returning |(0 choice 2)|.  This is a free design decision:
a suspending semantics would be equally easy to express.

\subsection{Unification: the easy bits}

Next we study unification, again in \cref{fig:general-axioms}.
Rules \rulename{ulit}, \rulename{ucon}, \rulename{ux1} and \rulename{ux2}
are the standard rules for unification. The only special thing
to note is that |(k_1 = k_2)| either fails if the values differ,
or yields common value if they are the same; and similarly for
tuples\footnote{An alternative choice would
for unification to yield |emptytup| on success; but it does not make much difference
either way.}.  \rulename{ux1} and \rulename{ux2} make unfication fail if
you attempt to unify an integer with a tuple.
Slightly less obviously,
\rulename{ux3} through \rulename{ux6} make unification fail if you attempt to unify a
lambda or a built-in operation with any other value.  We discuss this design choice further in \cref{sec:unifying-lambda}.

\subsection{Unification: the \rulename{subst} rule}

The key innovation in \versecalc{} is the way bindings (that is, just ordinary equalities)
of logical variables are propagated.
The key rule is \rulename{subst}:
$$
\rulename{subst} \quad |X^[x==v]|  \movesto  |(subst X x v)^[x==v]|
        \quad \text{if $|x| \in \freevars{X}$, $|x| \notin \freevars{v}$} \\
$$
It says that if we have an equality |(x=v)| in a term,
we can replace ``nearby'' occurrences of |x| by |v|.  In this rule, |X| is an \emph{execution context},
defined in \cref{fig:syntax}.
In general, a ``context'' is a syntactic construct (in this case an expression)
with a single hole, written |hole|, in it.
The notation |X^[e]| is the expression obtained by filling the hole in |X| with |e|.
For example |(y = tup (hole, 4))| is a possible execution context, |X|; filling its hole with 99
gives the expression $|X^[99]| \equiv |(y = tup (99,4))|$.
The notation |subst X x v| is the context obtained by replacing all free occurences of |x| in |X| by |v|.

So \rulename{subst} says that we can replace all occurrences of |x| by |v| in the
|X|-context surrounding an equality |(x=v)|, thereby propagating |x|'s value to those occurrence sites.
There are several things to notice:
\begin{itemize}
\item \rulename{subst} fires only when the right-hand side of the unification is a \emph{value} |v|, so that the substitution does not risk duplicating work, or choices. This
restriction is what implements so-called \emph{call-time choice} \cite{hanus:survey}.
We do not need a heap, or thunks, or updates; the equalities of the program elegantly
suffice to express the necessary sharing.
The fact that we substitute only values for variables also ensures that substitution is syntactically
valid; if we tried to substitute an expression |e| for a variable in |(x,2)| we would get the
ill-formed expression |(e,2)|.
\item \rulename{subst}
  replaces the occurrences of |x| in |X|, \emph{but leaves the original |(x=v)| undisturbed}.
  For example, we can rewrite |((y=x+1; x=3); z=x+3)| to |((y=3+1; x=3); z=x+3)|, using
  |X = (y=x+1; hole)|.  We must not drop the |(x=v)| because there may be other occurrences of |x|
  ``further out'', such as |(z=x+3)| in this example.
\item If the equality |(x=v)| turns out to be the sole remaining occurence
      of |x| in the body of a |defKW|, then rule \rulename{def-elim} can drop both the |defKW|
    and the unification equation.
\item
Note carefully  that \emph{execution contexts |X| do not include |oneKW|, |allKW|, and choice}.
That restriction is the key to localising unification (\cref{sec:mixing}), and
the flexible/rigid distinction of
\cref{sec:rigid}. For example, consider the expression
|(y = ((x=3; x*2) choice (x=4)); (x+1,y))|.
We must not propagate the binding |(x=3)| to the expression |(x+1)|,
because the latter is outside the choice, and a different branch of the choice binds |x| to |4|.
But rule \rulename{subst}
\emph{can} propagate it locally
within the first arm of the choice, thus\footnote{
You may wonder if this local propagation is useful, a point we return
to in \cref{sec:local-prop}.}:
$$
|y = ((x=3; x*2) choice (x=4)); tup (x+1,y)|
\movesto
|y = ((x=3; 3*2) choice (x=4)); tup (x+1,y)|
$$
To make further progress, we need a rule for choice, \cref{sec:choice-rules}.
\item The $(x \in \freevars{X})$ in \rulename{subst} prevents the rule from
  firing infinitely often, doing nothing each time.  The other guard $(x \not\in \freevars{v})$
  also prevents \rulename{subst} from applying infinitely often, this time unrolling a recursive loop; but
  see \cref{sec:recursion}.
\end{itemize}


\subsection{Unification: recursion and the occurs check} \label{sec:occurs} \label{sec:recursion}

The classic ``occurs check'', rule \rulename{ux-occurs}, makes an equality like |(x=tup(x,3))| fail, because the only successful
unification would yield an infinitely-nested pair.
$$
    \rulename{ux-occurs} \quad |x=V^[x]| \quad \movesto \quad |fail| \quad \text{if }|V| \ne |hole|
$$
In this rule, |V| is an \emph{value context}, defined in \cref{fig:syntax}.
So \rulename{uc-occurs} says that the equality fails if there is a |V| such the equality looks like |(x=V^[x])|;
that is, if |x| occurs under a ``shell'' of tuples.  The guard $|V| \not= |hole|$ is to avoid failure for |x=x|.

What if |x| occurs inside a lambda?  That is very useful, and we do not want to fail; for example
$$
|factorial = lam n (if (n=0) then 1 else n * apply1 factorial (n-1))|
$$
Our solution, \rulename{subst-rec}, is standard: if the occurrence of |x| is inside a lambda |(lam y e)|,
we create a new local binding for |x| under the lambda.  Now the original equality is
non-recursive, and \rulename{subst} can apply.


\subsection{Unification: administrative rules}

A number of rules are administrative: they shuffle expressions about in order to make
the main unification rules apply.
Rule \rulename{def-float} moves a |defKW| ``out of the way''. For example,
the expression |(y = (def x (tup (x,2))); dots)| does not bind |y| to a value, but if we float the |defKW| out
we get |def x (y=tup(x,2); dots)|, and now we may be able to make progress by substituting for |y|.
Similarly rules \rulename{unify-seql} and \rulename{unify-seqr} move sequencing out of unification,
again to make it more likely to get the form |(x=v)|.  Notice that |e_1|, |e_2| and |e_3| stay in the
same left-to-right order, there is no risk of re-ordering choices (c.f. \cref{sec:mixing}).

Rules \rulename{unify-unifyl} and \rulename{unify-unifyr} move a unification between values out of another unification.
Without these rules |(x=y)=1;y>0| would be stuck; but \rulename{unify-unifyl}
rewrites it to |((x=y);(y=1);y>0|, thus exposing |y=1| for further use by \rulename{subst}. Furthermore, in order
to prove confluence, we need to be able to freely move the equation |(v_1=v_2)|,
as discussed in \cref{sec:conf-move-eqn}.
%% The \label has been commented out.
% These rules could be more general, see \cref{sec:variations-liberate-unify}.

Finally, rule \rulename{swap} swaps the arguments to unification to make \rulename{subst} apply.

\subsection{Local substitution}

Consider this tricky term:
$$
|def x (x = (if (x=0; x>1) then 33 else 55))|
$$
What should this do? At first sight you might think it was stuck; how can we simplify the |if|
when its condition mentions |x| which is not yet defined.  But in fact rule \rulename{subst} allows
us to substitute \emph{locally} in the |X|-neighbourhood of the equality |(x=0)| thus:
$$
\begin{array}[t]{r@@{\;}l}
           & |def x (x = if (x=0; x>1) then 33 else 55)| \\
  \movesto & |def x (x = if (x=0; 0>1) then 33 else 55)| \\
  \movesto & |def x (x = if fail then 33 else 55)|\\
  \movesto & |def x (x = 55)| \quad \movesto |55|
  \end{array}
$$
But this is example is arguably a weird corner case.  Perhaps we can simplify the system, by only allowing
substitution over the entire scope of the |defKW|, and combining it with \rulename{def-elim} thus:
$$
\rulename{global-subst} \quad |def x (X^[x==v])|  \movesto  |(subst X x v)^[v]|
        \quad \text{if $|x| \in \freevars{X}$, $|x| \notin \freevars{v}$} \\
$$
Now the tricky term is stuck; but maybe that does not matter. After
all, even with local substitution, the tricky term would be stuck if we
replaced |(x=0)| with |(x=33)|, as the reader may care to verify.
\emph{But the real reason we need local substitution is not expressiveness, but confluence.}
Suppose we use \rulename{global-subst} instead of \rulename{subst} and consider
$$
|def x (x = if (def y (y=x; y=0; y>1)) then 33 else 55)|
$$
In the condition of the |if| we have introduced a local variable |y|, and set it equal to |x|.
By using \rulename{global-subst} on |(y=x)|, we could recover the original tricky term, which is
stuck.

\emph{But by using \rulename{global-subst} on |(y=0)|, instead, we could simplify the |if|.}
Two different rewrites starting from the same term lead to non-joinable terms.
In effect, local substitution via \rulename{subst} has the same effect as introducing a local
alias for the unified variable.  This is a tricky point!

\subsection{Rules for choice} \label{sec:choice-rules}

The rules for choice are given in \cref{fig:choice}. Rules \rulename{fail-l} and \rulename{fail-r}
say that |fail| is the unit for choice |choice|.  Rules \rulename{one-value}, \rulename{one-choice}
and \rulename{one-fail} describe the semantics for |oneKW|, just as in \cref{sec:one}.
The \rulename{assoc-choice} is an administrative rule, to allow \rulename{one-choice} to fire.

The most interesting rule is \rulename{choose} which, just as described in \cref{sec:choice},
``floats the choice outwards'', duplicating the surrounding context.
But what ``surrounding context'' precisely? We use two new contexts, |SX| and |CX|, both
defined in \cref{fig:syntax}. A \emph{choice context} |CX| is like an
execution context |X|, \emph{but with no possible
  choices to the left of the hole}:
$$
CX ::= \hole \vb |CX == e| \vb |ce == CX| \vb |CX ; e| \vb |ce; CX|
\vb |def x CX|
$$
Here, |ce| is guaranteed-choice-free expression (syntax in \cref{fig:syntax}).
This syntactic condition is necessarily conservative; for example, a call |apply1 f x|
is considered not guaranteed-choice-free, because it depends on what function |f| does.
We must guarantee not to have choices to the left so that we preserve order --
see \cref{sec:mixing}.

The context |SX| (\cref{fig:syntax}) is a \emph{scope context}; it
ensures that |CX| is as large as possible.  This is a subtle point:
without this restriction we lose confluence.  To see this, consider\footnote{Remember, |if| is syntactic sugar for
  a use of |oneKW|, see \cref{sec:one}, but using |if| makes the example easier to understand.}:
$$
\begin{array}{ll}
\multicolumn{2}{l}{|def x (if (x>0) then 55 else 44; (x=1); (77 choice 99))|}\\
\movesto \{\rulename{subst}\} & |def x (if (1>0)  then 55 else 44; (x=1); (77 choice 99))|\\
\movesto \{\text{simplify |if|}\} & |def x (55; (x=1); (77 choice 99))|\\
\movesto \{\rulename{seq}\} & |def x ((x=1); (77 choice 99))|\\
\movesto \{\rulename{def-elim}\} & |1; (77 choice 99)|\\
\movesto \{\rulename{seq}\} & |77 choice 99|\\
\end{array}
$$
But suppose instead we floated the choice out, \emph{part-way}, like this:
$$
\begin{array}{ll}
\multicolumn{2}{l}{|def x (if (x>0) then 55 else 44; (x=1); (77 choice 99))|}\\
\movesto \{\text{Bogus \rulename{choose}}\} & |def x (if (x>0) then 55 else 44; ((x=1); 77) choice ((x=1); 99))|\\
\end{array}
$$
Now the |(x=1)| is inside the choice branches, so we cannot use \rulename{subst} to rewrite the
condition of the |if|.  Nor can we use \rulename{choose} again to float the choice further out,
because the |if| is not guaranteed choice-free (for example, the branches might contains choices).
So, alas, we are stuck!

Our not-entirely-satisfying solution is to force \rulename{choose} to float the choice all the way
to the innermost enclosing scope construct; hence the |SX| in the rule.

\subsection{\versecalc{} is lenient}

\versecalc{} is a \emph{lenient}, not lazy (call-by-need), nor strict (call-by-value).  Under lenient
evaluation, everything is eventually evaluated, but functions can run
before their arguments have a value.
Consider a function
call |apply1 f e|.  In \versecalc{} applications are in administrative normal form (ANF),
so we must actually write |(x=e; apply1 f x)|. This expression will not return a value until
|e| reduces to a value: that is, everything is eventually evaluated.  But even so |apply1 f x| can proceed
to beta-reduce (\cref{sec:beta}), assuming we know the definition of |f|.
This is not just a way to get \emph{parallelism}, which was the motivation for introducing
lenience in the dataflow language Id \cite{lenience}; in a functional logic language
it is critical for \emph{semantics}.  Suppose |y = lamhdr (tup(p,q)) (p=7;q)| and consider
$$
|def (f,y) (f = (lam p (x=7;p)); y = (if (x>0) then 7 else 8); apply1 f y|
$$
Here |y| does not get a value until |x| is known; but |x| does not get its
value (in this case |7|) until |f| is called.  Without lenience this program would be stuck.


\subsection{Examples}
\lennart{I think we should have a few more examples of reduction sequences.}

\section{Confluence}

The rules of our rewrite semantics can be applied anywhere, in any order, and they give meaning to programs without committing to a particular evaluation strategy. But then it had better be the case that no matter how the rules are applied, one always obtains the same meaning for the program!

If the rewrite rules were terminating, then we would expect strong normalization: all possible execution paths lead to the same normal form. But since {\versecalc} has recursion, we can aim for \emph{confluence} instead: If $e_1 \movesto^\ast e_2$ and $e_1 \movesto^\ast e_3$, then we can find $e_4$ so that $e_2 \movesto^\ast e_4$ and $e_3 \movesto^\ast e_4$.

Confluence will give us the assurance that we will not get different results when choosing different rules, or get stuck with some rules and not with others.

\subsection{subst vs. subst}

The main source for possible non-confluence is rule \rulename{subst}: If we have multiple equations for a variable $x$ available, we can use any of them to rewrite. A typical example would be
\[ |def x (x=v_1; x=v_2; e)| \]
which we can rewrite by \rulename{subst} (and administrative \rulename{def-elim} and \rulename{seq}) to either
\[ |v_1=v_2; subst e x v_1| \quad \text{or} \quad |v_2=v_1; subst e x v_2|. \]
Can we still join these two expressions? The unification rules allow us to decompose an equality |v_1 = v_2| completely: Either it reduces to |fail|, which will swallow the whole expression, or to a sequence of equalities with a variable on (at least) one side. Using \rulename{subst} on these equations, we can join the occurrences of $v_1$ resp.\ $v_2$ in |subst e x v_1| resp.\ |subst e x v_2|.

So it seems that confluence is within reach? Unfortunately, not quite:

\subsection{Confluence up-to}

\begin{figure}
\[\begin{array}{lrcl}
    \rulename{def-swap} & |def (x,y) e| & \movesto & |def (y,x) e|\\
    \rulename{unify-move} & |X^[v_1=v_2]| & \movesto & |v_1=v_2;X^[v_2]| \\
    \rulename{var-swap} & |x=y| & \movesto & |y=x| \\
\end{array}\]
  \caption{Additional rewrite rules for confluence}
    \label{fig:rules-confluence}
\end{figure}

\label{sec:conf-move-eqn}

There some sources of non-confluence that we would consider insignificant. For example
| (def x (x=z));(def y (y=z)) |
can be reduced to
| (def (x,y) (x=z;y=z)) |
or
| (def (y,x) (x=z;y=z)) |, which are different expressions, but the difference is one we do not care about. We can capture this idea that our notion of confluence may ignore the order of |defKW|s by adding an additional rule, \rulename{def-swap} in \cref{fig:rules-confluence}, to the rewrite system. This is obviously looping, but this is not an issue.

Similarly, we do not care about where, within the same $X$-context, an equality is asserted, so we allow ourselves to move any such equation to the front of the $X$-context, using \rulename{unify-move}. We leave $v_2$ in place, just like \rulename{def-elim} does. Adding this rules does not allow more or fundamentally different evaluations; this motivated adding the rules \rulename{unify-unifyl} and \rulename{unify-unifyr}.

Finally, to make equality on values commutative, we add \rulename{var-swap}; else we’d get non-confluence from |def x (x=y;x=z)|, which can reduce to |y=z;z| or to |z=y;y|.

\subsection{Skew-confluence}

With that out of the way, there still remains a source of non-confluence, due to recursion. In
\[ |def (x,y) (x=tup (1,y); y=(lam z x); x)| \]
we can apply \rulename{subst} to $x$ first and obtain the following reduction (omiting spelling out the application of rules \rulename{def-elim} and \rulename{seq}):
\[
\begin{array}{ll}
\movesto \{\rulename{subst}\}     & |def y (y=(lam z (tup (1,y))); tup (1,y))| \\
\movesto \{\rulename{subst-rec}\} & |def y (y=v_y; tup (1,y))| \quad \text{ where } |v_y| = |(lam z (def y (y = (lam z (tup (1,y))); tup (1,y))))| \\
\movesto \{\rulename{subst}\}     & |tup (1,v_y)|  \\
\end{array}
\]
Alternatively, starting with |y = (lam z x)|, we obtain
\[
\begin{array}{ll}
\movesto \{\rulename{subst}\}     & |def x (x=tup (1,lam z x); x)| \\
\movesto \{\rulename{subst-rec}\} & |def x (x=v_x; x)| \quad \text{ where } |v_x| = |tup (1,lam z (def x (x = tup (1,(lam z x)); x)))| \\
\movesto \{\rulename{subst}\}     & |v_x|  \\
\end{array}
\]
Can we join |tup (1,v_y)| and |v_x|? Given $|v_x| \movesto^\ast |tup (1,lam z v_x)|$ and $|v_y| \movesto^\ast |lam z (tup (1, v_y))|$, we get
\[
\begin{array}{c@@{\,}c@@{\,}c@@{\,}c@@{\,}c@@{\,}c@@{\,}c@@{\,}c}
& |tup (1,v_y)| & \movesto^\ast & |tup (1,lam z (tup (1, v_y)))| & \movesto^\ast & |tup (1, lam z (tup (1, (lam z (tup (1, v_y))))))| & \movesto^\ast & \cdots \\
|v_x|     & \movesto^\ast & |tup (1,lam z v_x)| & \movesto^\ast & |tup (1, lam z (tup (1, (lam z v_x))))| & \movesto^\ast && \cdots
\end{array}
\]
and we \emph{cannot} join them; they keep “missing each other”.

But notice that as we keep evaluating these fixed-points, they produce the same “infinite expression” |tup (1, lam z (tup (1, (lam z dots))))|. This phenomenon has been studied before and led to a various generalizations of confluence to allow this behavior.

In particular the notion of \emph{skew-confluence} as introduced by \citet{ariola:skew-confluence}, applies here nicely: Slightly paraphrased it states that given $e_1 \movesto^\ast e_2$ and $e_1 \movesto^\ast e_3$, it suffices that for every \emph{finite} observation on expressions, both $e_2$ and $e_3$ eventually reduce to expression that are indistinguishable by that finite observation.

Until we have concluded the formal proof of skew-confluence, including the precise definition of finite observation, we state our aspiration as a conjecture:
\begin{conjecture}{Skew-confluence}
  The reduction system $\movesto$ is skew-confluent.
\end{conjecture}


\section{Related work} \label{sec:related}

Now that we know what \versecalc{} is, we can identify its distinctive features,
compared to other approaches.

\subsection{Choice and non-determinism}

A significant difference between our presentation and earlier works is our
treatment of choice. Consider an expression like |(3 + (20 choice 30))|.
This is typically handled by a pair non-deterministic rewrite rules:
$$
|e_1 choice e_2| \movesto |e_1| \hspace{1cm} |e_1 choice e_2| \movesto |e_2|
$$
So our expression rewrites (non-deterministically) to either |(3+20)| or |(3+30)|;
and that in turn allows the addition to make progress.
Of course, including non-deterministic choice makes the rules are non-confluent by
construction.

In contrast, our rules never pick one side or the other of a choice.
And yet |(3 + (20 choice 30))| can make progress, by floating out the
choice (rule \rulename{choose} in \cref{fig:general-axioms}), thus
|(3+20) choice (3+30)|.  In effect, \emph{choices are laid out in space} (in
the syntax of the term), rather than being explored by
non-determinstic selection.  Rule \rulename{choose} is not a new idea -- indeed
it is common in calculi for calculi with choice (e.g. \cite{what}) --
but it does not appear to have been used to describe functional logic languages before.

\subsection{One and all}

Logical variables, choice, and equalities are present in many functional logic
languages.  However |oneKW| and |allKW| are distinctive features of \versecalc{}.

First, notice that |allKW| \emph{reifies} choice (a control operator) into an array (a data structure);
for example, |all (1 choice 7 choice 2)| returns the array |tup (1,7,2)|),
In the other direction, indexing turns an array into choice (e.g. |def i (apply1 (tup (1,7,2)) i)| yields |(1 choice 7 choice 2)|).
Other languages can reify choices into a (non-deterministic)
list, via an operator called \verb!bagof!, or a mechanism called
\emph{encapsulated search} in Curry.  But this is regarded as a somewhat sophisticated feature,
whereas it is part of the fabric of \versecalc{}.

Second, even under the reification of |allKW|,
\versecalc{} is \emph{deterministic}.  Choice is not
non-determinstic choice, it is takes pains to maintain order, so that
when reifying choice into an array, the order of elements in that
array is completely determined.  Anything else would leak
non-determinism into the language.  This determinism has a price: as we saw in \cref{sec:mixing}
and \cref{sec:rules-for-choice}, we have to take pains to maintain the
left-to-right order of choices.

Maintaining that order has other payoffs. For example, it is relatively easy to add
effects other than choice, including mutable varialbes and input/output, to \versecalc{}.
To substantiate this claim, \cref{sec:store} gives the additional syntax and rewrite rules
for mutable variables.

Thirdly, |oneKW| allows us to reify failure; to try something and take different actions
depending on whether or not it succeeds.  \simon{Can one do this in Curry?}

Finally, |oneKW| and |allKW| neatly encapsulate the idea of ``flexible'' vs ``rigid'' logical variables.
As we saw in \cref{sec:rigid}, logical variables bound outside |oneKW| cannot be unifies inside it;
they are ``rigid''.  This notion is nicely captured by the execution context |X|, which limits the
scope of the \rulename{subst} rule, and specifically excludes |oneKW| and |allKW|.

\subsection{The semantics of logical variables} \label{sec:magical}

Our variables, introduced by |defKW|, are usually called \emph{logical variables},
or \emph{extra variables} in the literature.  The latter term arises because they
are typically introduced as variables that appear on the right-hand side of a function
defintion, but are not bound on the left.  For example, in Curry we can write
\begin{verbatim}
     f x | x =:= (a,b) = a where a,b free
\end{verbatim}
Here \verb!a! and \verb!b! are logical variables, not bound on the left; they
get their values through unificaiton.   In Curry they are explicitly introduced
by the ``\verb!where a,b free!'' clause, while in many other papers their
introduction is implicit in the top-level rules, simply by not being bound
on the left.  Logical variables are
at the heart of the ``logic'' part of functional logic programming.

To give a high-level semantics for non-strict, non-deterministic
functional logic languages, the brand leader is Constructor based
ReWrite logic (CRWL).
It is a ``big-step'' rewrite semantics,
one that rewrites a term to value in one big step.
\Citet{fraguas:rewriting} make a powerful case for
instead giving the semantics of a functional logic language using ``small-step'' rewrite
rules, more like those of the lambda calculus, that successively rewrite the term, one step at
a time, until it reaches a normal form.
they offer set of small-step rewrite rules, that is provably equivalent to
the CRWL framework.  Their key insight (inspired by 
\citet{ariola:call-by-need}'s formalisation of the call-by-need lambda calculus)
is to use |let| to make sharing explicit.

However both CRWL and Fraguas \emph{et al} suffer from a major problem:
they require something we call \emph{magical rewriting}. A key rewrite rule is this;
$$
\begin{array}{l}
f( \theta(e_1),\ldots, \theta(e_n) ) \movesto \theta( rhs ) \\
\text{if}
\begin{array}[t]{l}
  (e_1, \ldots, e_n) \movesto rhs
   \; \text{is a top-level function binding} \\
\text{and $\theta$ is a substitution mapping variables to closed values} \\
\text{and $dom(\theta) = \mathit{fvs}(e_1, \ldots, e_n, rhs)$}
\end{array}
\end{array}
$$
The substitution for the free variables of the left hand side can
readily be chosen by matching the left hand side against the call.
But the substitution for the extra variables must be chosen ``magically''
\cite[Section 7]{fraguas:rewriting} or clairvoyantly, so as to make the future execution
work out. This is admirably high level, because it hides everything
about unification, but it is not much help to a programmer trying to
understand a program, nor is it directly executable.

To explain actual execution, the state of the art is described by
\citet{hanus:semantics}.  They give both a big-step operational
semantics (in the style of \citet{launchbury:natural}), and
a small-step operational semantics.  These two approaches both thread a
\emph{heap}, which holds
the unification variables and their unification state; and the small-step
semantics also has a \emph{stack}, to specify the focus of execution.
The trouble is that heaps and stacks are difficult to explain to a programmer,
and do not make it easy to reason about program equivalance.
To model residuation requires a new level of complexity, needing concurrency.

In contrast, our rewrite rules give a complete, executable (i.e. no ``magic'')
account of logical variables and choice, directly as small-step
rewrites on the original program, rather than as the evolution of a
(heap, control, stack) configuration. Moreover, we have no problem with
residuation.

\subsection{Flat vs higher order}

When giving the semantics
of functional logic languages, a first-order presentation is almost universal.
User-defined functions can be defined at top level only;
and function symbols (the names of such functions) are syntactically
distinguished from ordinary variables. As Hanus describes, it is possible
to translate a higher order program into a first-order
form\footnote{
  He does not mention this, but for a language with arbitrary nested lambdas
  one would need to do lambda-lifting as well, but that is a small point.}
using defunctionalisation \cite[Section 3.3]{hanus:survey}, and a buit-in
function |apply| function.
But this encoding is hardly a natural rendition of the lambda calculus, and it
obstructs the goal of using rewrite rules to explain to programmers how
their program might execute.

A strength of our presentation is that it deals natively with the full lambda calculus.

\subsection{Intermediate language}

Hanus's \emph{Flat Language} \cite[Fig 1]{hanus:semantics} plays the same role as \versecalc{}:
it is a small core language
into which a larger surface language can be desugared.
There are some common features: variables, literals, constructor applications,
and sequencing (written \verb!hnf! in the Flat Language).
However, it seems that \versecalc{} has a greater economy of concepts.
In particular, the Flat Language
has two forms of equality \verb!(==)! and \verb!(=:=)!, and two forms
of case-expression, \verb!case! and \verb!fcase!.  In each pair, the former suspends
if it encounters a logical variable; the latter unifies or narrows
respectively. In contrast, \versecalc{}
has a single equality |(=)|, and the orthogonal |oneKW| construct, to deal with all four concepts.

The Flat Language has let-expressions \verb!(let x=e in b)!, where \versecalc{} uses
|defKW| and (again) unification.
The Flat Language also uses the same construct for a different purpose, to
bring a logical variable into scope, using the strange binding \verb!x=x!, thus \verb!(let x=x in e)!.
In contrast, |def x e| seems more direct.

\section{Variations and choices}

In a calculus like \versecalc{} there is room for many design variations.
We discuss some of them here.

\subsection{Unifying lambdas} \label{sec:unifying-lambda}

In \versecalc{} we say that any attempt to unify two lambdas (or primitive operations) fails -- even if the
lambdas are semantically identical.  The problem, obviously, is that it is hard
to tell if they \emph{are} semantically identical.

We cannot say that the attempt to unify gets stuck, because that leads to non-confluence.
Here is an expression that could rewrite in two different ways:
$$
\begin{array}{lll}
 \multicolumn{3}{l}{|def x (x = (lam p 1); x = (lam q 2); x emptytup)|} \\
&    \movesto \{\textit{use \rulename{SUSBST} on second equality}\} & |(lam p 1) = (lam q 2); 1| \\
OR & \movesto \{\textit{use \rulename{SUSBST} on second equality}\} & |(lam q 2) = (lam p 1); 2|
\end{array}
$$
These two outcomes cannot be joined.  But making unification on lambdas fail, both outcomes
lead to |fail|, and confluence is restored.

\subsection{Choice and failure} \label{sec:choice-and-failure}

In \versecalc{} an expression delivers a \emph{sequence} of results, not a bag,
and certainly not a set (\cref{sec:choice}).  That suggests that a good evaluation strategy might be to
only explore the left branch of a choice first, and try the right branch only if former fails.  But consider
$$
|(if (x > 2) then e1 else e2); (x = 1 choice  fail)|
$$
The |if| is stuck until we have a value for |x|; we cannot float the choice out with \rulename{choose} because
|if| is no choice-free.  With the rules in \cref{fig:rules2} we \emph{can} use \rulename{fail-r} to rewrite
$|(x=1 choice fail)| \movesto |(x=1)|$; and now we can propagate the |(x=1)| to the |if| via \rulename{subst}
and make progress.  But to do this we must look at the right branch of the choice, before we know whether
the left branch will reach a value or fail.  There is nothing wrong with this, but it is arguably a bit peculiar that
|(x=1 choice e)| can substitute for |x| if and only if |e| fails.  Even |(x=1) choice (x=1)| cannot substitute
for |x|.

An alternative would be to drop \rulename{fail-l} and \rulename{fail-r}, and drop the |fail| terms only in
the rules for |oneKW| and |allKW|.  Both alternatives seems reasonable choices.

% \subsection{Liberating more equations} \label{sec:variations-liberate-unify}
% 
% Rules \rulename{unify-unifyl} and \rulename{unify-unifyr} help to make more equations available for rewriting with \rulename{subst}. For example, |(x=y)=1;y>2| rewrites to |x=y;y=1;y>2|, which allows \rulename{subst} to rewrite with |y=1|, and the whole expression collapses to |fail|.
% 
% But the quite similar expression |(apply0 f =y)=1;y>2| is stuck, because \rulename{unify-unifyr} requires the inner unification between \emph{values}. We chose the more restrictive variant because confluence calls for just that, but one could have a more complete set of rules instead, covering all cases where two values ought to be equated:
% \[\begin{array}{rcl@@{\hspace{10mm}}rcl}
% |(v_1 = e) = v_2| & \movesto & |(v_1 = e); (v_1 = v_2)|
% & |(e = v_1) = v_2| & \movesto & |(e = v_1); (v_1 = v_2)| \\
% |v_1 = (e = v_2)| & \movesto & |(v_1 = v_2); (e = v_2)|
% & |v_1 = (v_2 = e)| & \movesto & |(v_1 = v_2); (v_2 = e)| \\
%  \end{array}\]
% 
% \joachim{Not sure if this section is worth it, or does a good job of explaining why we don’t use the full rules. Also these rules are not confluent.}

\subsection{Not restricting application and arrays}

We have opted to restrict array formation and function application to only allow \emph{values} and not \emph{expressions}.
This is not a necessary restriction, but allowing expressions would require many more rules.
The extra rules would describe how to distribute, e.g., application over sequencing.
There is no loss of expressive power using just values, just slightly more verbose terms.

\subsection{Restricting unification}

The system becomes slightly simpler if we restrict the syntax of equalities so that the left-hand-side is always a value.
Thus $e ::= \ldots \vb |v = e| \vb \ldots$. It is easy to get the program into this form:
$$
|e_1 = e_2| \quad \text{desugars to} \quad |def x (x = e_1; x = e_2)|
$$
The payoff is that we can then drop \rulename{unify-seql} and \rulename{unify-unifyl}.
The difference is modest, but it would make our language a little smaller.

\subsection{Ordering and choices}

As we discussed in \cref{sec:choice-rules}, rule \rulename{choose} is less than satisfying, for two reasons.
First, the |CX| context uses a conservative, syntactic analysis for choice-free expressions; and second,
the |SX| context is needed to force |CX| to be maximal.  A rule like this would be more satisfying:
$$
\rulename{simpler-choose} \quad   |X^[ e_1 choice e_2 ]| \movesto |X^[e_1] choice X^[e_2]|
$$
The trouble with that is that it may change the order of the results (\cref{sec:mixing}).
Another possiblity would be to accept the the results may come out in the ``wrong'' order, but have
some kind of sorting mechanism to put them back into the ``right'' order.  Something like this:
$$
\rulename{labelled-choose} \quad   |X^[ e_1 choice e_2 ]| \movesto |X^[L; e_1] choice X^[R; e_2]|
$$
Here the two branches are labelled with |L| and |R|.  We can add new rules to reorder such
labelled expressions, something like
$$
\rulename{sort-l} \quad   |(R; e_1) choice (L; e_2)| \movesto |(L; e_2) choice (R; e_1)|
$$
We believe this can be made to work, but it adds unwelcome clutter to program terms, and the cure
may be worse than the disease.

\subsection{Generalizing |oneKW| and |allKW|}

In \versecalc{} we introduced |oneKW| and |allKW| as the primitive choice-consuming operators, and neither is more general than the other, as discussed in \cref{sec:all}. We could have, however, introduced a more general operator |splitKW| as $e ::=  |dots| \vb |split e v_1 v_2|$ and rules
\[\begin{array}{lrcll}
    \rulename{split-fail} & |split fail f g| & \movesto & |apply0 f| \\
    \rulename{split-choice} & |split (v `choice` e) f g| & \movesto & |applyN g (v, thunk e)| & \text{|x| fresh}  \\
    \rulename{split-value} & |split v f g| & \movesto & |applyN g (v, thunk fail)| \\
\end{array}\]
The intuition behind |splitKW| is that it distinguishes a failing computation from one that returns at least one value. If |e| fails, it calls |f|, and if |e| returns at least one value, passes that to |g| together with the remaining computation, safely tucked away behind a lambda.

Indeed, this is more general, as we can implement |oneKW| and |allKW| with |splitKW|:
\[\begin{array}{r@@{\;}l@@{\;}l@@{\;}l}
    |one e| & \equiv |apply1 f x := fail; | & |applyN g (x,y):= x;| & |split e f g| \\
    |all e| & \equiv |apply1 f x := emptytup;|    & |applyN g (x,y):= applyN cons (x, split (apply0 y) f g);| & |split e f g|
\end{array}\]
For this paper we stuck to the arguably simpler |oneKW| and |allKW|,
to avoid confusing the presentation with these higher-order encoding, but we do not expect any complications using |splitKW| instead.



\section{Conclusions and further work}
\lennart{This needs a little fleshing out.}

\subsection{Types}
The full Verse language has types and static checking of those.
In the dynamic semantics the types can be represented by partial
identity function---identity of the values of the type and |fail| otherwise.
We have a reduction rules for this, but they are left out for space and exposition reasons.

\subsection{Assignment}
The full Verse language has updateable references in a store, in the style of SML.
We also have (tentative) reduction rules for this as well.
They are non-trivial, because store updates have to be transactions, e.g.,
when the condition of an |if| fails any store effects in the condition are rolled back.
See \cref{sec:store} for the relevant rules.

\subsection{Effects}
In this paper we have seen one kind of effect, namely being multi-valued.
Adding assignable references add another set of effects, as does I/O.
To accomodate this, real Verse has an effect system in addition to the type system.
We have reduction rules for effect checking as well, but for the real language this is done at compile time.

% -------------------- Acknowledgements ----------------
\begin{acks}                            %% acks environment is optional
                                        %% contents suppressed with 'anonymous'
\end{acks}

% -------------------- Bibliography ----------------
\bibliography{../verse}

\appendix

\section{Updateable references} \label{sec:store}

\begin{figure}
  $$
  \begin{array}{l}
    \textbf{Syntax extension} \\
    \begin{array}{lrcl}
      \text{References}         & r \\
      \text{Expressions}        & e   & ::= & \cdots \vb |store s e| \\
      \text{Primops}            & op  & ::= & \cdots \vb |alloc| \vb |read| \vb |write| \\
      \text{Head values}        & hnf & ::= & \cdots \vb r \\
      \text{Execution contexts} & X   & ::= & \cdots \vb |store s X| \\
      \text{Store}              & s   & ::= & |cell r v| \vb |s,s| \vb |emp| \\
      \text{Store context}      & S   & ::= & \hole \vb |S,s| \vb |s,S| \\
    \end{array} \\
\\

  \textbf{Axiom extensions} \\
  \textit{Reference ops} \\
  \begin{array}{lrcll}
    \rulename{ref-alloc} & |store s (apply1 alloc v)| & \movesto & |store (s, cell r v) r| & \text{|r| fresh} \\
    \rulename{ref-read} & |store (S^[cell r v]) (apply1 read r)| & \movesto & |store (S^[cell r v]) v| \\
    \rulename{ref-write} &
        | store (S^[cell r v_1]) (applyN write (r,v_2))|
        & \movesto & |store (S^[cell r v_2]) emptytup| \\
  \end{array} \\
    \\
  \textit{Store propagation} \\
  \begin{array}{lrcll}
    % SEQ
    \rulename{st-seq-in} & |store s (e_1; e_2)|  & \movesto & |store s e_1; e_2| \\
    \rulename{st-seq-r} & |store s v; e|  & \movesto & |v; store s e| \\
    % UNIFY
    \rulename{st-unify-in} & |store s (e_1 = e_2)|  & \movesto & |store s e_1 = e_2| \\
    \rulename{st-unify-r} & |store s v = e|  & \movesto & |v = store s e| \\
    \rulename{st-unify} & |v_1 = store s v_2|  & \movesto & |v_1 = v_2; store s v_1| \\
    % CHOICE
    \rulename{st-choice} & |store s (e_1 `choice` e_2)| & \movesto & |store s e_1 `choice` e_2| \\
    % DEF
    \rulename{st-def-in} & |store s (def x e)| & \movesto & |def x (store s e)| & e \not\in v \\
    \rulename{st-def-out} & |def x (store s v)| & \movesto & |store s (def x v)| \\
    % SPLIT
    \rulename{st-split-value} & |split (store s v) f g| & \movesto &
        |store s (split v f g)| \\

    \rulename{st-split-choice} & |split (store s (v `choice` e)) f g| & \movesto &
      |store s (split (v `choice` e) f g)| \\

    % FAIL
    %% \rulename{st-wrong} & |store s wrong| & \movesto & |wrong| \\
  \end{array} \\
 \\
  \textit{Store copying} \\
  \begin{array}{lrcll}
    % SPLIT
    \rulename{st-split-copy} &
      \begin{array}[t]{@@{}l}
        | storeTwoLines s (split e f g)|
      \end{array} & \movesto & 
      \begin{array}[t]{@@{}l}
        | storeTwoLines s (split (store s e) f g)| \\
      \end{array} & \text{|canCopy e|} \\
    % FAIL
    \rulename{st-fail} & |store s fail| & \movesto & |fail| \\
    % STORE
    \rulename{st-store} & |store s_1 (store s_2 v)| & \movesto & |store s_2 v| \\
  \end{array} \\
    \\
    \textit{Unification} \\
    \begin{array}{lrcll}
      \text{Extension with the obvious axioms making equal references unify,
        and anything else fail.}
    \end{array} \\
  \\
  \textit{Top level} \\
  \begin{array}{lrcll}
    \text{Start top level reduction of |e| with |store emp e|.}
  \end{array} \\

    \\
    \text{|canCopy e| is true if |e| contains no |storeKW|, |e| isn't |v|, |e| isn't |v `choice` e_1|, nor |fail|}

  \end{array}
   $$
  \caption{The Verse calculus: store axioms} \label{fig:store-axioms}
\end{figure}


To add updateable references we extend the system with syntax and rules from figure \ref{fig:store-axioms}.
The |store s e| indicates that |e| should be reduced using the store |s|.
A store is simply a mapping from references to values (one mapping being |cell r v|).
A reference is some opaque type that supports equality (unification) and creation of a new reference.

There are three new operations, |alloc|, |read|, and |write|.
Their interaction with the store can be seen from the axioms.
The |apply1 alloc v| operation creates a new reference and adds a binding with |v| to the store.
The |apply1 read r| operation retrieves the value for reference |r| from the store,
and |applyN write (r, v)| updates the reference |r| with |v| in the store.
Note that these reductions only happen when the store is exactly at the operation.

Most of the store axioms just move the store around from left-to-right, ``visiting'' every subexpression,
and moving on when the subexpression has reduced to a value.
This way every subexpression (in a non-looping program) will get a ``visit'' and all reference operations
can reduce.

The interesting rules are those that cover |splitKW|.
When reducing the |e| in |split e f g| the reduction of |e| is treated as a transaction, i.e.,
if the reduction fails any changes to the store are rolled back.
This is handled by a combination of \rulename{st-split-copy}, \rulename{st-fail}, and \rulename{st-store}.
First, when a |storeKW| is at a |split e f g| then the store is copied to reduce the |e|.
If the evaluation of |e| fails, then rule \rulename{st-fail} will simply abandon the copy of the store.
If the evaluation of |e| is successful then the \rulename{st-value} or \rulename{st-choice} will move the store out of the |splitKW|.
In that position the \rulename{st-store} rule can ``commit'' and throw away the old store, replacing it with the new.
Note that the \rulename{st-split-copy} has a side condition to stop it from firing when it has just fired,
and also when the |e| is a result.

To support limited store operations (e.g., |read|, but not |write|) we can equip the store with a set of currently allowed operations.
We also need some extra expressions that modify this set as the store ``moves'' around.

\lennart{These rewrite rules assume that choices/|fail| are always inside a |splitKW|, otherwise the store can get lost.}


\end{document}
