%include lhs2TeX.fmt
%include lhs2TeX.sty
%include polycode.fmt
%include ../calculus.fmt
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


%% For later:
%% \newcommand{\go}[1]{\mathbin{\smash{\mathop{\xrightarrow{}}\limits_{\smash{\hbox{\raise0.55ex\hbox{$\scriptstyle#1\mskip13mu$}}}}}}}


%format substNP (e) x (v) = "\subst{" e "}{" x "}{" v "}"
%format applyNP (a) (b) = a "\," b
%format Omega = "\mathit{\Omega}"
%format ctrdots = "\cdots"
%format lowdots = "\ldots"
%format commadots = "\;\ldots"

\newcommand{\rinfo}{\mathcal{I}}
\newcommand{\vcOmega}{\mathcal{VC}_\mathit{\Omega}}
\newcommand{\omegavc}{\omega_{\scriptscriptstyle\mathrm{VC}}}
\newcommand{\leqomegavc}{\mathrel{\leq_{\omegavc}}} 

\newcommand{\skewtodo}[1]{\par\vskip6pt\textbf{[#1]}\par\vskip6pt}


\section{Skew Confluence} \label{sec:skew-confluence}

\skewtodo{This is a sketch of an incomplete proof of skew confluence.}

We now consider a version of \versecalc{} that fully supports recursive substitution,
and lifts the pesky no-recursion precondition on the confluence theorem.
Rather than lifting the side condition $|x| \not\in \freevars{v}$
on rule \rulename{subst}, which avoids using a recursive equation for substitution,
we take another approach that we believe leads to a simpler proof:
we introduce the familiar, conventional fixpoint operator |mu x hnf|, but allow it to be applied only to head values, not to general expressions.
A new unification rule \rulename{u-occurs-wrap} can turn a recursive equation into one that is not recursive
(by packaging up the right-hand side within a fixpoint), after which rule \rulename{subst} may be applied.
A corresponding new rule \rulename{u-unwrap} can expand a fixpoint by applying the conventional rewrite rule
$|mu x hnf| \movesto |subst hnf x (mu x hnf)|$. While this rule allows infinite application,
a sensible evaluation strategy would apply this rule ``only when needed''---when it is on either side of an equation,
or when it is in the function position of an application. If we regard any \versecalc{} data structure as tree,
the fixpoint construct in effect can label any subtree in such a way that any node beneath it can have a ``back pointer'' up to the labeled node
by referring to the bound variable that serves as the label. 

The overall plan is to adapt the proof strategy
of \cref{sec:confluence-sketch} and \cref{sec:confluence} for skew confluence.
To do this we need a new result: if two relations are skew confluent
with respect to the same information content function and commute, then their union is also skew confluent.
(In fact, it is not required that the two relations fully commute: a slightly weaker precondition suffices.)
Using this result, our plan
is to (i) define an appropriate information content function for \versecalc{} expressions;
(ii) prove that all the rewrite rules for \versecalc{} are monotonic in this information content function;
(iii) prove that the Unification rules (modified to permit recursive substitution) together with the Normalization rules are skew confluent;
(iv) prove that this combined set of rules commutes in the necessary way with the rules for Application, Elimination, and Choice (which taken together are already known to be confluent);
and (v) then apply our new result to show that the entire set of rewrite rules is skew confluent.
At this time steps (iii) and (iv) are incomplete, so we emphasize that we do not yet have a complete proof of skew confluence for \versecalc{}.

\subsection{Free Variables}


We use the conventional notation $\freevars{|e|}$ to denote the set of variables that occur free
in the expression |e|. Variables are bound by the constructs |def x e|, |lam x e|, and |mu x hnf|.
\Cref{fig:fvs} contains a formal definition of this function for \versecalc{}.

We use the variation $\freevarsoutsidelambdas{|e|}$ to denote the set of variables that occur free
in the expression |e| in at least one position that is not within the body of a
lambda expression\footnote{
  ``$\freevarsoutsidelambdas{\cdot}$'' abbreviates ``free variables outside lambda''}.
As an example, $\freevarsoutsidelambdas{|def x (tup (x, f, g, lam y (tup (x, g, y, z))))|} = \{ f, g \}$ because:
\begin{itemize}
\item |x| is bound by |defKW|, so it is not free. 
\item |f| is free in a position not within the body of a lambda expression.
\item |g| is free in at least one position not within the body of a lambda expression (it also happens to be free in a second position that is within the body of a lambda expression).
\item |y| is bound by |lambda|, so it is not free. 
\item |z| is free, but appears only in a position that is within the body of a lambda expression.
\end{itemize}
\Cref{fig:fvsol} contains a formal definition of this function.
Unlike $\freevars{e}$, $\freevarsoutsidelambdas{|v|}$ is only ever appplied to a value $|v|$.

\subsection{Substitution}

We use the notation |subst e x v| to denote the expression that results from performing standard capture-avoiding substitution of the value |v| for every occurrence of the variable |x| within the expression |e| (it turns out that, for \versecalc{}, substitution of general expressions for variables is not required, only substitution of values for variables).
\Cref{fig:substitution} contains a formal definition of how this notation applies to the \versecalc{} grammar (compare \cite[\S 2.1.15]{Barendregt}).

\subsection{Modified grammar and rewrite rules}

Let us modify the grammar for \versecalc{} to have one additional kind of value, a~\emph{fixpoint} value |mu x hnf|:
\begin{quote}
$
\begin{array}{lr@@{\;}c@@{\;}l}
      \text{Values} & |v| & ::= & |x|  \vb \hnf \vb |mu x hnf| \\
\end{array}
$
\end{quote}
and a modify the set of Unification rewrite rules $\runi$ so that rule \rulename{u-occurs}
\begin{quote}
$
\begin{array}{@@{\quad}lr@@{\;}c@@{\;} l l}
  \rulename{u-occurs} & |x=V^[x]| & \movesto & |fail| & \text{if $|V| \ne |hole|$} \\
\end{array}
$
\end{quote}
is replaced by the two rules\footnote{These two rules allow equations to be recursive through lambda expressions and possibly also tuples, but not through tuples only;
  thus equations such as |f = lam y (tup (y, f))| and |x = (tup (1,  lam y (tup (y, x))))| can be processed by rule \rulename{u-occurs-wrap},
  but the equation |x = (tup (1, x))| can be processed only by rule \rulename{u-occurs-fail}.
  An alternate design using the single rule
  \begin{quote}
$
\begin{array}{@@{\quad}lr@@{\;}c@@{\;} l l}
  \rulename{u-occurs-wrap} & |x=hnf| & \movesto & |x=mu x hnf| & \text{if $|x| \in \freevars{|hnf|}$} \\
\end{array}
$
\end{quote}
plus rule \rulename{u-unwrap} could be used instead to support recursion through tuples only as well as through lambda expressions.}\footnote{
  \rulename{u-occurs-fail} is identical to \rulename{u-occurs} in its effect,
  but it is re-expressed using $\freevarsoutsidelambdas{\cdot}$, which we need anyway for \rulename{u-occurs-wrap}.  Now we can drop the context $V$, which was only
  used in \rulename{u-occurs}.
}
\begin{quote}
$
\begin{array}{@@{\quad}lr@@{\;}c@@{\;} l l}
  \rulename{u-occurs-fail} & |x=hnf; e| & \movesto & |fail| & \text{if $|x| \in \freevarsoutsidelambdas{|hnf|}$} \\
  \rulename{u-occurs-wrap} & |x=hnf; e| & \movesto & |x=mu x hnf; e| & \text{if $|x| \in \freevars{|hnf|}$ and $|x| \not\in \freevarsoutsidelambdas{|hnf|}$} \\
\end{array}
$
\end{quote}
Let us also add this rewrite rule:
\begin{quote}
$
\begin{array}{@@{\quad}lr@@{\;}c@@{\;} l l}
  \rulename{u-unwrap} & |mu x hnf| & \movesto & |subst hnf x (mu x hnf)| \\
\end{array}
$
\end{quote}
As we will see, rule \rulename{u-unwrap} is confluent but not Noetherian.

The resulting grammar is not confluent; in particular, it suffers from the even-odd problem
described in \cref{sec:even-odd} \cite[Example 4.1]{ariola:skew-confluence}.
Therefore we will modify the proof of confluence for $\runi$ to become a proof of \emph{skew confluence} \cite{ariola:skew-confluence},
and then go on to prove that \versecalc{} itself, with this modification, is skew confluent.


\begin{figure}
$
\begin{array}{r@@{\;\;}c@@{\;\;}l@@{\qquad}l}
  \freevars{|x|} & = & \{ |x| \} \\
  \freevars{|k|} & = & \{\,\} \\
  \freevars{|op|} & = & \{\,\} \\
  \freevars{|tup (v_1, commadots, v_n)|} & = & \freevars{|v_1|} \cup \cdots \cup \freevars{|v_n|} \\
  \freevars{|lam x e|} & = & \freevars{|e|} \setminus \{ x \} \\
  \freevars{|mu x e|} & = & \freevars{|e|} \setminus \{ x \} \\
  \freevars{|eq; e|} & = & \freevars{|eq|} \cup \freevars{|e|} \\
  \freevars{|v = e|} & = & \freevars{|v|} \cup \freevars{|e|} \\
  \freevars{|def x e|} & = & \freevars{|e|} \setminus \{ x \} \\
  \freevars{|fail|} & = & \{\,\}  \\
  \freevars{|e_1 choice e_2|} & = & \freevars{|e_1|} \cup \freevars{|e_2|} \\
  \freevars{|v_1 v_2|} & = & \freevars{|v_1|} \cup \freevars{|v_2|} \\
  \freevars{|one e|} & = & \freevars{|e|} \\
  \freevars{|all e|} & = & \freevars{|e|} \\
\end{array}
$
\caption{Definition of the free-variables function $\freevars{|e|}$}\label{fig:fvs}
\Description{Definition of the free-variables function ``fvs''}
\end{figure}

\begin{figure}
$
\begin{array}{r@@{\;\;}c@@{\;\;}l@@{\qquad}l}
  \freevarsoutsidelambdas{|x|} & = & \{ |x| \} \\
  \freevarsoutsidelambdas{|k|} & = & \{\,\} \\
  \freevarsoutsidelambdas{|op|} & = & \{\,\} \\
  \freevarsoutsidelambdas{|tup (v_1, commadots, v_n)|} & = & \freevarsoutsidelambdas{|v_1|} \cup \cdots \cup \freevarsoutsidelambdas{|v_n|} \\
  \freevarsoutsidelambdas{|lam x e|} & = & \{\,\} \\
%   \freevarsoutsidelambdas{|mu x e|} & = & \freevarsoutsidelambdas{|e|} \setminus \{ x \} \\
%  \freevarsoutsidelambdas{|eq; e|} & = & \freevarsoutsidelambdas{|eq|} \cup \freevarsoutsidelambdas{|e|} \\
%  \freevarsoutsidelambdas{|v = e|} & = & \freevarsoutsidelambdas{|v|} \cup \freevarsoutsidelambdas{|e|} \\
%  \freevarsoutsidelambdas{|def x e|} & = & \freevarsoutsidelambdas{|e|} \setminus \{ x \} \\
%  \freevarsoutsidelambdas{|fail|} & = & \{\,\}  \\
%  \freevarsoutsidelambdas{|e_1 choice e_2|} & = & \freevarsoutsidelambdas{|e_1|} \cup \freevarsoutsidelambdas{|e_2|} \\
%  \freevarsoutsidelambdas{|v_1 v_2|} & = & \freevarsoutsidelambdas{|v_1|} \cup \freevarsoutsidelambdas{|v_2|} \\
%  \freevarsoutsidelambdas{|one e|} & = & \freevarsoutsidelambdas{|e|} \\
%  \freevarsoutsidelambdas{|all e|} & = & \freevarsoutsidelambdas{|e|} \\
\end{array}
$
\caption{Definition of the free-variables-outside-lambdas function $\freevarsoutsidelambdas{|v|}$}
\label{fig:fvsol}
\Description{Definition of the free-variables-outside-lambdas function ``fvsol''}
\end{figure}


\begin{figure}
$
\begin{array}{r@@{\;\;}c@@{\;\;}l@@{\qquad}l}
  |subst x x v| & \equiv & |v| \\
  |subst y x v| & \equiv & |y| & \text{if $y \neq x$}  \\
  |subst k x v| & \equiv & |k| \\
  |subst op x v| & \equiv & |op| \\
  |substNP (arr (v_1, dots, v_n)) x v| & \equiv & |arr (subst v_1 x v, dots, subst v_n x v)| \\
  |subst (lam y e) x v| & \equiv & |lam y (subst e x v)| &  \text{if $y \not\in \freevars{|x|, |v|}$ [use $\alpha$]} \\
  |subst (mu y v') x v| & \equiv & |mu y (subst v' x v)| &  \text{if $y \not\in \freevars{|x|, |v|}$ [use $\alpha$]} \\
  |subst (eq; e) x v| & \equiv &  |subst eq x v; subst e x v| \\
  |subst (v' = e) x v| & \equiv &  |subst v' x v = subst e x v| \\
  |subst (def y e) x v| & \equiv & |def y (subst e x v)| &  \text{if $y \not\in \freevars{|x|, |v|}$ [use $\alpha$]} \\
  |subst fail x v| & \equiv & |fail| \\
  |subst (e_1 choice e_2) x v| & \equiv &  |subst e_1 x v`choice`subst e_2 x v| \\
  |subst (apply1N v_1 v_2) x v| & \equiv &  |applyNP (subst v_1 x v) (subst v_2 x v)| \\
  |subst (one e) x v| & \equiv & |one (subst e x v)| \\
  |subst (all e) x v| & \equiv & |all (subst e x v)| \\
\end{array}
$
\caption{Definition of the substitution notation |subst e x v|}\label{fig:substitution}
\Description{Definition of the substitution notation used in this paper}
\end{figure}


\subsection{Unwrapping of Fixpoints is Confluent but not Noetherian}

\begin{lemma}\label{lem:unwrap-confluent}
The rule \rulename{u-unwrap} is confluent.
\end{lemma}

\begin{proof}
  Suppose that $\rstep{\rulename{u-unwrap}}{|e|}{|e_1|}$ and $\rstep{\rulename{u-unwrap}}{|e|}{|e_2|}$ for distinct redexes within $e$.
  
  If the redexes are disjoint, then \cref{lem:disjoint} applies.

  Otherwise, without loss of generality assume the redex for $\rstep{\rulename{u-unwrap}}{|e|}{|e_1|}$ contains the redex for $\rstep{\rulename{u-unwrap}}{|e|}{|e_2|}$;
  let $e$ must be of the form |C_1^[mu x (C_2^[mu y hnf])]| ($\alpha$-conversion may be used to ensure that $x$ and $y$ are distinct variables).

  Then |e_1 = C_1^[substNP (C_2^[mu y hnf]) x (mu x (C_2^[mu y hnf]))]|
  and |e_2 = C_1^[mu x (C_2^[subst hnf y (mu y hnf)])]|.

  From |e_2| we can take just one more \rulename{u-unwrap} step, using the outermost redex |mu x (C_2^[cdots])|, obtaining |e' = C_1^[subst (C_2^[subst hnf y (mu y hnf)]) x (mu x (C_2^[subst hnf y (mu y hnf)]))]|.

  \skewtodo{More to come.}

  Thus we have $\rsteps{\rulename{u-unwrap}}{|e_1|}{|e'|}$ and $\rstep{\rulename{u-unwrap}}{|e_2|}{|e'|}$, so \rulename{u-unwrap} is strongly confluent,
  and therefore by \cref{lem:strong} is confluent.
\end{proof}

To see that \rulename{u-unwrap} is not Noetherian, observe that
$$\ensuremath{|mu x (tup (1,x))| \go{\rulename{u-unwrap}}  |tup (1, mu x (tup (1,x)))| \go{\rulename{u-unwrap}} |tup (1, (tup (1, mu x (tup (1,x)))))| \go{\rulename{u-unwrap}} \cdots}$$
is an unending sequence of reduction steps.

\subsection{Information Content}

We define a second grammar, for a second language $\vcOmega$, by adding one more value |Omega|, which indicates a lack of information as to just what will be computed.
When |Omega| appears in a context where only a value is permitted, it indicates uncertainty as to what value will be provided; 
when |Omega| appears in a context where any expression permitted, it furthermore indicates uncertainty as to how many values will be provided (possibly none).
\begin{quote}
$
\begin{array}{lr@@{\;}c@@{\;}l}
      \text{Values} & |v| & ::= & |x| \vb \hnf \vb |mu x hnf| \vb |Omega| \\
\end{array}
$
\end{quote}

For every term of \versecalc{} there is a corresponding term of $\vcOmega$, identical in structure and appearance and containing no occurrence of |Omega|;
we identify such terms and regard the set of terms of \versecalc{} as simply a subset of the terms of $\vcOmega$.

The definition of substitution (\cref{fig:substitution}) is extended in the expected trivial manner: $|subst Omega x v| \equiv |Omega|$.

\begin{definition}
(after \cite[Definition 2.3]{ariola:skew-confluence})
Let |T| be a set of terms over a signature that includes the constant |Omega|.
Define $\leqomega^T$ be the relation such that $|Omega| \leqomega^T |M|$ for every term $|M| \in |T|$;
then define $\leqomega$ to be the transitive, reflexive, and compatible closure of $\leqomega^T$.
\end{definition}

\Cref{fig:info-rewrite-rules} shows a system $\rinfo$ of rewrite rules on $\vcOmega$.
These may be compared to similar rules for the $\lambda\circ$ calculus \cite[Definition 5.20]{ariola:skew-confluence}.

\begin{figure}
 $
  \begin{array}{l}
\begin{array}{@@{\quad}lr@@{\;}c@@{\;} l l}
  \rulegroupname{Information Content:~$\rinfo${}}
  \rulename{info-fix} & |mu x v| & \movesto & |Omega| \\
  \rulename{info-seq} & |eq; e| & \movesto & |Omega| \\
  \rulename{info-exi} & |def x e| & \movesto & |Omega| \\
  \rulename{info-fail-l} & |fail choice e| & \movesto & |Omega| \\
  \rulename{info-fail-r} & |e choice fail| & \movesto & |Omega| \\
  \rulename{info-choice-omega} & |Omega choice e| & \movesto & |Omega| \\
  \rulename{info-choice-assoc} & |(e_1 choice e_2) choice e_3| & \movesto & |Omega| \\
  \rulename{info-app-omega} & |apply1N Omega v| & \movesto & |Omega| \\
  \rulename{info-app-hnf} & |apply1N hnf v| & \movesto & |Omega| \\
  \rulename{info-one} & |one e| & \movesto & |Omega| \\
  \rulename{info-all} & |all e| & \movesto & |Omega| \\
\end{array}
\end{array}
$
\caption{Rewrite rules on $\vcOmega$ for defining the function $\omegavc(e)$}\label{fig:info-rewrite-rules}
\Description{Rewrite rules on $\vcOmega$ for defining the function $\omegavc(e)$}
\end{figure}

\begin{lemma}\label{lem:critpairs}\cite[Lemma 3.1]{Huet80}
The relation $\go{\rr}$ is locally confluent iff for every critical pair |(e_1, e_2)| of $\rr$, |e_1| and |e_2| can be \emph{joined}---that is, there exists |e_3| such that
 $|e_1| \go{\rr} |e_3|$ and  $|e_2| \go{\rr} |e_3|$.
\end{lemma}


\begin{lemma}[$\rinfo$-Confluent] \label{lem:rinfo-c}
$\rinfo$ is locally confluent and Noetherian; therefore $\rinfo$ is confluent.
\end{lemma}

\begin{proof}
  Consider all critical pairs of $\rinfo$:
  \begin{itemize}
\item  Rules \rulename{info-fail-l} and \rulename{info-fail-r} produce the critical pair |(Omega, Omega)|.

\item  Rules \rulename{info-fail-l} and \rulename{info-choice-omega} produce no critical pairs.

\item  Rules \rulename{info-fail-l} and \rulename{info-choice-assoc} produce the critical pair |(Omega choice e, Omega)|.

\item  Rules \rulename{info-fail-r} and \rulename{info-choice-omega} produce the critical pair |(Omega, Omega)|.

\item  Rules \rulename{info-fail-r} and \rulename{info-choice-assoc} produce the critical pairs |(Omega, Omega)| and |(Omega choice e, Omega)|.

\item  Rules \rulename{info-choice-omega} and \rulename{info-choice-assoc} produce the critical pair |(Omega choice e, Omega)|.

\item  No other pairs of rules produce any critical pairs.
\end{itemize}
The critical pair |(Omega, Omega)| can be trivially joined at |Omega|.
The critical pair |(Omega choice e, Omega)| can be joined at |Omega| in one step by using rule \rulename{info-choice-omega} on |Omega choice e|.

All critical pairs can be joined; therefore by \cref{lem:critpairs} $\rinfo$ is locally confluent.

Let the size of a term of $\vcOmega$ be the number of tokens it contains other than parentheses.
Each of the rewrite rules in \cref{fig:info-rewrite-rules} strictly decreases the number of such tokens.
Because any given term has a finite number $n$ of such tokens, and the number of tokens cannot be less than zero,
for every term every rewriting sequence from that term can have no more than $n$ steps. So $\rinfo$ is bounded and therefore Noetherian.

Because $\rinfo$ is locally confluent and Noetherian, it is confluent by Newman's lemma.
\end{proof}

Because $\rinfo$ is confluent and Noetherian, it follows that $\rinfo$ defines unique normal forms for $\vcOmega$. Therefore we are justified in defining
$\omegavc(e)$ to be the function that takes any term in $\vcOmega$ and returns the term that is its normal form under $\rinfo$.

\begin{definition}
The comparison $e \leqomegavc e'$ is defined to mean $\omegavc(e) \leqomega \omegavc(e')$.
\end{definition}

\skewtodo{Need to prove that $\leqomegavc$ is monotonic with respect to $\leqomega$; this should be routine \cite[Proposition 5.21]{ariola:skew-confluence}.}

\begin{lemma}[\versecalc{} monotonic] \label{lem:vc-monotonic}
Every rewrite rule in \versecalc{} is monotonic with respect to $\leqomegavc$.
\end{lemma}

\begin{proof}
  For every rewrite rule in $\rapp \cup \runi \cup \rcon \cup \rgc \cup \rspc$,
  the left-hand side is an expression that is mapped to |Omega| by the function $\omegavc$,
  and no matter what expression |e| is the result of applying $\omegavc$ to the right-hand side,
  we have $|Omega| \leqomega |e|$.
 \end{proof}

\subsection{Preliminaries about Skew Confluence}


\begin{comment}
\subsection{Overview of Skew Confluence (VERSION 1)} \label{sec:skew-overview}


If a reduction relation $\rr$ is confluent, then for all |e|, |e_1|, and |e_2|,
if $\rstep{\rr}{|e|}{|e_1|}$ and $\rstep{\rr}{|e|}{|e_2|}$ then there exists |e'| such that
if $\rstep{\rr}{|e_1|}{|e'|}$ and $\rstep{\rr}{|e_2|}{|e'|}$.
This is useful because the $\rr$-normal form (if it exists) of any expression is if $\rr$ is confluent.

But what can we say if $\rr$ is not confluent---for example, what if $\rr$ has the even-odd problem
discussed in \cref{sec:even-odd}? Ariola and Blom introduced a notion of \emph{infinite normal form},
which is a (possibly infinite) set of (possibly \emph{erased}) terms rather than a single term.
We summarize this idea, using our own terminology, as follows:
\begin{itemize}
\item Let an \emph{erasure} of a term be a copy in which one or more subterms (possibly the entire term)
  have been replaced with |Omega|, a special term that means ``unknown'' or ``we don't know yet.''
\item We can compare possibly erased terms with the partial order $\leqomega$, which is the transitive,
  reflexive, and compatible closure of the relation in which |Omega| is less than any other term.
  Observe that if |e'| is an erasure of |e| then $|e'| \leqomega |e|$.
\item Define the $\rr$-\emph{information content} $\omegarr(|e|)$ of an expression |e| to be the unique erasure of |e| in which \emph{every} redex
  has been replaced by |Omega|. Any non-|Omega| structure in the result is therefore ``permanent'':
  no further reductions under $\rr$ can alter that structure.
\item For any expression |e|, consider the (possibly infinite) set $A$ of all its $\rr$-reducts, that is, \emph{all} expressions |e'| such that
  $\rstep{\rr}{e}{e'}$, then let $B = \{ \omegarr(|e|) \mid |e| \in A\}$.
  The \emph{infinite $\rr$-normal form} of |e| is a set $C$, the \emph{downward closure} of $B$ under $\leqomega$; that is,
  for every element of $B$, $C$ contains not only that element but also every possible erasure of that element.
\end{itemize}
It is the step of taking a downward closure under $\leqomega$ that makes it possible to prove that infinite normal forms
are unique for certain reduction relations.

But working with these infinite sets directly is tricky. Fortunately, there is a simpler path, because Ariola and Blom also prove
an important theorem: a reduction relation that is \emph{monotonic} (meaning that no rewrite rule ever decreases the information content)
has unique infinite normal forms if and only if the reduction relation is \emph{skew confluent} \cite[Theorem 5.4]{ariola:skew-confluence}---and skew
confluence is much easier to prove, using techniques that do not involve infinite sets, but are fairly similar to proofs of ordinary confluence
that use commutative diagrams and case analysis.

Here, then, is the idea behind skew confluence: for all |e|, |e_1|, and |e_2|,
if $\rstep{\rr}{|e|}{|e_1|}$ and $\rstep{\rr}{|e|}{|e_2|}$, then there may not be any |e'| such that
if $\rstep{\rr}{|e_1|}{|e'|}$ and $\rstep{\rr}{|e_2|}{|e'|}$. Instead, you can choose either of |e_1| and |e_2| (let's say |e_2|),
and then there does exists some |e'| such that $\rstep{\rr}{|e_1|}{|e'|}$ and |e'| has at least as much information content as |e_2|,
that is, $\omegarr(|e_2|) \leqomega \omegarr(|e'|)$. This works both ways, by symmetry, because the choice of |e_1| or |e_2| is arbitrary.

In the case of the even-odd problem discussed in \cref{sec:even-odd}, further reductions on the two expressions
|def y (y == lam z (tup (1, y); tup (1,y))| and |def x (x = tup (1, lam z x)); x| might produce longer expressions such as
|def y (y == lam z (tup (1, y)); tup (1, lam z (tup (1, lam z (tup (1, y))))))|
and |def x (x = tup (1, lam z x); tup (1, lam z (tup (1, lam z (tup (1, lam z x))))))|.
One can see that each of these expressions
will produce a value of the form |tup (1, lam z (tup (1, lam z (tup (1, ctrdots))))))|. Confluence cannot be achieved,
because no matter how many times one substitutes for |y| in the first expression or for |x| in the second expression,
the first expression will always have one more occurrence of ``|1|'' than ``|lamhdr z|'' but the second expression
will always have the same number of occurrences of ``|1|'' than ``|lamhdr z|''; on the other hand, no matter
how long one expression becomes through substitution, the other one can be made even longer,
so as to have at least as much information content (and in fact more, in this example). The theory of skew confluence makes this idea precise.

Another important fact is that if a reduction relation is confluent and monotonic, then it is skew confluent
\cite[Corollary 5.5]{ariola:skew-confluence}. In \cref{sec:skew-confluence} we prove a new result: if two relations are skew confluent
with respect to the same information content function and commute, then their union is also skew confluent.
(In fact, it is not required that the two relations fully commute; a slightly weaker precondition suffices.)
This provides a detailed strategy for reusing parts of the proof of confluence for \versecalc{} sketched in \cref{sec:confluence-sketch}
and detailed in \cref{sec:confluence}: our plan is to (i) define an appropriate information content function for \versecalc{} expressions;
(ii) prove that all the rewrite rules for \versecalc{} are monotonic in this information content function;
(iii) prove that the Unification rules (modified to permit recursive substitution) together with the Normalization rules are skew confluent;
(iv) prove that this combined set of rules commutes in the necessary way with the rules for Application, Elimination, and Choice (which taken together are already known to be confluent);
and (v) then apply our new result to show that the entire set of rewrite rules is skew confluent.
At this time steps (iii) and (iv) are incomplete, so we emphasize that \emph{we do not yet have a complete proof of skew confluence for \versecalc{}}.
\end{comment}

\begin{comment}

further reductions on the two terms
|def y (y == lam z (tup (1, y); tup (1,y))| and |def x (x = tup (1, lam z x)); x| might produce longer terms such as
|def y (y == lam z (tup (1, y)); tup (1, lam z (tup (1, lam z (tup (1, y))))))|
and |def x (x = tup (1, lam z x); tup (1, lam z (tup (1, lam z (tup (1, lam z x))))))|.
One can see that each of these terms
will produce a value of the form |tup (1, lam z (tup (1, lam z (tup (1, ctrdots))))))|. Confluence cannot be achieved,
because no matter how many times one substitutes for |y| in the first expression or for |x| in the second expression,
the first expression will always have one more occurrence of ``|1|'' than ``|lamhdr z|'' but the second expression
will always have the same number of occurrences of ``|1|'' than ``|lamhdr z|''; on the other hand, no matter
how long one expression becomes through substitution, the other one can be made even longer,
so as to have at least as much information content (and in fact more, in this example). The theory of skew confluence makes this idea precise.

Old...
and either term can be further reduced so as to contain \emph{all} the permanent structure of the other.
This provides a notion of convergence even for infinite reduction sequences.

In the case of the even-odd problem discussed in \cref{sec:even-odd}, further reductions on the two terms
|def y (y == lam z (tup (1, y); tup (1,y))| and |def x (x = tup (1, lam z x)); x| might produce longer terms such as
|def y (y == lam z (tup (1, y)); tup (1, lam z (tup (1, lam z (tup (1, y))))))|
and |def x (x = tup (1, lam z x); tup (1, lam z (tup (1, lam z (tup (1, lam z x))))))|.
One can see that each of these terms
will produce a value of the form |tup (1, lam z (tup (1, lam z (tup (1, ctrdots))))))|. Confluence cannot be achieved,
because no matter how many times one substitutes for |y| in the first expression or for |x| in the second expression,
the first expression will always have one more occurrence of ``|1|'' than ``|lamhdr z|'' but the second expression
will always have the same number of occurrences of ``|1|'' than ``|lamhdr z|''; on the other hand, no matter
how long one expression becomes through substitution, the other one can be made even longer,
so as to have at least as much information content (and in fact more, in this example). The theory of skew confluence makes this idea precise.
\end{comment}


Why use skew confluence? Ordinary confluence is useful because if term |e| has an $\rr$-normal form, then that normal form is \emph{unique} if $\rr$ is confluent.
Ariola and Blom define a related notion, which we will refer to as \emph{$\rr$-skew-normal form}\footnote{Ariola and Blom call it the ``infinite normal form''; this is a bit misleading because in fact not all such forms are infinite.}, and prove that under certain conditions, if term |e| has an $\rr$-skew-normal form, then that normal form is unique if $\rr$ is skew confluent.

A  $\rr$-skew-normal form is not a single term, but rather a possibly infinite set of \emph{erased} terms.
We summarize this idea, using our own terminology, as follows:
\begin{itemize}
\item Let an \emph{erasure} of a term be a copy in which some number of subterms (possibly zero, and possibly the entire term)
  have been replaced with |Omega|, a special term that means ``unknown'' or ``we don't know yet.''
\item We can compare erased terms with the partial order $\leqomega$, which is the transitive,
  reflexive, and compatible closure of the relation in which |Omega| is less than any other term.
  Observe that if |e'| is any erasure of |e| (including |e| itself) then $|e'| \leqomega |e|$.
\item Define the $\rr$-\emph{information content} $\omegarr(|e|)$ of a term |e| to be the unique erasure of |e| in which \emph{every} redex
  has been replaced by |Omega|. Any non-|Omega| structure in the result is therefore ``permanent'':
  no further reductions under $\rr$ can alter that structure.
\item Define the \emph{downward closure} $\downclose A$ of a set of terms $A$ to be the set of all elements of $A$ and all possible erasures of those elements,
  that is, $\downclose A = \{ |e'| \mid |e| \in A, |e'| \leqomega \}$.
\item
  Define the \emph{$\rr$-skew-normal form} of |e| to be the downward closure of the set of information contents of all possible $\rr$-reducts of |e|,
  that is, $\downclose \{ \omega(|e'|) \mid \rsteps{\rr}{e}{e'} \}$.
\end{itemize}
Taking the downward closure with respect to $\leqomega$ is crucial; without that step, it would not be possible to prove that skew-normal forms
are unique for certain reduction relations.

Skew confluence is a natural extension of confluence, in this sense: if $\rr$ is skew confluent,
then an expression |e| has a unique $\rr$-normal form
if and only if its $\rr$-skew-normal form is the (finite) set consisting of all possible erasures of that $\rr$-normal form.
(For example, |tup (1,2)| is the unique normal form of |def x (x=2; tup (1,x))|, and the  $\rr$-skew-normal form of that same term
is $\{ |tup (1,2)|, |tup (Omega,2)|, |tup (1,Omega)|, |tup (Omega,Omega)|, |Omega| \}$.)

But working with possibly infinite sets directly is tricky. Fortunately, there is a simpler path, because Ariola and Blom prove
an important theorem:
Define the partial order $|e| \preceqomegarr |e'|$ to mean $\omegarr(|e|)\leqomega\omegarr(|e'|)$; then
a reduction relation that is \emph{monotonic} in $\preceqomegarr$ ($\rstep{\rr}{e}{e'} \implies e \preceqomegarr e'$)
has unique skew-normal forms if and only if the reduction relation is skew confluent \cite[Theorem 5.4]{ariola:skew-confluence}---and skew
confluence is much easier to prove, using techniques that do not involve possibly infinite sets, but are fairly similar to proofs of ordinary confluence
that use commutative diagrams and case analysis. They also prove that if a reduction relation is confluent and monotonic, then it is skew confluent
\cite[Corollary 5.5]{ariola:skew-confluence}.




\begin{definition}
  Reduction relation $\rr$ over the set of terms $T$ is \emph{skew confluent} using quasi order $\preceq$ if for all $a, b, c \in T$,
  if $a \gos{\rr} b$ and $a \gos{\rr} c$, there exists $d\in T$ such that $b \gos{\rr} d$ and $c \preceq d$. As a diagram:
  
% https://q.uiver.app/?q=WzAsNCxbMCwwLCJhIl0sWzEsMCwiYiJdLFsxLDEsImQiXSxbMCwxLCJjIl0sWzAsMSwiXFxyciIsMCx7InN0eWxlIjp7ImhlYWQiOnsibmFtZSI6ImVwaSJ9fX1dLFsxLDIsIlxccnIiLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJkb3R0ZWQifSwiaGVhZCI6eyJuYW1lIjoiZXBpIn19fV0sWzAsMywiXFxyciIsMix7InN0eWxlIjp7ImhlYWQiOnsibmFtZSI6ImVwaSJ9fX1dLFszLDIsIlxccnIiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJkb3R0ZWQifSwiaGVhZCI6eyJuYW1lIjoiZXBpIn19fV0sWzIsMywiXFxwcmVjZXEiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJub25lIn0sImhlYWQiOnsibmFtZSI6ImVwaSJ9fX1dLFs2LDUsIiIsMSx7InNob3J0ZW4iOnsic291cmNlIjoyMCwidGFyZ2V0IjoyMH0sInN0eWxlIjp7ImJvZHkiOnsibmFtZSI6Im5vbmUifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dXQ==
\[\begin{tikzcd}
	a & b \\
	c & d
	\arrow["\rr", two heads, from=1-1, to=1-2]
	\arrow[""{name=0, anchor=center, inner sep=0}, "\rr", dotted, two heads, from=1-2, to=2-2]
	\arrow[""{name=1, anchor=center, inner sep=0}, "\rr"', two heads, from=1-1, to=2-1]
	\arrow["\rr"', dotted, two heads, from=2-1, to=2-2]
	\arrow["\preceq"', no body, two heads, from=2-2, to=2-1]
	\arrow[draw=none, from=1, to=0]
\end{tikzcd}\]
\end{definition}

\skewtodo{More to come.}

\subsection{New Lemmas about Skew Confluence}

\begin{lemma}
If relation $\rr$ is monotonic in some quasi order $\preceq$ and confluent, then it is skew confluent using $\preceq$.
\end{lemma}
\begin{proof}
  By the definition of confluence,
  % https://q.uiver.app/?q=WzAsNCxbMCwwLCJhIl0sWzEsMCwiYyJdLFswLDEsImIiXSxbMSwxLCJkIl0sWzAsMiwiXFxyciIsMix7InN0eWxlIjp7ImhlYWQiOnsibmFtZSI6ImVwaSJ9fX1dLFswLDEsIlxccnIiLDAseyJzdHlsZSI6eyJoZWFkIjp7Im5hbWUiOiJlcGkifX19XSxbMSwzLCJcXHJyIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiZG90dGVkIn0sImhlYWQiOnsibmFtZSI6ImVwaSJ9fX1dLFsyLDMsIlxccnIiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJkb3R0ZWQifSwiaGVhZCI6eyJuYW1lIjoiZXBpIn19fV1d
\[\begin{tikzcd}
	a & c \\
	b & d
	\arrow["\rr"', two heads, from=1-1, to=2-1]
	\arrow["\rr", two heads, from=1-1, to=1-2]
	\arrow["\rr", dotted, two heads, from=1-2, to=2-2]
	\arrow["\rr"', dotted, two heads, from=2-1, to=2-2]
\end{tikzcd}\]
Because $\rr$ is monotonic, $\gos{\rr} \subset \inccon{\preceq}{\rr}$, therefore 
% https://q.uiver.app/?q=WzAsNCxbMCwwLCJhIl0sWzEsMCwiYyJdLFswLDEsImIiXSxbMSwxLCJkIl0sWzAsMiwiXFxyciIsMix7InN0eWxlIjp7ImhlYWQiOnsibmFtZSI6ImVwaSJ9fX1dLFswLDEsIlxccnIiLDAseyJzdHlsZSI6eyJoZWFkIjp7Im5hbWUiOiJlcGkifX19XSxbMSwzLCJcXHJyIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiZG90dGVkIn0sImhlYWQiOnsibmFtZSI6ImVwaSJ9fX1dLFsyLDMsIlxccnIiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJkb3R0ZWQifSwiaGVhZCI6eyJuYW1lIjoiZXBpIn19fV0sWzMsMiwiXFxwcmVjZXEiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJub25lIn0sImhlYWQiOnsibmFtZSI6ImVwaSJ9fX1dXQ==
\[\begin{tikzcd}
	a & c \\
	b & d
	\arrow["\rr"', two heads, from=1-1, to=2-1]
	\arrow["\rr", two heads, from=1-1, to=1-2]
	\arrow["\rr", dotted, two heads, from=1-2, to=2-2]
	\arrow["\rr"', dotted, two heads, from=2-1, to=2-2]
	\arrow["\preceq"', no body, two heads, from=2-2, to=2-1]
\end{tikzcd}\]
\end{proof}

\begin{definition}
  Let reduction relation $\rr$ be monotonic in quasi order $\preceq$ and skew confluent using $\preceq$.
  Let reduction relation $\rr_\preceq^{\leftarrow}$ be defined by a set of rewrite rules that are converses of those rewrite rules of $\rr$ whose converses are used in the proof that $\rr$ is skew
  confluent.
  Then we say that $\rr$ is \emph{skew confluent using $\preceq$ and  $\rr_\preceq^{\leftarrow}$}.
\end{definition}


\begin{lemma}\label{lem:four-diagrams}
  Let relation $\rr$ be monotonic in quasi order $\preceq$ and skew confluent using $\preceq$ and $\rr_\preceq^{\leftarrow}$;
  similarly let $\rs$ be monotonic in the same quasi order $\preceq$ and skew confluent using $\preceq$ and $\rs_\preceq^{\leftarrow}$.
  Suppose furthermore that $\rr$ and $\rs$ commute and that $\rr$ and $\rs_\preceq^{\leftarrow}$ commute. 
  Then the following four diagrams hold:

% https://q.uiver.app/?q=WzAsNCxbMCwwLCJhIl0sWzEsMCwiYiJdLFsxLDEsImQiXSxbMCwxLCJjIl0sWzAsMSwiXFxyciIsMl0sWzEsMiwiXFxyciIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImRvdHRlZCJ9LCJoZWFkIjp7Im5hbWUiOiJlcGkifX19XSxbMCwzLCJcXHJyIiwyLHsic3R5bGUiOnsiaGVhZCI6eyJuYW1lIjoiZXBpIn19fV0sWzMsMiwiXFxyclxcY3VwXFxycyIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImRvdHRlZCJ9LCJoZWFkIjp7Im5hbWUiOiJlcGkifX19XSxbMiwzLCJcXHByZWNlcSIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6Im5vbmUifSwiaGVhZCI6eyJuYW1lIjoiZXBpIn19fV0sWzYsNSwiIiwxLHsic2hvcnRlbiI6eyJzb3VyY2UiOjIwLCJ0YXJnZXQiOjIwfSwic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
\[\begin{tikzcd}
	a & b \\
	c & d
	\arrow["\rr"', from=1-1, to=1-2]
	\arrow[""{name=0, anchor=center, inner sep=0}, "\rr", dotted, two heads, from=1-2, to=2-2]
	\arrow[""{name=1, anchor=center, inner sep=0}, "\rr"', two heads, from=1-1, to=2-1]
	\arrow["\rr\cup\rs"', dotted, two heads, from=2-1, to=2-2]
	\arrow["\preceq"', no body, two heads, from=2-2, to=2-1]
	\arrow[draw=none, from=1, to=0]
      \end{tikzcd}
      \qquad\qquad
% https://q.uiver.app/?q=WzAsNCxbMCwwLCJhIl0sWzEsMCwiYiJdLFsxLDEsImQiXSxbMCwxLCJjIl0sWzAsMSwiXFxycyIsMl0sWzEsMiwiXFxyciIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImRvdHRlZCJ9LCJoZWFkIjp7Im5hbWUiOiJlcGkifX19XSxbMCwzLCJcXHJyIiwyLHsic3R5bGUiOnsiaGVhZCI6eyJuYW1lIjoiZXBpIn19fV0sWzMsMiwiXFxyclxcY3VwXFxycyIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImRvdHRlZCJ9LCJoZWFkIjp7Im5hbWUiOiJlcGkifX19XSxbMiwzLCJcXHByZWNlcSIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6Im5vbmUifSwiaGVhZCI6eyJuYW1lIjoiZXBpIn19fV0sWzYsNSwiIiwxLHsic2hvcnRlbiI6eyJzb3VyY2UiOjIwLCJ0YXJnZXQiOjIwfSwic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
\begin{tikzcd}
	a & b \\
	c & d
	\arrow["\rs"', from=1-1, to=1-2]
	\arrow[""{name=0, anchor=center, inner sep=0}, "\rr", dotted, two heads, from=1-2, to=2-2]
	\arrow[""{name=1, anchor=center, inner sep=0}, "\rr"', two heads, from=1-1, to=2-1]
	\arrow["\rr\cup\rs"', dotted, two heads, from=2-1, to=2-2]
	\arrow["\preceq"', no body, two heads, from=2-2, to=2-1]
	\arrow[draw=none, from=1, to=0]
      \end{tikzcd}
      \qquad\qquad
    % https://q.uiver.app/?q=WzAsNCxbMCwwLCJhIl0sWzEsMCwiYiJdLFsxLDEsImQiXSxbMCwxLCJjIl0sWzEsMCwiXFxyciJdLFsxLDIsIlxccnIiLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJkb3R0ZWQifSwiaGVhZCI6eyJuYW1lIjoiZXBpIn19fV0sWzAsMywiXFxyciIsMix7InN0eWxlIjp7ImhlYWQiOnsibmFtZSI6ImVwaSJ9fX1dLFszLDIsIlxccnJcXGN1cFxccnMiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJkb3R0ZWQifSwiaGVhZCI6eyJuYW1lIjoiZXBpIn19fV0sWzIsMywiXFxwcmVjZXEiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJub25lIn0sImhlYWQiOnsibmFtZSI6ImVwaSJ9fX1dLFs2LDUsIiIsMSx7InNob3J0ZW4iOnsic291cmNlIjoyMCwidGFyZ2V0IjoyMH0sInN0eWxlIjp7ImJvZHkiOnsibmFtZSI6Im5vbmUifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dXQ==
\begin{tikzcd}
	a & b \\
	c & d
	\arrow["\rr", from=1-2, to=1-1]
	\arrow[""{name=0, anchor=center, inner sep=0}, "\rr", dotted, two heads, from=1-2, to=2-2]
	\arrow[""{name=1, anchor=center, inner sep=0}, "\rr"', two heads, from=1-1, to=2-1]
	\arrow["\rr\cup\rs"', dotted, two heads, from=2-1, to=2-2]
	\arrow["\preceq"', no body, two heads, from=2-2, to=2-1]
	\arrow[draw=none, from=1, to=0]
      \end{tikzcd}
      \qquad\qquad
% https://q.uiver.app/?q=WzAsNCxbMCwwLCJhIl0sWzEsMCwiYiJdLFsxLDEsImQiXSxbMCwxLCJjIl0sWzEsMCwiXFxycyJdLFsxLDIsIlxccnIiLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJkb3R0ZWQifSwiaGVhZCI6eyJuYW1lIjoiZXBpIn19fV0sWzAsMywiXFxyciIsMix7InN0eWxlIjp7ImhlYWQiOnsibmFtZSI6ImVwaSJ9fX1dLFszLDIsIlxccnJcXGN1cFxccnMiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJkb3R0ZWQifSwiaGVhZCI6eyJuYW1lIjoiZXBpIn19fV0sWzIsMywiXFxwcmVjZXEiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJub25lIn0sImhlYWQiOnsibmFtZSI6ImVwaSJ9fX1dLFs2LDUsIiIsMSx7InNob3J0ZW4iOnsic291cmNlIjoyMCwidGFyZ2V0IjoyMH0sInN0eWxlIjp7ImJvZHkiOnsibmFtZSI6Im5vbmUifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dXQ==
\begin{tikzcd}
	a & b \\
	c & d
	\arrow["\rs", from=1-2, to=1-1]
	\arrow[""{name=0, anchor=center, inner sep=0}, "\rr", dotted, two heads, from=1-2, to=2-2]
	\arrow[""{name=1, anchor=center, inner sep=0}, "\rr"', two heads, from=1-1, to=2-1]
	\arrow["\rr\cup\rs"', dotted, two heads, from=2-1, to=2-2]
	\arrow["\preceq"', no body, two heads, from=2-2, to=2-1]
	\arrow[draw=none, from=1, to=0]
      \end{tikzcd}\]
  \end{lemma}
  \begin{proof}
    
    (1) Because $\rr$ is skew confluent, we have

    % https://q.uiver.app/?q=WzAsNCxbMCwwLCJhIl0sWzEsMCwiYiJdLFsxLDEsImQiXSxbMCwxLCJjIl0sWzAsMSwiXFxyciIsMl0sWzEsMiwiXFxyciIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImRvdHRlZCJ9LCJoZWFkIjp7Im5hbWUiOiJlcGkifX19XSxbMCwzLCJcXHJyIiwyLHsic3R5bGUiOnsiaGVhZCI6eyJuYW1lIjoiZXBpIn19fV0sWzMsMiwiXFxyciIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImRvdHRlZCJ9LCJoZWFkIjp7Im5hbWUiOiJlcGkifX19XSxbMiwzLCJcXHByZWNlcSIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6Im5vbmUifSwiaGVhZCI6eyJuYW1lIjoiZXBpIn19fV0sWzYsNSwiIiwxLHsic2hvcnRlbiI6eyJzb3VyY2UiOjIwLCJ0YXJnZXQiOjIwfSwic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
\[\begin{tikzcd}
	a & b \\
	c & d
	\arrow["\rr"', from=1-1, to=1-2]
	\arrow[""{name=0, anchor=center, inner sep=0}, "\rr", dotted, two heads, from=1-2, to=2-2]
	\arrow[""{name=1, anchor=center, inner sep=0}, "\rr"', two heads, from=1-1, to=2-1]
	\arrow["\rr"', dotted, two heads, from=2-1, to=2-2]
	\arrow["\preceq"', no body, two heads, from=2-2, to=2-1]
	\arrow[draw=none, from=1, to=0]
      \end{tikzcd}\]

\noindent    Because $\inccon{\preceq}{\rr} \subset \inccon{\preceq}{\rr\cup\rs}$, the first diagram follows.

        (2) Because $\rr$ and $\rs$ commute, we have

% https://q.uiver.app/?q=WzAsNCxbMCwwLCJhIl0sWzEsMCwiYiJdLFsxLDEsImQiXSxbMCwxLCJjIl0sWzAsMSwiXFxycyIsMl0sWzEsMiwiXFxyciIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImRvdHRlZCJ9LCJoZWFkIjp7Im5hbWUiOiJlcGkifX19XSxbMCwzLCJcXHJyIiwyLHsic3R5bGUiOnsiaGVhZCI6eyJuYW1lIjoiZXBpIn19fV0sWzMsMiwiXFxycyIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImRvdHRlZCJ9LCJoZWFkIjp7Im5hbWUiOiJlcGkifX19XSxbNiw1LCIiLDEseyJzaG9ydGVuIjp7InNvdXJjZSI6MjAsInRhcmdldCI6MjB9LCJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJub25lIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XV0=
\[\begin{tikzcd}
	a & b \\
	c & d
	\arrow["\rs"', from=1-1, to=1-2]
	\arrow[""{name=0, anchor=center, inner sep=0}, "\rr", dotted, two heads, from=1-2, to=2-2]
	\arrow[""{name=1, anchor=center, inner sep=0}, "\rr"', two heads, from=1-1, to=2-1]
	\arrow["\rs"', dotted, two heads, from=2-1, to=2-2]
	\arrow[draw=none, from=1, to=0]
\end{tikzcd}\]

\noindent    Because $\gos{\rs} \subset \inccon{\preceq}{\rr\cup\rs}$, the second diagram follows.

(3) The following diagram clearly holds if the sequence of reduction steps from |b| to |d| is the same as the sequence of reduction steps from |b| to |a| to |d|:

% https://q.uiver.app/?q=WzAsNCxbMCwwLCJhIl0sWzEsMCwiYiJdLFsxLDEsImQiXSxbMCwxLCJjIl0sWzEsMCwiXFxyciJdLFsxLDIsIlxccnIiLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJkb3R0ZWQifSwiaGVhZCI6eyJuYW1lIjoiZXBpIn19fV0sWzAsMywiXFxyciIsMix7InN0eWxlIjp7ImhlYWQiOnsibmFtZSI6ImVwaSJ9fX1dLFszLDIsIlxcZXF1aXYiLDEseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJub25lIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbNiw1LCIiLDEseyJzaG9ydGVuIjp7InNvdXJjZSI6MjAsInRhcmdldCI6MjB9LCJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJub25lIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XV0=
\[\begin{tikzcd}
	a & b \\
	c & d
	\arrow["\rr", from=1-2, to=1-1]
	\arrow[""{name=0, anchor=center, inner sep=0}, "\rr", dotted, two heads, from=1-2, to=2-2]
	\arrow[""{name=1, anchor=center, inner sep=0}, "\rr"', two heads, from=1-1, to=2-1]
	\arrow["\equiv"{description}, draw=none, from=2-1, to=2-2]
	\arrow[draw=none, from=1, to=0]
      \end{tikzcd}\]

    \noindent    Because $\equiv \subset \inccon{\preceq}{\rr\cup\rs}$, the third diagram follows.

    (4) Because $\rr$ and $\rs_\preceq^{\leftarrow}$ commute, we have

    % https://q.uiver.app/?q=WzAsNCxbMCwwLCJhIl0sWzEsMCwiYiJdLFsxLDEsImQiXSxbMCwxLCJjIl0sWzEsMCwiXFxycyJdLFsxLDIsIlxccnIiLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJkb3R0ZWQifSwiaGVhZCI6eyJuYW1lIjoiZXBpIn19fV0sWzAsMywiXFxyciIsMix7InN0eWxlIjp7ImhlYWQiOnsibmFtZSI6ImVwaSJ9fX1dLFsyLDMsIlxccnMiLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJkb3R0ZWQifSwiaGVhZCI6eyJuYW1lIjoiZXBpIn19fV0sWzYsNSwiIiwxLHsic2hvcnRlbiI6eyJzb3VyY2UiOjIwLCJ0YXJnZXQiOjIwfSwic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
\[\begin{tikzcd}
	a & b \\
	c & d
	\arrow["\rs", from=1-2, to=1-1]
	\arrow[""{name=0, anchor=center, inner sep=0}, "\rr", dotted, two heads, from=1-2, to=2-2]
	\arrow[""{name=1, anchor=center, inner sep=0}, "\rr"', two heads, from=1-1, to=2-1]
	\arrow["\rs", dotted, two heads, from=2-2, to=2-1]
	\arrow[draw=none, from=1, to=0]
      \end{tikzcd}\]

        \noindent    Because $\ungos{\rs} \subset \inccon{\preceq}{\rr\cup\rs}$, the fourth diagram follows.

        \skewtodo{It may turn out to be impossible to prove for our specific application that $\rr$ and $\rs_\preceq^{\leftarrow}$ commute.
          In that case, it may be necessary to use a more complicated or more subtle precondition.
        The important thing is to prove the fourth diagram somehow.}
\end{proof}






  
\begin{lemma}\label{lem:skewrowinduction}
  Let relation $\rr$ be monotonic in quasi order $\preceq$ and skew confluent using $\preceq$ and $\rr_\preceq^{\leftarrow}$;
  similarly let $\rs$ be monotonic in the same quasi order $\preceq$ and skew confluent using $\preceq$ and $\rs_\preceq^{\leftarrow}$.
  Suppose furthermore that $\rr$ and $\rs$ commute and that $\rr$ and $\rs_\preceq^{\leftarrow}$ commute. 
  Then

  % https://q.uiver.app/?q=WzAsNCxbMCwwLCJhIl0sWzEsMCwiYiJdLFsxLDEsImQiXSxbMCwxLCJjIl0sWzAsMSwiXFxyclxcY3VwXFxycyIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImRvdHRlZCJ9LCJoZWFkIjp7Im5hbWUiOiJlcGkifX19XSxbMSwwLCJcXHByZWNlcSIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6Im5vbmUifSwiaGVhZCI6eyJuYW1lIjoiZXBpIn19fV0sWzEsMiwiXFxyciIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImRvdHRlZCJ9LCJoZWFkIjp7Im5hbWUiOiJlcGkifX19XSxbMCwzLCJcXHJyIiwyLHsic3R5bGUiOnsiaGVhZCI6eyJuYW1lIjoiZXBpIn19fV0sWzMsMiwiXFxyclxcY3VwXFxycyIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImRvdHRlZCJ9LCJoZWFkIjp7Im5hbWUiOiJlcGkifX19XSxbMiwzLCJcXHByZWNlcSIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6Im5vbmUifSwiaGVhZCI6eyJuYW1lIjoiZXBpIn19fV0sWzcsNiwiIiwxLHsic2hvcnRlbiI6eyJzb3VyY2UiOjIwLCJ0YXJnZXQiOjIwfSwic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
\[\begin{tikzcd}
	a & b \\
	c & d
	\arrow["\rr\cup\rs"', dotted, two heads, from=1-1, to=1-2]
	\arrow["\preceq"', no body, two heads, from=1-2, to=1-1]
	\arrow[""{name=0, anchor=center, inner sep=0}, "\rr", dotted, two heads, from=1-2, to=2-2]
	\arrow[""{name=1, anchor=center, inner sep=0}, "\rr"', two heads, from=1-1, to=2-1]
	\arrow["\rr\cup\rs"', dotted, two heads, from=2-1, to=2-2]
	\arrow["\preceq"', no body, two heads, from=2-2, to=2-1]
	\arrow[draw=none, from=1, to=0]
\end{tikzcd}\]
\end{lemma}
\begin{proof}
  By induction on the size of the top edge of the diagram. At each step one of the four diagrams from \cref{lem:four-diagrams} will be used.

  
 \skewtodo{More to come.}
\end{proof}


\begin{lemma}\label{lem:skewrow}
  Let relation $\rr$ be monotonic in quasi order $\preceq$ and skew confluent using $\preceq$ and $\rr_\preceq^\leftarrow$;
  similarly let $\rs$ be monotonic in the same quasi order $\preceq$ and skew confluent using $\preceq$ and $\rs_\preceq^\leftarrow$.
  Then

  % https://q.uiver.app/?q=WzAsNCxbMCwwLCJhIl0sWzEsMCwiYiJdLFsxLDEsImQiXSxbMCwxLCJjIl0sWzAsMSwiXFxyclxcY3VwXFxycyIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImRvdHRlZCJ9LCJoZWFkIjp7Im5hbWUiOiJlcGkifX19XSxbMSwwLCJcXHByZWNlcSIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6Im5vbmUifSwiaGVhZCI6eyJuYW1lIjoiZXBpIn19fV0sWzEsMiwiXFxyclxcY3VwXFxycyIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImRvdHRlZCJ9LCJoZWFkIjp7Im5hbWUiOiJlcGkifX19XSxbMCwzLCJcXHJyXFxjdXBcXHJzIiwyXSxbMywyLCJcXHJyXFxjdXBcXHJzIiwyLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiZG90dGVkIn0sImhlYWQiOnsibmFtZSI6ImVwaSJ9fX1dLFsyLDMsIlxccHJlY2VxIiwyLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJlcGkifX19XSxbNyw2LCIiLDEseyJzaG9ydGVuIjp7InNvdXJjZSI6MjAsInRhcmdldCI6MjB9LCJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJub25lIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XV0=
\[\begin{tikzcd}
	a & b \\
	c & d
	\arrow["\rr\cup\rs"', dotted, two heads, from=1-1, to=1-2]
	\arrow["\preceq"', no body, two heads, from=1-2, to=1-1]
	\arrow[""{name=0, anchor=center, inner sep=0}, "\rr\cup\rs", dotted, two heads, from=1-2, to=2-2]
	\arrow[""{name=1, anchor=center, inner sep=0}, "\rr\cup\rs"', from=1-1, to=2-1]
	\arrow["\rr\cup\rs"', dotted, two heads, from=2-1, to=2-2]
	\arrow["\preceq"', no body, two heads, from=2-2, to=2-1]
	\arrow[draw=none, from=1, to=0]
\end{tikzcd}\]

\end{lemma}
\begin{proof}
By case analysis on whether left edge uses $\rr$ or $\rs$; then project that left edge into $\rr^*$ or $\rs^*$ respectively and apply \cref{lem:skewrowinduction}.

  
 \skewtodo{More to come.}
\end{proof}


\begin{lemma}\label{lem:skewgrid}
  Let relation $\rr$ be monotonic in quasi order $\preceq$ and skew confluent using $\preceq$ and $\rr_\preceq^\leftarrow$;
  similarly let $\rs$ be monotonic in the same quasi order $\preceq$ and skew confluent using $\preceq$ and $\rs_\preceq^\leftarrow$.
  Then

% https://q.uiver.app/?q=WzAsNCxbMCwwLCJhIl0sWzEsMCwiYiJdLFswLDEsImMiXSxbMSwxLCJkIl0sWzAsMiwiXFxyclxcY3VwXFxycyIsMix7InN0eWxlIjp7ImhlYWQiOnsibmFtZSI6ImVwaSJ9fX1dLFswLDEsIlxccnJcXGN1cFxccnMiLDIseyJzdHlsZSI6eyJoZWFkIjp7Im5hbWUiOiJlcGkifX19XSxbMSwzLCJcXHJyXFxjdXBcXHJzIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiZG90dGVkIn0sImhlYWQiOnsibmFtZSI6ImVwaSJ9fX1dLFsyLDMsIlxccnJcXGN1cFxccnMiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJkb3R0ZWQifSwiaGVhZCI6eyJuYW1lIjoiZXBpIn19fV0sWzMsMiwiXFxwcmVjZXEiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJub25lIn0sImhlYWQiOnsibmFtZSI6ImVwaSJ9fX1dLFsxLDAsIlxccHJlY2VxIiwyLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJlcGkifX19XV0=
\[\begin{tikzcd}
	a & b \\
	c & d
	\arrow["\rr\cup\rs"', two heads, from=1-1, to=2-1]
	\arrow["\rr\cup\rs"', two heads, from=1-1, to=1-2]
	\arrow["\rr\cup\rs", dotted, two heads, from=1-2, to=2-2]
	\arrow["\rr\cup\rs"', dotted, two heads, from=2-1, to=2-2]
	\arrow["\preceq"', no body, two heads, from=2-2, to=2-1]
	\arrow["\preceq"', no body, two heads, from=1-2, to=1-1]
\end{tikzcd}\]
 \end{lemma}
\begin{proof}
  By induction on the size of the left edge of the diagram. 

  \begin{description}
  \item[Base case] This diagram clearly holds by letting the bottom edge be the same as the top edge:

% https://q.uiver.app/?q=WzAsNCxbMCwwLCJhIl0sWzEsMCwiYiJdLFswLDEsImMiXSxbMSwxLCJkIl0sWzAsMiwiXFxlcXVpdiIsMSx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6Im5vbmUifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFswLDEsIlxccnJcXGN1cFxccnMiLDIseyJzdHlsZSI6eyJoZWFkIjp7Im5hbWUiOiJlcGkifX19XSxbMSwzLCJcXGVxdWl2IiwxLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzIsMywiXFxyclxcY3VwXFxycyIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImRvdHRlZCJ9LCJoZWFkIjp7Im5hbWUiOiJlcGkifX19XSxbMywyLCJcXHByZWNlcSIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6Im5vbmUifSwiaGVhZCI6eyJuYW1lIjoiZXBpIn19fV0sWzEsMCwiXFxwcmVjZXEiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJub25lIn0sImhlYWQiOnsibmFtZSI6ImVwaSJ9fX1dXQ==
\[\begin{tikzcd}
	a & b \\
	c & d
	\arrow["\equiv"{description}, draw=none, from=1-1, to=2-1]
	\arrow["\rr\cup\rs"', two heads, from=1-1, to=1-2]
	\arrow["\equiv"{description}, draw=none, from=1-2, to=2-2]
	\arrow["\rr\cup\rs"', dotted, two heads, from=2-1, to=2-2]
	\arrow["\preceq"', no body, two heads, from=2-2, to=2-1]
	\arrow["\preceq"', no body, two heads, from=1-2, to=1-1]
\end{tikzcd}\]

    and it implies this diagram:

% https://q.uiver.app/?q=WzAsNCxbMCwwLCJhIl0sWzEsMCwiYiJdLFswLDEsImMiXSxbMSwxLCJkIl0sWzAsMiwiKFxccnJcXGN1cFxccnMpXjAiLDJdLFswLDEsIlxccnJcXGN1cFxccnMiLDIseyJzdHlsZSI6eyJoZWFkIjp7Im5hbWUiOiJlcGkifX19XSxbMSwzLCJcXHJyXFxjdXBcXHJzIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiZG90dGVkIn0sImhlYWQiOnsibmFtZSI6ImVwaSJ9fX1dLFsyLDMsIlxccnJcXGN1cFxccnMiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJkb3R0ZWQifSwiaGVhZCI6eyJuYW1lIjoiZXBpIn19fV0sWzMsMiwiXFxwcmVjZXEiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJub25lIn0sImhlYWQiOnsibmFtZSI6ImVwaSJ9fX1dLFsxLDAsIlxccHJlY2VxIiwyLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJlcGkifX19XV0=
\[\begin{tikzcd}
	a & b \\
	c & d
	\arrow["{(\rr\cup\rs)^0}"', from=1-1, to=2-1]
	\arrow["\rr\cup\rs"', two heads, from=1-1, to=1-2]
	\arrow["\rr\cup\rs", dotted, two heads, from=1-2, to=2-2]
	\arrow["\rr\cup\rs"', dotted, two heads, from=2-1, to=2-2]
	\arrow["\preceq"', no body, two heads, from=2-2, to=2-1]
	\arrow["\preceq"', no body, two heads, from=1-2, to=1-1]
\end{tikzcd}\]
    
	\item[Inductive case] Assume the diagram holds for left edges of all sizes up to $n$. Then this diagram:

% https://q.uiver.app/?q=WzAsNCxbMCwwLCJhIl0sWzEsMCwiYiJdLFswLDEsImMiXSxbMSwxLCJkIl0sWzAsMiwiKFxccnJcXGN1cFxccnMpXntuKzF9IiwyXSxbMCwxLCJcXHJyXFxjdXBcXHJzIiwyLHsic3R5bGUiOnsiaGVhZCI6eyJuYW1lIjoiZXBpIn19fV0sWzEsMywiXFxyclxcY3VwXFxycyIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImRvdHRlZCJ9LCJoZWFkIjp7Im5hbWUiOiJlcGkifX19XSxbMiwzLCJcXHJyXFxjdXBcXHJzIiwyLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiZG90dGVkIn0sImhlYWQiOnsibmFtZSI6ImVwaSJ9fX1dLFszLDIsIlxccHJlY2VxIiwyLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJlcGkifX19XSxbMSwwLCJcXHByZWNlcSIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6Im5vbmUifSwiaGVhZCI6eyJuYW1lIjoiZXBpIn19fV1d
\[\begin{tikzcd}
	a & b \\
	c & d
	\arrow["{(\rr\cup\rs)^{n+1}}"', from=1-1, to=2-1]
	\arrow["\rr\cup\rs"', two heads, from=1-1, to=1-2]
	\arrow["\rr\cup\rs", dotted, two heads, from=1-2, to=2-2]
	\arrow["\rr\cup\rs"', dotted, two heads, from=2-1, to=2-2]
	\arrow["\preceq"', no body, two heads, from=2-2, to=2-1]
	\arrow["\preceq"', no body, two heads, from=1-2, to=1-1]
\end{tikzcd}\]

    \noindent
    follows from this diagram:

% https://q.uiver.app/?q=WzAsNixbMCwwLCJhIl0sWzEsMCwiYiJdLFswLDEsImMiXSxbMSwxLCJkIl0sWzAsMiwiZSJdLFsxLDIsImYiXSxbMCwyLCIoXFxyclxcY3VwXFxycylebiIsMl0sWzAsMSwiXFxyclxcY3VwXFxycyIsMix7InN0eWxlIjp7ImhlYWQiOnsibmFtZSI6ImVwaSJ9fX1dLFsxLDMsIlxccnJcXGN1cFxccnMiLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJkb3R0ZWQifSwiaGVhZCI6eyJuYW1lIjoiZXBpIn19fV0sWzIsMywiXFxyclxcY3VwXFxycyIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImRvdHRlZCJ9LCJoZWFkIjp7Im5hbWUiOiJlcGkifX19XSxbMywyLCJcXHByZWNlcSIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6Im5vbmUifSwiaGVhZCI6eyJuYW1lIjoiZXBpIn19fV0sWzEsMCwiXFxwcmVjZXEiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJub25lIn0sImhlYWQiOnsibmFtZSI6ImVwaSJ9fX1dLFsyLDQsIlxccnJcXGN1cFxccnMiLDJdLFszLDUsIlxccnJcXGN1cFxccnMiLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJkb3R0ZWQifSwiaGVhZCI6eyJuYW1lIjoiZXBpIn19fV0sWzQsNSwiXFxyclxcY3VwXFxycyIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImRvdHRlZCJ9LCJoZWFkIjp7Im5hbWUiOiJlcGkifX19XSxbNSw0LCJcXHByZWNlcSIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6Im5vbmUifSwiaGVhZCI6eyJuYW1lIjoiZXBpIn19fV0sWzEyLDEzLCJcXGxhYmVsY3JlZntsZW06c2tld3Jvd30iLDEseyJzaG9ydGVuIjp7InNvdXJjZSI6MjAsInRhcmdldCI6MjB9LCJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJub25lIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XV0=
\[\begin{tikzcd}
	a & b \\
	c & d \\
	e & f
	\arrow["{(\rr\cup\rs)^n}"', from=1-1, to=2-1]
	\arrow["\rr\cup\rs"', two heads, from=1-1, to=1-2]
	\arrow["\rr\cup\rs", dotted, two heads, from=1-2, to=2-2]
	\arrow["\rr\cup\rs"', dotted, two heads, from=2-1, to=2-2]
	\arrow["\preceq"', no body, two heads, from=2-2, to=2-1]
	\arrow["\preceq"', no body, two heads, from=1-2, to=1-1]
	\arrow[""{name=0, anchor=center, inner sep=0}, "\rr\cup\rs"', from=2-1, to=3-1]
	\arrow[""{name=1, anchor=center, inner sep=0}, "\rr\cup\rs", dotted, two heads, from=2-2, to=3-2]
	\arrow["\rr\cup\rs"', dotted, two heads, from=3-1, to=3-2]
	\arrow["\preceq"', no body, two heads, from=3-2, to=3-1]
	\arrow["{\labelcref{lem:skewrow}}"{description}, draw=none, from=0, to=1]
\end{tikzcd}\]

\noindent
  where the top half is the inductive hypothesis and the bottom half follows from \cref{lem:skewrow}.
\end{description}
\end{proof}


\begin{lemma}\label {lem:skew-union}
  Let relation $\rr$ be monotonic in quasi order $\preceq$ and skew confluent using $\preceq$ and $\rr_\preceq^\leftarrow$;
  similarly let $\rs$ be monotonic in the same quasi order $\preceq$ and skew confluent using $\preceq$ and $\rs_\preceq^\leftarrow$.
  If $\rr$ commutes with $\rs$, then $\rt=\rr\cup\rs$ is monotonic in $\preceq$ and skew confluent using $\preceq$ and $\rt_\preceq^\leftarrow$.
\end{lemma}
\begin{proof}

\end{proof}



\subsection{Proof that \texorpdfstring{\versecalc{}}{VC} Is Skew Confluent}


\skewtodo{This is just a brief proof sketch.}

First prove that the modified $\runi$ is skew confluent. (In doing so we will define $\runi_{\leqomegavc}^{\leftarrow}$.)

Then use existing proofs to demonstrate that $\rapp \cup \rcon \cup \rgc \cup \rspc$ is confluent.
Because they are also monotonic, they are therefore skew confluent, and $(\rapp \cup \rcon \cup \rgc \cup \rspc) _{\leqomegavc}^{\leftarrow}$ is trivial.

Prove that $\rapp \cup \rcon \cup \rgc \cup \rspc$ commutes with $\runi_{\leqomegavc}^{\leftarrow}$.

Then apply \cref {lem:skew-union} to show that $\runi \cup (\rapp \cup \rcon \cup \rgc \cup \rspc)$ is skew confluent.
