%include lhs2TeX.fmt
%include lhs2TeX.sty
%include polycode.fmt
%include ../calculus.fmt
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Semantics} \label{sec:semantics}

Invariants that are maintained by the contexts and rules

%
\mypara{Stratification of Values}
%
We layer values into \emph{scalars} (integers, variables)
and \emph{heap values} (tuples and $\lambda$-terms).
%
Only scalars are allowed inside tuples (\ie no tuples or $\lambda$-terms in tuples).
%
Note this is not a limitation: you can ANF to name the
inner elements (tuples or $\lambda$-terms).
%
Use \rulename{utup} to first destruct or ``unpack'' the tuple and
then you can access the values.
%
We do this stratification to prevent the following undesirable behaviors.

\mypara{Non-Confluence: JB POPL 5.3}
Substituting $\lambda$ under tuples (and tuples under $\lambda$)
breaks confluence as shown below. (The same happens just with mutually
recursive $\lambda$-under-$\lambda$~\cite{Ariola2002}.)

\begin{verbatim}
ex x,y. x = <1,y>; y = \z.x; x
-->> ex x. x = <1, \z.x>; x         (1)

ex x,y. x = <1,y>; y = \z.x; x
-->> ex x. y = \z.<1, y>; <1, y>    (2)
\end{verbatim}

\mypara{Divergence: JB on Slack}
%
Even just substituting tuple under tuple can lead to non-termination
which breaks the confluence via Newman's lemma route for the
unification-rules fragment. Note we need termination as we can only
directly prove \emph{local} confluence for the unification reductions
as \cref{lem:join-h-h} requires multiple steps to reach a peak.

\begin{verbatim}
    ex x, y. x = <1,y>; y = <2,x>; x
--> ex x, y. x = <1,y>; y = <2,x>; <1,y>
--> ex x, y. x = <1,y>; y = <2,x>; <1,<2,x>>
--> ex x, y. x = <1,y>; y = <2,x>; <1,<2,<1,y>>>
--> ...
\end{verbatim}

\mypara{Well-formedness}
%
The WF predicate checks each variable is defined \emph{once};
you \emph{can} have cyclic dependencies and those do not fail.
%
(Lemma [Self-Tuple] from the notes is not needed, it was only
required when substituting a tuple under a tuple.)


\mypara{Example: Lazy List}

\begin{verbatim}
    x = <1,x>; <y1, <y2, <y3, z>>> = x; y1 + y2 + y3
--> x = <1,x>; <y1, <y2, <y3, z>>> = <1, x>; y1 + y2 + y3
--> x = <1,x>; y1 = 1; <y2, <y3, z>> = x; y1 + y2 + y3
--> x = <1,x>; y1 = 1; <y2, <y3, z>> = <1, x>; y1 + y2 + y3
--> x = <1,x>; y1 = 1; y2 = 1; <y3, z> = x; y1 + y2 + y3
--> x = <1,x>; y1 = 1; y2 = 1; <y3, z> = <1, x>; y1 + y2 + y3
--> x = <1,x>; y1 = 1; y2 = 1; y3 = 1; z = x; y1 + y2 + y3
--> ...; 1 + 1 + 1
--> 3
\end{verbatim}


\figureRulesI

% included in figureRulesI
% \figureRulesII

\figureContexts

\figureWellformed