%include lhs2TeX.fmt
%include lhs2TeX.sty
%include polycode.fmt
%include ../calculus.fmt
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{POPL-style syntax and semantics} \label{sec:popl-style}

\figureSyntaxPOPL
\figureRulesPOPL
\figureRulesPOPLconj

For the POPL-style rules:
\begin{itemize}
\item \Cref{fig:popl-syntax} (c.f. \Cref{fig:syntax}) gives the POPL-style syntax.  We have no ``|c|'' form (except in the syntax of results |r|, an dit is optional even there).  Sequencing, unification, function application all have expressions on both sides.

\item \Cref{fig:popl-structural-equiv} (c.f. \Cref{fig:structural-equiv}) gives the structural rules.
  The old \rulename{swap-e} has become \rulename{conj-swap-eq} in the
  non-structural (directed) rules.

\item There are no desugaring rules (c.f. \Cref{fig:desugar}).

\item \Cref{fig:popl-rulesI} are the main rewrite rule. Just the same as the Fresh-style ones (\Cref{fig:rulesI}) except
  for fixing \rulename{app-beta}.

\item \Cref{fig:popl-rulesN} and \ref{fig:popl-rulesN-alt} are two alternative sets of ``conjunction'' rules, which
  float existentials and equalities out of expressions:
  \begin{itemize}
  \item \Cref{fig:popl-rulesN} tackles it head on.
  \item \Cref{fig:popl-rulesN-alt} uses \rulename{conj-flat-eq} and \rulename{conj-falt-fun} to normalise equalities and function applications to have a value on the LHS.  That reduces the number of other rules, and makes them simpler too.  In effect,
    a rule like \rulename{conj-arg-cst} has to do flattening on the fly.
  \end{itemize}
  Lennart and I like the second set best, and it's nearest to your existing syntax.  Indeed, if you want, the two flattening rules can be applied ``at the start'', yielding a language in which equality always has form |(v = e)|, and function application always has form |(v e)|.  Once the expression is in this form, no rewrite destroys the form.
\end{itemize}
The six existential-floating rules could be replaced with one if we had a suitable context
$$
X ::= |hole| \vb |X == e| \vb |e == X| \vb |X; e| \vb |e ; X| \vb |apply1N X e| \vb |apply1N e X|
$$
The six existential-floating rules could be replaced with one if we had a suitable context
$$
X ::= |hole| \vb |X == e| \vb |e == X| \vb |X; e| \vb |e ; X| \vb |apply1N X e| \vb |apply1N e X|
$$
The six existential-floating rules could be replaced with one if we had a suitable context
$$
X ::= |hole| \vb |X == e| \vb |e == X| \vb |X; e| \vb |e ; X| \vb |apply1N X e| \vb |apply1N e X|
$$
The six existential-floating rules could be replaced with one if we had a suitable context
$$
X ::= |hole| \vb |X == e| \vb |e == X| \vb |X; e| \vb |e ; X| \vb |apply1N X e| \vb |apply1N e X|
$$
