%include lhs2TeX.fmt
%include lhs2TeX.sty
%include polycode.fmt
%include ../calculus.fmt
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Confluence for \calcname} \label{sec:confluence}

\begin{definition}[Reductions] \label{def:reductions}
Let $\rr$ be the reduction relation defined as $\runi \cup \rapp \cup \rcon \cup \rgc \cup \rfail$,
each of which is defined as the reduction relation induced by the respective rules in \cref{fig:rulesI}
%
\begin{itemize}
\item $\runi$ : Unification,
\item $\rapp$ : Application,
\item $\rcon$ : Conjunction,
\item $\rgc$ : Elimination,
\item $\rfail$ : Fail Propagation.
\item $\rspc$ : Speculation.
\end{itemize}
%
\end{definition}

\begin{figure}
\begin{tabular}{rrrrrrrr}
\toprule
					& 	  	& \runi 		 & \rapp 		   & \rcon 			 & \rfail 		   & \rgc 			 & \rspc \\
\midrule
Unification 		& \runi & \ref{lem:u-c}  & \ref{lem:u-a-c} & \ref{lem:u-c-c} & \ref{lem:u-f-c} & \ref{lem:u-e-c} & \ref{lem:u-s-c} \\
$\beta$-Application & \rapp &      			 & \ref{lem:a-c}   & \ref{lem:a-c-c} & \ref{lem:a-f-c} & \ref{lem:a-e-c} & \ref{lem:a-s-c} \\
Conjunction 		& \rcon &      			 &                 & \ref{lem:c-c}   & \ref{lem:c-f-c} & \ref{lem:c-e-c} & \ref{lem:c-s-c} \\
Fail Propagation	& \rfail &      	         &                 &                 & \ref{lem:f-c}   & \ref{lem:f-e-c} & \ref{lem:f-s-c} \\
Elimination 		& \rgc &      		     &                 &                 &                 & \ref{lem:e-c}   & \ref{lem:e-s-c} \\
Speculation 		& \rspc &				 & 			       & 	             & 	   			   &           	     & \ref{lem:s-c}   \\
\bottomrule
\end{tabular}
\caption{Summary of the confluence and commutativity of the reductions in \cref{def:reductions}.
The lemmas on the diagonal (resp. non-diagonal) entries establish confluence (resp. commutativity)
for the respective relation (resp. pairs of relations).}
\label{fig:confluence-summary}
\end{figure}

\begin{theorem}[Confluence] \label{thm:confluence}
The reduction relation $\rr$ is confluent.
\end{theorem}

\begin{proof}
First, we partition $\rr$ into the relations $\runi$, $\rapp$, $\rcon$, $\rfail$,
$\rgc$ and $\rspc$.
%
Next, we show that each of these relations is confluent and pairwise
commutative.
%
Finally, we use \cref{lem:comm-n} to prove their union $\rr$ is confluent.
\end{proof}

\subsection{Unification}

\begin{lemma}[$\runi$-Confluent]\label{lem:u-c}
    $\runi$ is confluent.
\end{lemma}

\begin{proof}
%
\textbf{Step 1:} We define a relation $\rord$ that is parameterized
by a rank function $\rnk$ that maps each identifier name to a rank,
and restricts \rulename{deref-s} to only allow substituting |x| for
|y| if |x| is \emph{ranked below} |y|.
%
\textbf{Step 2:} Next, via \verb|TODO: [Deref-Com-1]| and \verb|TODO: [Deref-Com-2]|
(which depends on \verb|TODO: [Eq-Join]|) we show that $\rord$
is \emph{locally} confluent.
%
\textbf{Step 3:} Next we show that $\rord$ is terminating \verb|TODO: [uni-term]|
and hence by Newman's lemma \cite{Barendregt} $\rord$ is confluent.
%
\textbf{Step 4:} Finally, define $\runi'$ as $\exists \rnk. \rord^{*}$.
By \cref{lem:closure}, $\runi'$, and hence, its reflexive transitive
closure, $\rr$ are confluent.
\end{proof}

\subsection{Unification and Application}

\subsubsection{Application}

\begin{lemma}[$\rapp$-Confluent]\label{lem:a-c}
    $\rapp$ is confluent.
\end{lemma}

\begin{proof}
Lemma \verb|TODO: [Beta-Comm]| shows that $\rapp$ satisfies
the diamond property and hence is confluent by \cref{lem:diamond}.
\end{proof}

\begin{lemma}[$\runi$-$\rapp$-Commute]\label{lem:u-a-c}
    $\runi$ and $\rapp$ strongly commute.
\end{lemma}

\begin{proof}
Lemma \verb|TODO: [Deref-Beta]| shows that $\runi$ and $\rapp$ strongly commute.
\end{proof}

\begin{lemma}[$\rspc$-Confluent]\label{lem:s-c}
    $\rspc$ is confluent.
\end{lemma}

\subsection{Pairwise Commutativity}

\begin{lemma}[$\rspc$-Conf] \label{lem:s-c}
	$\rspc$ is confluent.
\end{lemma}

\begin{proof}
\ranjit{TODO:Fresh-Verse.pdf}
\end{proof}

\begin{lemma}[$\runi-\rcon$-Comm] \label{lem:u-c-c}
	$\runi$ and $\rcon$ commute.
\end{lemma}

\begin{proof}
\ranjit{TODO:Fresh-Verse.pdf}
\end{proof}

\begin{lemma}[$\runi-\rfail$-Comm] \label{lem:u-f-c}
	$\runi$ and $\rfail$ commute.
\end{lemma}

\begin{proof}
\ranjit{TODO:Fresh-Verse.pdf}
\end{proof}

\begin{lemma}[$\runi-\rgc$-Comm] \label{lem:u-e-c}
	$\runi$ and $\rgc$ commute.
\end{lemma}

\begin{proof}
\ranjit{TODO:Fresh-Verse.pdf}
\end{proof}

\begin{lemma}[$\runi-\rspc$-Comm] \label{lem:u-s-c}
	$\runi$ and $\rspc$ commute.
\end{lemma}

\begin{proof}
\ranjit{TODO:Fresh-Verse.pdf}
\end{proof}