%include lhs2TeX.fmt
%include lhs2TeX.sty
%include polycode.fmt
%include ../calculus.fmt
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newpage

\section{Metatheory} \label{sec:metatheory}

The rules of our rewrite semantics can be applied anywhere,
in any order, and they give meaning to programs without
committing to a particular evaluation strategy.
%
But then it had better be the case that no matter how the
rules are applied, one always obtains the same result!
%
If the rewrite rules were terminating, then we would expect
that all possible execution paths lead to the same normal form.
%
However, since \calcname has recursion, we instead show that
the rules are \emph{confluent}.

\subsection{Confluence}

\mypara{Reductions and Confluence}
%
A \emph{reduction} \rr is a binary relation on a set of terms \univ.
%
We write $\rr^{k}$ for the $k$-step closure of $\rr$ and $\rr^{*}$
for the reflexive and transitive closure of $\rr$, \ie $\rr^{*} \equiv \cup_{0 \leq k} \rr^k$.
%
We write $\rstep{\rr}{e}{e'}$ ($a$ \emph{steps to} $b$) if $(e, e') \in \rr$ and
$\rsteps{\rr}{e}{e'}$ ($a$ \emph{reduces to} $b$) if $(e, e') \in \rr^{*}$.
%
A reduction $\rr$ is \emph{confluent} if whenever $\rsteps{\rr}{e}{e_1}$ and $\rsteps{\rr}{e}{e_2}$,
there exists an $e'$ such that $\rsteps{\rr}{e_1}{e'}$ and $\rsteps{\rr}{e_2}{e'}$.
%
Confluence gives us the assurance that we will not get
different results when choosing different rules, or get
stuck with some rules and not with others.

\mypara{Normal Forms}
%
A term $e$ is an $\rr$-\emph{Normal Form} if there does
not exist any $e'$ such that $\rstep{\rr}{e}{e'}$.
Confluence implies that ultimately, rewriting terminates
with at most one unique normal form, regardless of the
evaluation strategy \cite{Barendregt}.

\begin{lemma}[Unicity] \label{lem:nf}
If $\rr$ is confluent then every term reduces to at most one normal form.
\end{lemma}

Our main result is that the reduction relation $\rr$ for \calcname,
defined as the union of the reduction relations
$\runi$ (Unification),
$\rapp$ ($\beta$-Application),
$\rcon$ (Conjunction),
$\rgc$  (Garbage Collection),
$\rfail$ (Fail Propagation) and
$\rspc$ (Speculation)
in \cref{FIXME} % \cref{fig:rulesI}
is confluent.

\begin{theorem}[Confluence] \label{thm:confluence}
The reduction relation $\rr \equiv \runi \cup \rapp \cup \rcon \cup \rgc \cup \rfail$ is confluent.
\end{theorem}

A reduction $R$ \emph{commutes} with $S$ if for all terms
$e, e_1, e_2$ such that $\rsteps{R}{e}{e_1}$ and $\rsteps{S}{e}{e_2}$
there exists $e'$ such that $\rsteps{S}{e_1}{e'}$ and $\rsteps{R}{e_2}{e'}$.
%
We establish confluence by proving each sub-relation is confluent,
and that they pairwise commute \cite{Huet}.

\begin{lemma}[Commutativity] \label{lem:comm}
If $R$ and $S$ are confluent and commute, then $R \cup S$ is confluent.
\end{lemma}

\subsection{Design}

\calcname deliberately layers values into \emph{scalars} (integers, variables)
and \emph{heap values} (tuples and $\lambda$-terms) and then only
allows scalars (no tuples or $\lambda$-terms) in tuples to ensure 
confluence.
%
Further, the \rulename{deref-h} rule prevents substituting heap values
under $\lambda$ and tuples, and prevents ``copying'' heap values
in constraints as all of these lead to non-confluence.

\mypara{Non-Confluence due to nested $\lambda$}
%
Suppose we eliminated \rulename{deref-h} and instead used a single
rule like \rulename{deref-s} for \emph{all} values.
%
Then we could substitute $\lambda$ under tuples (and tuples under $\lambda$)
which would violate confluence as shown by the following example.
%
\ranjit{TODO:flesh out steps}
%
\begin{verbatim}
ex x,y. x = <1,y>; y = \z.x; x
-->> ex x. x = <1, \z.x>; x         (1)

ex x,y. x = <1,y>; y = \z.x; x
-->> ex x. y = \z.<1, y>; <1, y>    (2)
\end{verbatim}
%
(A similar phenomenon arises just with inlining mutually recursive
$\lambda$-terms~\cite{Ariola2002}.)

\mypara{Non-confluence due to nested tuples}
%
It is possible to prove confluence even for potentially diverging
(non-terminating) reductions, most famously, for the $\lambda$-calculus.
%
One general strategy is to find a \emph{sub-relation} that satisfies
three properties.
%
First, it is \emph{locally} confluent, meaning if $e$ (single-)steps to $e_1$
and $e_2$ then $e_1$ and $e_2$ can be joined at some $e'$.
%
Second, it is terminating. Newman's Lemma \cite{Barendregt} then implies 
the relation is confluent (and hence, so is its reflexive transitive closure.)
%
Third, that the \emph{closure} of the sub-relation is the same as the full reduction
relation, which then implies that the full reduction relation is also confluent.

We follow this route to establish the confluence of the $\runi$
reductions. The other relations can be proved confluence by
establishing a stronger \emph{diamond} property, namely, two
different reduction steps can be joined at a common
term in a fixed number of steps.
%
But for unification, we require multiple steps to reach
a common join if we have a term $x = |arr(s_1,...s_n)|; x = |arr(t_1,...,t_n)|; E[x]$,
which can be reduced to $e_1$ and $e_2$ by respectively substituting
the first or second tuple in \rulename{deref-h}, and which must then
be joined by individually unifying the elements of $E[|arr(s_1,...,s_n)|]$
and $E[|arr(t_1,...,t_n)|]$.

As we need potentially unbounded steps to join, we require using Newman's
lemma, and hence, that $\runi$ terminates. However, just substituting tuples
under tuples leads to divergence which blocks this proof strategy.
For example, consider the term
%
\begin{verbatim}
    ex x, y. x = <1,y>; y = <2,x>; x
--> ex x, y. x = <1,y>; y = <2,x>; <1,y>
--> ex x, y. x = <1,y>; y = <2,x>; <1,<2,x>>
--> ex x, y. x = <1,y>; y = <2,x>; <1,<2,<1,y>>>
--> ...
\end{verbatim}
%
We avoid this problem by preventing \rulename{deref-h} from substituting
tuples under tuples. Instead these are left as ``heap-constraints'' which
can be used whenever we actually need to access the contents of the tuple
via pattern matching (or eliminated via the garbage collection rules.)

\mypara{Non-confluence due to Heap Copying}
%
Finally, notice that we disallow ``copying'' heap values in constraints.
%
That is, \rulename{deref-h} does not allow substituting heap values (e.g. $x = h$)
into variable-variable equalities (e.g. $y = x$).
%
This is because such copying could prevent confluence
as some reductions might fail while others succeed.
%
For example consider the term $|x == (lam y e); x == y; y == x|$.
%
\begin{verbatim}
      (ex x,y. x = \z. z; y = x; x = y; x) 3
-->   ex f. f = (ex x,y. x = \z. z; y = x; x = y; x); f 3
-->>  ex f,x,y. x = \z. z; y = x; f = x; f 3
-->>  ex f,x,y. x = \z. z; y = x; f = x; x 3
-->>  ex f,x,y. x = \z. z; y = x; f = x; (\z.z) 3
-->>  3
\end{verbatim}
%
However, if we \emph{allow} copying heap values, then the above term could also fail
%
\begin{verbatim}
     (ex x,y. x = \z. z; y = x; x = y; x) 3
-->  (ex x,y. x = \z. z; y = \z.z; x = y; x) 3
-->> (ex x,y. x = \z. z; y = \z.z; \z.z = \z.z; x) 3
-->  FAIL
\end{verbatim}
%
An alternate ``big-step'' strategy that substituted
identifiers with values \emph{everywhere} they were
in a single swoop would avoid the above problem, but
we were unable to prove confluence using such a ``big-step''
rule.

\mypara{Discussion}
%
Value stratification is not a limitation: in essence we ANF to name the
inner elements (tuples or $\lambda$-terms), and then use the
unification rule \rulename{utup} to destruct or ``unpack'' the
tuple and to access its contents.
%
However, not substituting under lambda is a tiresome restriction: e.g.,
compilers will certainly want to do so. \ranjit{Simon -- add more discussion here? Skew Confluence?}
