%include lhs2TeX.fmt
%include lhs2TeX.sty
%include polycode.fmt
%include ../calculus.fmt

% ------------------- Local notational macros -----------------
\newcommand{\vtt}[1]{\text{\tt #1}}

%%\newcommand{\clamsymx}{\Rightarrow}
%%\newcommand{\clamsym}{$\clamsymx$}
\newcommand{\clamx}[2]{lam (#1) in #2}
\newcommand{\clam}[2]{\vtt{\clamx{#1}{#2}}}

\newcommand{\tuple}[1]{\langle #1 \rangle}
\newcommand{\Value}{W}
\newcommand{\Env}{Env}
\newcommand{\equa}[1]{\begin{equation*}#1\end{equation*}}
\newcommand{\mcompr}[2]{[ #1 \vb #2 ]}
\newcommand{\mbind}[2]{ #1 \leftarrow #2 }
\newcommand{\malt}{\diamond}

\newcommand{\setform}[1]{\left\{ #1 \right\}}
\newcommand{\scompr}[2]{\{ #1 \vb #2 \}}
\newcommand{\sbind}[2]{ #1 \in #2 }

\newcommand{\semfun}[1]{\mathcal{#1}}
\newcommand{\semapp}[3]{\semfun{#1}\llbracket #2 \rrbracket\, #3}
\newcommand{\envapp}[2]{\semfun{R}\llbracket #1 \rrbracket #2}
\newcommand{\liftrel}[1]{\hat{#1}}
\newcommand{\cons}{{:}}           % Cons for value sequences, V*
\newcommand{\eff}{\mathit{eff}}   % Non-terminal for effects

% extend envt
\newcommand{\extendrho}[2]{#1[#2]}

% The collection type
\newcommand{\mtype}{\Value^{*}}
\newcommand{\mtyped}{\Value^{?}}
\newcommand{\mempty}{\mathit{empty}}
\newcommand{\munit}{\mathit{unit}}
\newcommand{\munits}[1]{\munit(#1)}
\newcommand{\mintersect}{\Cap}
\newcommand{\munion}{\Cup}
\newcommand{\msemi}{\fatsemi}
\newcommand{\mone}{\mathit{one}}
\newcommand{\mall}{\mathit{all}}
\newcommand{\mtag}{\mathit{tag}}
\newcommand{\mtage}[1]{([],#1)}
\newcommand{\mlift}[1]{#1_{\bot}}
\newcommand{\mplus}{\oplus}
\newcommand{\powerset}[1]{{\mathcal P}(#1)}
\newcommand{\wrong}{\mathrm{WRONG}}
\newcommand{\apply}{\mathit{apply}}
\newcommand{\sconcat}{\bowtie}
\newcommand{\shead}{\mathit{head}}
\newcommand{\stuple}{\mathit{tuple}}
\newcommand{\scons}{:}

\section{A denotational semantics for \versecalc} \label{sec:den-sem}


\begin{figure}
$
\begin{array}{l}
\text{\bf Domains} \\
\begin{array}{rcll}
\Value & = & \mathbb{Z} + \tuple{\Value} + (\Value \rightarrow \mtype) \\
\tuple{\Value} & = & \text{a finite tuple of values $\Value$} \\
\Env & = & Ident \rightarrow \Value \\
\end{array}
\\[7mm]
\text{\bf Semantics of expressions and values} \\

  \begin{array}{rcll}
  \semapp{E}{|e|}{} & : & \Env \rightarrow \mtype \\
  \semapp{E}{|v|}{\rho} & = & \munit(\semapp{V}{v}{\rho}) \\
  \semapp{E}{|fail|}{\rho} & = & \mempty \\
  \semapp{E}{|e_1 choice e_2|}{\rho} & = & \semapp{E}{|e_1|}{\rho} \; \munion     \; \semapp{E}{|e_2|}{\rho} \\
  \semapp{E}{|e_1 = e_2|}{\rho}      & = & \semapp{E}{|e_1|}{\rho} \; \mintersect \; \semapp{E}{|e_2|}{\rho} \\
  \semapp{E}{|e_1 ; e_2|}{\rho}      & = & \semapp{E}{|e_1|}{\rho} \; \msemi      \; \semapp{E}{|e_2|}{\rho} \\
  \semapp{E}{|apply1N v_1 v_2|}{\rho} & = & \apply(\semapp{V}{|v_1|}{\rho}, \; \semapp{V}{|v_2|}{\rho}) \\
  \semapp{E}{|def x e|}{\rho} & = & \bigcup_{w \in \Value}\, \semapp{E}{e}{(\extendrho{\rho}{x \mapsto w})} \\
  \semapp{E}{|one e|}{\rho} & = & \mone (\semapp{E}{e}{\rho}) \\
  \semapp{E}{|all e|}{\rho} & = & \munit(\mall (\semapp{E}{e}{\rho})) \\
% Commenting out hastype for now
%   \semapp{E}{|hastype e_1 e_2|}{\rho} & = & \begin{array}[t]{ll}
%     r & \text{if $\semapp{E}{|def (x t) (x=e_1; t=e_2; t x|}{\rho} = r$} \\
%       \wrong & \text{otherwise} \\
%      \multicolumn{2}{l}{\text{where $r = \semapp{E}{e_1}{\rho}$}}
%         \end{array} \\
\\
  \semapp{V}{|v|}{} & : & \Env \rightarrow \Value \\
  \semapp{V}{|x|}{\rho} & = & \rho(x) \\
  \semapp{V}{|k|}{\rho} & = & k \\
  \semapp{V}{op}\rho & = & \semapp{O}{op} \\
  \semapp{V}{|lam x e|}{\rho} & = & \lambda w . \semapp{E}{e} (\extendrho{\rho}{x \mapsto w}) \\
  \semapp{V}{|tup (v_1,dots,v_n)|}{\rho} & = & \tuple{\semapp{V}{v_1}{\rho}, \cdots, \semapp{V}{v_n}{\rho}} \\
\\
  \semapp{O}{|op|}{} & : & \Value \\
  \semapp{O}{|add|} & = &  \lambda w.\, \text{\bf if}\, (w = |tup(k_1,k_2)|)
      \,\text{\bf then}\, \munit(k_1+k_2)
      \,\text{\bf else}\, \wrong \\
  \semapp{O}{|gt|}  & = &  \lambda w.\, \text{\bf if}\, (w = |tup(k_1,k_2)| \wedge k_1 > k_2)
      \,\text{\bf then}\, \munit(k_1)
      \,\text{\bf else}\, \mempty \\

  \semapp{O}{|int|}  & = &  \lambda w.\, \text{\bf if}\, (w = k)
      \,\text{\bf then}\, \munit(k)
      \,\text{\bf else}\, \mempty
  \end{array} \\ \\
  
\begin{array}{rcll}
apply & : & (\Value \times \Value) \rightarrow \mtype \\
apply( k, w) & = & \wrong & k \in \mathbb{Z} \\
apply( \tuple{v_0, \ldots, v_n}, k) & = & \munit(v_k) & 0 \leq k \leq n \\
  & = &  \mempty & \text{otherwise} \\
apply( f, w ) & = & f(w) & f \in \Value \rightarrow \mtype \\
\end{array}
\end{array}
$
\caption{Expression semantics}
\label{fig:den-sem1}
\end{figure}

\begin{figure}
$$
\begin{array}{l}
\text{\bf Domains} \\
\mtype  =  \mlift{(\wrong + \powerset{\Value})} \\[2mm]
\text{\bf Operations} \\
\begin{array}{rrcll}
\text{Empty}        & \mempty & : &  \mtype \\
                    & \mempty & = & \{ \, \} \\[1mm]
\text{Unit}    & \munit & : & \Value \rightarrow \mtype \\
               & \munit(w) & = & \{ w  \} \\[1mm]
\text{Union}   & \munion & : & \mtype \rightarrow \mtype \rightarrow \mtype \\
               &  s_1 \munion s_2 & = & s_1 \cup s_2 \\[1mm]
\text{Intersection} & \mintersect & : & \mtype \rightarrow \mtype \rightarrow \mtype \\
                    &  s_1 \mintersect s_2 & = & s_1 \cap s_2 \\[1mm]
\text{Sequencing}   & \msemi & : & \mtype \rightarrow \mtype \rightarrow \mtype \\
                    &  s_1 \msemi s_2 & = & s_2 & \text{ if $s_1$ is non-empty} \\
            &          & = & \{\,\} & \text{otherwise} \\[1mm]
\text{One}          &   \mone  & : & \mtype \rightarrow \mtype & \text{The result is either empty or a singleton}\\
  & \mone(s) & = & ??? \\[1mm]
\text{All}          &   \mall  & : & \mtype \rightarrow \tuple{\Value} \\
  & \mall(s) & = & ??? \\
\end{array} \\ \\
\text{All operations over $\mtype$ implicitly propagate $\bot$ and $\wrong$.  E.g.}\\
  \begin{array}{rcll}
    s_1 \munion s_2 & = & \bot & \text{if $s_1=\bot$ or $s_2=\bot$} \\
                    & = & \wrong & \text{if $(s_1=\wrong$ and $s_2 \neq \bot)$ or
                    $(s_2=\wrong$ and $s_1 \neq \bot)$} \\
                    & = & s_1 \cup s_2 & \text{otherwise}
  \end{array}
\end{array}
$$
\caption{Set semantics for $\mtype$}
\label{fig:set-sem}
\end{figure}

It is highly desirable to have a denotational semantics for \versecalc{}.
A denotational semantics says directly what an expression \emph{means} rather than
how it \emph{behaves}, and that meaning can be very perspicuous.  Equipped with a
denotational semantics we can, for example, prove that the left hand side and
right hand side of each rewrite rule have the same denotation; that is, the rewrites
are meaning-preserving.

But a denotational semantics for a functional logic language is tricky.  Typically one
writes a denotation function something like
$$
\semapp{E}{|e|}{}  :  \Env \rightarrow \Value
$$
where $\Env = Ident \rightarrow \Value$.  So $\semfun{E}$ takes an expession
|e| and an environment $\rho : \Env$ and returns the value, or denotation, of the
expresssion.  The environment binds each free variable of |e| to its value.
But what is the semantics of |def x e|?   We need to extend $\rho$ with a binding
for |x|, but what is |x| bound to?  In a functional logic language |x| is given
its value by various equalities scattered throughout |e|.

This section sketches our approach to this challenge.  It is not finished work, and
does not count as a contribution of our paper.  We offer it because we have found it
an illuminating alternative way to understand \versecalc{}, one that complements
the rewrite rules that are the substance of the paper.

\subsection{A first attempt at a denotational semantics}  \label{sec:collection}  \label{sec:unification-and-equality}

Our denotational semantics for \versecalc{} is given in \cref{fig:den-sem1}.
\begin{itemize}
\item We have one semantic function (here $\semfun{E}$ and $\semfun{V}$) for each syntactic
  non terminal (here $e$ and $v$ respectively.)
\item Each function has one equation for each form of the construct.
\item Both functions take an environment $\rho$ that maps in-scope identifiers to
  a \emph{single} value; see the definition $\Env = Ident \rightarrow \Value$.
\item The value function $\semfun{V}$ returns a \emph{single value} $\Value$, while
  the expression function $\semfun{E}$ returns a \emph{collection of values} $\mtype$ (\cref{sec:collection}).
\end{itemize}

The semantics is parameterised over the meaning of a ``collection of values $\mtype$''.
To a first approximation, think of $\mtype$ a (possibly infinite) set of values $\Value$,
with union, intersection etc having their ordinary meaning.

Our first interpretation,
given in \Cref{fig:set-sem}, is a little more refined: $\mtype$ includes $\bot$ and $\wrong$ as
well as a set of values.  Our second interpretation is given in \Cref{fig:labelled-sem}, and
discussed in \Cref{sec:order-sensitive}.

The equations themselves, in \cref{fig:den-sem1}
are beautifully simple and compositional, as a denotational semantics should be.

The equations for $\semfun{V}$ are mostly self-explanatory, but an equation like $\semapp{V}{|k|}{\rho} = k$ needs some
explanation: the $k$ on the left hand side (e.g. ``|3|'') is a piece of \emph{syntax}, but the $k$ on the right is the corresponding
element of the \emph{semantic world of values} $\Value$ (e.g. $3$). As is conventional, albeit a bit confusing,
we use the same $k$ for both.  Same for $op$, where the semantic equivalent is the corresponding mathematical function.

The equations for $\semfun{E}$ are more interesting.
\begin{itemize}
\item Values $\semapp{E}{|v|}{\rho}$: compute the single value for |v|, and return a singleton sequence of results.
The auxiliary function $\munit$ is defined at the bottom of \cref{fig:den-sem1}.
\item In particular, values include lambdas.  The semantics says that a lambda evaluates to a \emph{singleton}
collection, whose only element is a function value. But that function value has
type $\Value \rightarrow \mtype$; that is, it is a function that takes a single
value and returns a \emph{collection} of values.
\item Function application $\semapp{E}{|apply1N v_1 v_2|}{\rho}$ is easy, because $\semfun{V}$ returns a single value: just apply
  the meaning of the function to the meaning of the argument.  The $\apply$ function is defined in \Cref{fig:den-sem1}.
\item Choice $\semapp{E}{|e_1 choice e_2|}{\rho}$: take the union (written $\munion$) of the values
  returned by |e_1| and |e_2| respectively.  For bags this union operator is just bag union (\Cref{fig:set-sem}).
\item Unification $\semapp{E}{|e_1 choice e_2|}{\rho}$: take the \emph{intersection} of the
  values returned by |e_1| and |e_2| respectively. For bags, this ``intersection'' operator $\mintersect$
  is defined in \cref{fig:set-sem}. In this definition, the equality is mathematical equality of functions;
  which we can't implement for functions; see \Cref{sec:unification-and-equality}.
\item Sequencing $\semapp{E}{|e_1 ; e_2|}{\rho}$. Again we use an auxiliary function $\msemi$ to combine
  the meanings of |e_1| and |e_2|. For bags, the function $\msemi$ (\cref{fig:set-sem} again)
  uses a bag comprehension.  Again it does a cartesian product, but without the equality constraint of $\mintersect$.
\item The semantics of |(one e)| simply applies the semantic function $\mone : \mtype \rightarrow \mtype$ to the
  collection of values returned by |e|.  If |e| returns no values, so does |(one e)|; but if |e| returns one or more values,
  |(one e)| returns the first.  Of course that begs the question of what ``the first'' means -- for bags
  it would be non-deterministic. We will fix this problem in \Cref{sec:order-sensitive}, but for now we simply ignore it.
\item The semantics of |(all e)| is similar, but it always returns a singleton collection
  (hence the $\munit$ in the semantics of |all|) whose
  element is a (possibly-empty) tuple that contains all the values in the collection returned by |e|.
\end{itemize}
The fact that unification ``|=|'' maps onto intersection, and choice ``|choice|'' onto union, is
very satisfying.

The big excitement is the treatment of |defKW|. We must extend $\rho$, but what should we bind |x| to?  (Compare the
equation for $\semapp{V}{|lam x e|}{}$, where we have a value $w$ to hand.)
Our answer is simple: \emph{try all possible values, and union the results}:
$$
\semapp{E}{|def x e|}{\rho} \; = \; \bigcup_{w \in \Value}\, \semapp{E}{e}{(\extendrho{\rho}{x \mapsto w})}
$$
That $\bigcup_{w \in \Value}$ means: enumerate all values in $w \in \Value$, in some arbitrary order,
and for each: bind $x$ to $w$, find the semantics of $e$ for that value of $x$, namely
$\semapp{E}{e}{(\extendrho{\rho}{x \mapsto w})}$, and take the union
(in the sense of $\munion$) of the results.

Of course we can't possibly implement it like this, but it makes a great specification.
For example |def x (x=3)| tries all possible values for |x|, but only one of them succeeds, namely |3|, so
the semantics is a singleton sequence $[3]$.

\subsection{The denotational semantics is un-implementable}

This semantics is nice and simple, but we definitely can't implement it!  Consider
$$
|def x ((square x - x - 6) = 0; x|
$$
The semantics will iterate over all possible values for |x|, returning all those that satisfy the equality;
including $3$, for example. But unless our implementation can guarantee to solve quadratic equations,
we can't expect it to return 3.  Instead it'll get stuck.

Another way in which the implementation might get stuck is through unifying functions:
$$
|(lam x (x+x)) = (lam y (y * 2))|
\quad \text{or even} \quad
|(lam x (x+1)) = (lam y (y+1))|
$$
But not all unification-over-functions is ruled out.  We do expect the implementation to succeed with
$$
|def f (((lam x (x+1)) = f); apply1N f 3)|
$$
Here the |defKW| will iterate over all values of |f|, and the equality will pick out the (unique) iteration
in which |f| is bound to the incrementing function.

So our touchstone must be:
\begin{itemize}
\item If the implementation returns a value at all, it must be the value given by the semantics.
\item Ideally, the verifier will guarantee that the implementation does not get stuck, or go $\wrong$.
\end{itemize}

% This reminded Koen of something that Esterelle does.  (Details, Koen?)

\subsection{Getting $\wrong$ right}

Getting $\wrong$ right is a bit tricky.

\begin{itemize}
\item What is the value of |(3 = emptytup)|?  The intersection semantics would say $\mempty$, the empty collection
  of results, but we might want to say $\wrong$.
\item Should $\wrong$ be an element of $\Value$ or of $\mtype$?  We probably want |(one (3 `choice` wrong)|
  to return a $\munit(3)$ rather then $\wrong$?
\item What about |apply1 fst (tup (3,wrong))|?  Is that wrong or 3?
\end{itemize}

There is probably more than one possible choice here.


% -------------------------------------
\subsection{An order-sensitive denotational semantics} \label{sec:order-sensitive}

\newcommand{\LblValue}{LW}
\newcommand{\LblSeq}{[ \Lbl ]}
\newcommand{\Lbl}{L}
\newcommand{\lblleft}{{\bf L}}
\newcommand{\lblright}{{\bf R}}
\newcommand{\lblsemi}[2]{#1 \sconcat #2}
\newcommand{\lblunit}{\bullet}
\newcommand{\domplus}{\, + \,}
\newcommand{\msort}{\mathit{sort}}

\begin{figure}
$$
\begin{array}{l}
\text{\bf Domains} \\
\begin{array}{rcll}
\mtype & = & \mlift{(\wrong + \powerset{\LblValue})} \\
\mtyped & = & \setform{\Value} & \text{Set with 0 or 1 elements} \\
\LblValue & = & \LblSeq \times \Value & \text{Sequence of $\Lbl$ and a value} \\
\Lbl & = & \lblleft \domplus \lblright \\
\end{array} \\
\\
\text{\bf Operations} \\
\begin{array}{rrcll}
\text{Empty}        & \mempty & : &  \mtype \\
    & \mempty & = & \varnothing \\[1mm]
\text{Singleton}    & \munits{.} & : & \Value \rightarrow \mtype \\
     & \munits{w} & = & \setform{ ([], w) } \\[1mm]
\text{Union}        & \munion & : & \mtype \rightarrow \mtype \rightarrow \mtype \\
 &  s_1 \munion s_2 & = & \scompr{(\lblleft \scons l,w)}{\sbind{(l,w)}{s_1}} \cup
                                              \scompr{(\lblright \scons l, w)}{\sbind{(l,w)}{s_2}} \\[1mm]
\text{Intersection} & \mintersect & : & \mtype \rightarrow \mtype \rightarrow \mtype \\
 &  s_1 \mintersect s_2 & = & \scompr{(\lblsemi{l_1}{l_2},w_1)}{\sbind{(l_1,w_1)}{s_1}, \sbind{(l_2,w_2)}{s_2}, w_1=w_2} \\[1mm]
\text{Sequencing} & \msemi & : & \mtype \rightarrow \mtype \rightarrow \mtype \\
&  s_1 \msemi s_2 & = & \scompr{(\lblsemi{l_1}{l_2}, w_2)}{\sbind{(l_1,w_1)}{s_1}, \sbind{(l_2,w_2)}{s_2}} \\[1mm]
% \text{Tagging} & \mtag & : & [L] \times \mtype \rightarrow \mtype \\
% & \mtag(t, s) & = & \setform{(\lblsemi{t}{l}, w) \mid (l, w) \in s} \\[1mm]
\text{One}  & \mone & : & \mtype \rightarrow \mtype \\
& \mone(s) & = & \shead (\msort( s)) \\[1mm]
\text{All}  & \mall & : & \mtype \rightarrow \mtype \\
 & \mall(s) & = & \stuple (\msort( s) ) \\[1mm]
\text{Head}  & \shead & : & [\Value] \rightarrow \mtyped \\
             & \shead [] & = & empty \\
             & \shead (w \scons s) & = & \munits{w} \\[1mm]
\text{To tuple} & \stuple & : & [\Value] \rightarrow \tuple{\Value} \\
     & \stuple [w_1, \cdots, w_n] & = & \tuple{w_1,\cdots,w_n} \\
  \end{array} \\
\\
\begin{array}{rrcll}
\text{Sort} & \msort & : & \LblValue^{*} \rightarrow \mlift{([\Value] + \wrong)}\\
& \msort(s) & = & []  & \text{if $s$ is empty} \\
&  & = & \wrong & \text{if $ws$ has more than one element} \\
&& = & ws & \text{otherwise} \\
  &&& \sconcat \msort \scompr{(l,w)}{\sbind{(\lblleft:l,w)}{s}} \\
  &&& \sconcat \msort \scompr{(l,w)}{\sbind{(\lblright:l,w)}{s}} \\
&& \multicolumn{3}{l}{\text{where $ws = \mcompr{w}{\sbind{([],w)}{s}}$}}
\end{array}

\end{array}
$$
\caption{Labelled set semantics for $\mtype$}
\label{fig:labelled-sem}
\end{figure}

There is a Big Problem with this approach.  Consider |def x (x= (4 choice 3))|.  The existential
enumerates all possible values of x \emph{in some arbitrary order}, and takes the union (i.e. concatention)
of the results from each of these bindings.  Suppose that |defKW| enumerates |3| before |4|; then
the semantics of this expression is the sequence $[3,4]$, and not $[4,3]$ as it should be.
And yet returning a sequence (not a set nor a bag) is a key design choice in Verse.  What can we do?

\Cref{fig:labelled-sem} give a new denotational semantics that \emph{does} account for order.
The key idea (due to Joachim Breitner) is this: return a sequence of \emph{labelled} values;
and then sort that sequence (in $\mone$ and $\mall$) into canonical order
before exposing it to the programmer.

We do not change the equations for $\semfun{E}$, $\semfun{V}$, and $\semfun{O}$ at all; they remain
precisely as they are in \Cref{fig:den-sem1}.
However the semantics of a collection of values, $\mtype$, does change, and is given in \Cref{fig:labelled-sem}:
\begin{itemize}
\item
A collection of values $\mtype$ is now $\bot$ or $\wrong$ (as before), or
a \emph{set of labelled values}, each of type $\LblValue$.
\item A labelled value (of type $\LblValue$)
is just a pair $(\LblSeq \times \Value)$ of a \emph{label} and a value.
\item A label is a sequence of tags $\Lbl$, where a tag is just $\lblleft$ or $\lblright$.
\item The union (or concatention) operation $\munion$, defined in \cref{fig:labelled-sem},
adds a $\lblleft$ tag to the labels of the values in the left branch of the choice,
and a $\lblright$ tag to those coming from the right.
So the labels specify where in the tree the value comes from.
\item Sequencing $\msemi$ and $\mintersect$ both concatenate the labels from the values they combine.
\item Finally $\msort$ puts everything in the ``right'' order: first the values with
an empty label, then the values whose label starts with $\lblleft$ (notice the
recursive sort of the trimmed-down sequence), and then those that start with $\lblright$.
Notice that $\msort$ removes all the labels, leaving just a bare sequence of values $\mtype$.
\item Note that if $\msort$ encounters a set with more than one unlabelled element then this
considered $\wrong$.  This makes ambiguous expressions, like |one (def x x)|, $\wrong$.
\end{itemize}
Let us look at our troublesome example |def x (x= (4 choice 3))|, and assume that
|defKW| binds |x| to |3| and then |4|. The meaning of
this expression will be
$$
\semapp{E}{|def x (x= (4 choice 3))|}{\epsilon} \quad = \quad [ (\lblright,3), (\lblleft,4) ]
$$
Now if we take |allKW| of that expression we will get a singleton sequence
containing |tup (4,3)|, because |allKW| does a sort, stripping off all the tags.
$$
\semapp{E}{|all (def x (x= (4 choice 3)))|}{\epsilon} \quad = \quad [ ([], |tup (4,3)|) ]
$$

\subsection{Related work}

\cite{christiansen:denotational} gives another approach to a denotational semantics
for a functional logic language.  We are keen to learn of others.



% LocalWords:  Lennart Augustsson nnnn Peyton Joachim Breitner ccs ce
% LocalWords:  desc tup optimise defunctionalisation oneKW allKW lll
% LocalWords:  distils metaverse Ariola subst utup elim Hanus lrcl sx
% LocalWords:  residuation Primops isInt exprs CX SX rcl applyN defKW
% LocalWords:  modelled hnf dlet ys xs emptytup lamhdr xrest zs unary
% LocalWords:  boolean un rcll ANF flatMap cartesian behaviour fromTo
% LocalWords:  featureful multivalued mapFlat powerset denotational
% LocalWords:  Launchbury ulit ux localising uc seql seqr unifyl ary
% LocalWords:  unifyr neighbourhood ariola deterministically bagof xy
% LocalWords:  ReWrite CRWL fraguas formalisation rhs dom fvs hanus
% LocalWords:  launchbury fcase labelled splitKW lrcll SML emp Niklas
% LocalWords:  updateable alloc storeTwoLines canCopy storeKW Koen ss
% LocalWords:  subexpression Claessen equationally eliml fcn elimr
% LocalWords:  rll redex Formalising equational stepX compositionally
% LocalWords:  iff cuten stepXs stepXrs Rosser QuickCheck quickcheck
% LocalWords:  joinability metatheory fixpoint liguoro det lago CRLW
% LocalWords:  Smolka flavour FLC urefl onec obfuscatory opez Rodr
% LocalWords:  transactional iguez Hortal Ranjit Jhala ojemo anches
% LocalWords:  Hern andez Scheidecker stepA stepB stepAe stepArs cutr
% LocalWords:  stepAs stepBr stepAr stepAB stepABrs stepABs stepUn
% LocalWords:  Eqns ohlebusch
