%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
% \documentclass[acmsmall,review,anonymous,dvipsnames]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For double-blind review submission, w/ CCS and ACM Reference
\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For single-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission, w/ required CCS and ACM Reference
%\documentclass[acmsmall]{acmart}\settopmatter{}

\let\Bbbk\relax % polycode defines that again?
%include lhs2TeX.fmt
%include lhs2TeX.sty
%include polycode.fmt
%include ../calculus.fmt

\newcommand{\ie}{\emph{i.e.}\xspace}	% \mbox keeps the last period from
\newcommand{\Ie}{\emph{I.e.}\xspace}	% looking like an end-of-sentence.
\newcommand{\eg}{\emph{e.g.}\xspace}
\newcommand{\Eg}{\emph{E.g.}\xspace}
\newcommand{\etc}{\emph{etc.}\xspace}
\newcommand{\etal}{\emph{et al.}\xspace}
\newcommand{\vs}{\emph{vs.}\xspace}

\RequirePackage[utf8]{inputenc}

%% Journal information
%% Supplied to authors by publisher for camera-ready submission;
%% use defaults for review submission.
\acmJournal{PACMPL}
\acmVolume{1}
\acmNumber{ICFP} % CONF = POPL or ICFP or OOPSLA
\acmArticle{1}
\acmYear{2023}
\acmMonth{1}
\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}

%% -------------- Copyright information ------------------
%% Supplied to authors (based on authors' rights management selection;
%% see authors.acm.org) by publisher for camera-ready submission;
%% use 'none' for review submission.
\setcopyright{none}
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\copyrightyear{2018}           %% If different from \acmYear

%% ------------- Bibliography style -------------------
\bibliographystyle{ACM-Reference-Format}
\DeclareUnicodeCharacter{1D706}{$\lambda$}
\DeclareUnicodeCharacter{209B}{${}_s$}
%% Citation style
%% Note: author/year citations are required for papers published as an
%% issue of PACMPL.
\citestyle{acmauthoryear}   %% For author/year citations



%% Some recommended packages.
\usepackage{booktabs}   %% For formal tables:
                        %% http://ctan.org/pkg/booktabs
\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
                        %% http://ctan.org/pkg/subcaption

\usepackage{mathpartir}

\usepackage{stmaryrd}
\usepackage{colortbl}
\usepackage{bussproofs}
\usepackage{mathtools}
% \usepackage{amsmath}
\usepackage{verbatim}
\usepackage[capitalize]{cleveref}   % \cref and \Cref
\usepackage{textgreek} % \textalpha etc.
\usepackage{longtable}

% This makes a tt font that is slightly smaller.
% \usepackage[scaled=0.8]{beramono}

% This incantation makes tt be Latin Modern Mono Light Cond
% which is narrower than normal tt.
%\usepackage{lmodern}
%\DeclareFontFamily{T1}{lmttc}{\hyphenchar \font-1 }
%\DeclareFontShape{T1}{lmttc}{m}{n}
%     {<-> ec-lmtlc10}{}
%\DeclareFontShape{T1}{lmttc}{m}{it}
%     {<->sub*lmttc/m/sl}{}
%\DeclareFontShape{T1}{lmttc}{m}{sl}
%     {<-> ec-lmtlco10}{}
%\renewcommand{\ttdefault}{lmttc}

% ---------------- Boxed figures ------------------
\usepackage{float}
\usepackage{xspace}
\usepackage{quiver}
\floatstyle{boxed}
\restylefloat{figure}

% ---------------- Full Verse ----------------
\newif\iffullverse
\fullversefalse
% use \iffullverse \else \fi

% ---------------- Useful macros ----------------
\renewcommand{\arraystretch}{1.1}

% ----------------- Author comments ---------------------
\definecolor{dkcyan}{rgb}{0.1, 0.3, 0.3}
\definecolor{dkgreen}{rgb}{0,0.3,0}
\definecolor{olive}{rgb}{0.5, 0.5, 0.0}
\definecolor{dkblue}{rgb}{0,0.1,0.5}

\iftrue  % iftrue to enable, iffalse to disable
\newcommand{\simon}[1]{{\color{dkblue} [{SLPJ}: #1]}}
\newcommand{\lennart}[1]{{\color{dkcyan} [{LA}: #1]}}
\newcommand{\joachim}[1]{{\color{dkblue} [{JB}: #1]}}
\newcommand{\koen}[1]{{\color{dkred} [{KC}: #1]}}
\newcommand{\olin}[1]{{\color{dkgreen} [{OS}: #1]}}
\newcommand{\ranjit}[1]{{\color{orange} [{RJ}: #1]}}
\newcommand{\guy}[1]{{\color{red} [{GS}: #1]}}
\else
\newcommand{\simon}[1]{}
\newcommand{\lennart}[1]{}
\newcommand{\joachim}[1]{}
\newcommand{\koen}[1]{}
\newcommand{\olin}[1]{}
\newcommand{\ranjit}[1]{}
\newcommand{\guy}[1]{}
\fi

% ---------------- End of comments -----------------

% Loads the rewrite figures.
% They still need to be placed using \figureGeneralAxioms etc.
\input{rewrites}


\begin{document}

%% Title information
% \title{The essence of functional logic programming}
\title{The Verse Calculus: a Core Calculus for Functional Logic Programming}

                                        %% [Short Title] is optional;
                                        %% when present, will be used in
                                        %% header instead of Full Title.
% \titlenote{with title note}             %% \titlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'
% \subtitle{Subtitle}                     %% \subtitle is optional
% \subtitlenote{with subtitle note}       %% \subtitlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'


%% Author information
%% Contents and number of authors suppressed with 'anonymous'.
%% Each author should be introduced by \author, followed by
%% \authornote (optional), \orcid (optional), \affiliation, and
%% \email.
%% An author may have multiple affiliations and/or emails; repeat the
%% appropriate command.
%% Many elements are not rendered, but should be provided for metadata
%% extraction tools.

%% Author with single affiliation.
\author{Lennart Augustsson}
%\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
%  \position{Position1}
  \institution{Epic Games}            %% \institution is required
%   \streetaddress{Street1 Address1}
%  \city{City1}
%  \state{State1}
%  \postcode{Post-Code1}
  \country{Sweden}                    %% \country is recommended
}
\email{lennart.augustsson@@epicgames.com}          %% \email is recommended

\author{Joachim Breitner}
% \authornote{with author1 note}          %% \authornote is optional;
%                                         %% can be repeated if necessary
\orcid{0000-0003-3753-6821}             %% \orcid is optional
\affiliation{}
%  \position{Position1}
%   \institution{Epic Games}            %% \institution is required
%  \streetaddress{Street1 Address1}
%  \city{City1}
%  \state{State1}
%  \postcode{Post-Code1}
%  \country{Germany}                    %% \country is recommended
% }
\email{mail@@joachim-breitner.de}          %% \email is recommended

\author{Koen Claessen}
% \authornote{with author1 note}          %% \authornote is optional;
%                                         %% can be repeated if necessary
%\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
%  \position{Position1}
  \institution{Epic Games}            %% \institution is required
%  \streetaddress{Street1 Address1}
%  \city{City1}
%  \state{State1}
%  \postcode{Post-Code1}
  \country{Sweden}                    %% \country is recommended
}
\email{koen.claessen@@epicgames.com}          %% \email is recommended

\author{Ranjit Jhala}
% \authornote{with author1 note}          %% \authornote is optional;
%                                         %% can be repeated if necessary
%\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
%  \position{Position1}
  \institution{Epic Games}            %% \institution is required
%  \streetaddress{Street1 Address1}
%  \city{City1}
%  \state{State1}
%  \postcode{Post-Code1}
  \country{USA}                    %% \country is recommended
}
\email{ranjit.jhala@@epicgames.com}          %% \email is recommended

\author{Simon Peyton Jones}
% \authornote{with author1 note}          %% \authornote is optional;
%                                         %% can be repeated if necessary
%\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
%  \position{Position1}
  \institution{Epic Games}            %% \institution is required
%  \streetaddress{Street1 Address1}
%  \city{City1}
%  \state{State1}
%  \postcode{Post-Code1}
  \country{United Kingdom}                    %% \country is recommended
}
\email{simonpj@@epicgames.com}          %% \email is recommended

\author{Olin Shivers}
% \authornote{with author1 note}          %% \authornote is optional;
%                                         %% can be repeated if necessary
%\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
%  \position{Position1}
  \institution{Epic Games}            %% \institution is required
%  \streetaddress{Street1 Address1}
%  \city{City1}
%  \state{State1}
%  \postcode{Post-Code1}
  \country{USA}                    %% \country is recommended
}
\email{olin.shivers@@epicgames.com}          %% \email is recommended

\author{Guy Steele}
% \authornote{with author1 note}          %% \authornote is optional;
%                                         %% can be repeated if necessary
%\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
%  \position{Position1}
  \institution{Oracle Labs}            %% \institution is required
%  \streetaddress{Street1 Address1}
%  \city{City1}
%  \state{State1}
%  \postcode{Post-Code1}
  \country{USA}                    %% \country is recommended
}
\email{guy.steele@@oracle.com}          %% \email is recommended

\author{Tim Sweeney}
% \authornote{with author1 note}          %% \authornote is optional;
%                                         %% can be repeated if necessary
%\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
  \position{CEO}
  \institution{Epic Games}            %% \institution is required
%  \streetaddress{Street1 Address1}
%  \city{City1}
%  \state{State1}
%  \postcode{Post-Code1}
  \country{USA}                    %% \country is recommended
}
\email{tim.sweeney@@epicgames.com}          %% \email is recommended



%% ---------------- Abstract ---------------------
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
  Functional logic languages have a rich literature, but it is tricky
  to give them a satisfying semantics.  In this paper we describe the
  Verse calculus, \versecalc{}, a new core calculus for functional
  logical programming. Our main contribution is to equip \versecalc{} with a
  small-step rewrite semantics, so that we can reason
  about a \versecalc{} program in the same way as one does with lambda
  calculus; that is, by applying successive rewrites to it.
  We also show that the rewrite system is confluent.
%   This semantics elegantly fills the gap between very high level
%  ``magical'' semantics, and very low-level operational semantics.
  \par
  \vspace{1em}
\end{abstract}


%% ---------------- Classification ---------------------
%% 2012 ACM Computing Classification System (CSS) concepts
%% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10011007.10011006.10011008</concept_id>
<concept_desc>Software and its engineering~General programming languages</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10003456.10003457.10003521.10003525</concept_id>
<concept_desc>Social and professional topics~History of programming languages</concept_desc>
<concept_significance>300</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Software and its engineering~General programming languages}
\ccsdesc[300]{Social and professional topics~History of programming languages}
%% End of generated code


%% Keywords
%% comma separated list
% \keywords{keyword1, keyword2, keyword3}  %% \keywords are mandatory in final camera-ready submission
\keywords{%
  confluence,
  declarative programming,
  functional programming,
  lambda calculus,
  logic programming,
  rewrite rules,
  skew confluence,
  unification,
  Verse calculus,
  Verse language
}

%% \maketitle
%% Note: \maketitle command must come after title commands, author
%% commands, abstract environment, Computing Classification System
%% environment and commands, and keywords command.
\maketitle



% \textbf{Note} \textit{This is a working draft.  It's not secret, but it's very much
%   a work in progress.  So please don't share it; if there is anyone you think would be
%   interested/helpful, do suggest them to us.}

\section{Introduction}

Functional logic programming languages add expressiveness to functional programming
by introducing logical variables, equality constraints among those variables,
and choice to allow multiple alternatives to be explored.  Here is a tiny example:
$$
|def (x ^^ y ^^ z) (x==tup(y,3); x==tup(2,z); y)|
$$
This expression introduces three logical (or existential) variables |x,y,z|,
constrains them with two equalities |(x==tup(y,3)| and |(x==tup(2,z))|, and finally returns |y|.
The only solution to the two equalities is |y==2|, |z==3|, and |x==tup(2,3)|; so the result of the
whole expression is |2|.

Functional logic programming has a long history and a rich literature.
But it is somewhat tricky for programmers to \emph{reason} about
functional logic programs: they must think about logical variables,
narrowing, backtracking, Horn clauses, resolution, and the like.  This
contrasts with functional programming, where one can say ``just apply
rewrite rules, such as \textbeta-reduction, let-inlining, and
case-of-known-constructor.''
% Moreover, these rewrite rules apply even
% in the presence of free variables; {\eg}, under lambdas.
We therefore seek \emph{a precise expression of functional logic programming as
a term-rewriting system}, to give us both a formal semantics (via small-step reductions),
and a powerful set of equivalences that programmers can use to reason about their programs,
and that compilers can use to optimize them.

We make the following contributions in this paper.
First, we describe a new core calculus for functional logic programming,
the Verse calculus or \versecalc{} for short
(\cref{sec:informal}).
As in any functional logic language,
\versecalc{} supports logical variables, equalities, and choice, but it is distinctive in several ways:
\begin{itemize}
\item \versecalc{} natively supports \emph{higher-order functions}, just like the lambda calculus.
  Indeed, \emph{every lambda calculus program is a \versecalc{} program}.
  In contrast, most of the functional logic literature is rooted in a first-order world, and
  addresses higher-order features via an encoding called defunctionalization~\cite[3.3]{hanus:survey}.
\item All functional logic languages have some notion of ``flexible'' {\vs} ``rigid'' variables.
  \versecalc{} offers a new
  way to address these notions, through the operators |oneKW| (\cref{sec:one})
  and |allKW| (\cref{sec:all}).  This enables an elegant economy of concepts: for example,
  there is just one equality (other languages may have a suspending equality and a narrowing equality),
  and conditional expressions are driven by failure rather than booleans (\cref{sec:if}).
\item \emph{Choice and determinism.} Choice is a fundamental feature of all functional logic languages.
  In \versecalc{}, choice is expressed in the syntax of the term (``laid out in space'')
  rather than, as is more typical, handled by non-deterministic rewrites and backtracking (``laid out in time'').
  This makes \versecalc{} completely \emph{deterministic}, unlike most
  functional logic languages which are non-deterministic by design (\cref{sec:deterministic}).
\end{itemize}
As always with a calculus, the idea is that \versecalc{} distills the essence of functional logic programming.
Each construct does just one thing, and \versecalc{} cannot be made smaller without losing key features.
We believe that it is possible to use \versecalc{} as the compilation target for a variety of
functional logic languages such as Curry~\cite{hanus:curry-report}.
We are ourselves working on Verse, a new general purpose programming language, built directly on \versecalc{}; indeed,
our motivation for developing \versecalc{} is practical rather than theoretical.
No single aspect of \versecalc{} is unique, but we believe that their combination
is particularly harmonious and orthogonal.
We discuss the related work in \cref{sec:related}, and design alternatives in \cref{sec:variations}.

Our second contribution is to equip \versecalc{} with a
\emph{small-step term-rewriting semantics} (\cref{sec:rewrite-rules}).
We said that the lambda calculus is a subset of \versecalc{}, so it is natural to give its
semantics using rewrite rules, just like the lambda calculus.  That seems challenging, however, because
logical variables and unification involve sharing and non-local communication. How can that
be expressed in a rewrite system?

Happily, we can build on prior work: exactly the same difficulty arises with call-by-need in the lambda calculus.
For a long time, the only semantics of
call-by-need that was faithful to its sharing semantics (in which thunks are evaluated at most once)
was an operational semantics that sequentially threads a global heap through execution~\cite{launchbury:natural}.
But then Ariola {\etal}, in a seminal paper, showed how to \emph{reify the heap into the term itself},
and thereby build a rewrite system that is completely faithful to lazy evaluation~\cite{ariola:call-by-need}.
Inspired by their idea, we present a new rewrite system for functional logic programs that reifies
logical variables and unification into the term itself,
and replaces non-deterministic search with a (deterministic) tree of successful results.
% Just like rewriting for the lambda calculus, this semantics
% gives programmers a set of equivalences with which
% they can reason equationally about their programs.
For example, the expression above can be rewritten
as follows\footnote{The rule names come from \cref{fig:rules}, to be discussed in \cref{sec:rewrite-rules}; they
  are given here just for reference.}:
$$
\begin{array}{ll@@{\quad}ll}
\multicolumn{4}{l}{|def (x ^^ y ^^ z) (x == tup(y,3); x == tup(2,z); y)|} \\
\movestoR{subst}         & |def (x ^^ y ^^ z) (tup(2,z) == tup(y,3); x == tup(2,z); y)| &
\movestoR{eqn-elim}      & |def (y ^^ z) (tup(2,z) == tup(y,3); y)| \\
\movestoR{u-tup}         & |def (y ^^ z) (2==y; z==3; y)| &
\movestoR{eqn-elim}      & |def y (2==y; y)| \\
\movestoR{hnf-swap}      & |def y (y==2; y)| &
\movestoR{subst}         & |def y (y==2; 2)| \\
\movestoR{eqn-elim}      & |2| \\
\end{array}
$$
%
Rules may be applied anywhere they match, again just like the lambda calculus.
This freedom only makes sense, however, if each term ultimately reduces to a unique value,
regardless of its reduction path,
so we show that \versecalc{} is \emph{confluent}, in \cref{sec:metatheory}.

% There is a rich seam of related work, as we discuss in \cref{sec:related}.

\section{The Verse calculus, informally} \label{sec:informal}

\figureSyntax

We begin by presenting the Verse calculus, \versecalc, informally.  We will describe its
rewrite rules precisely in \cref{sec:rewrite-rules}.
The (abstract) syntax of \versecalc{} is given in \cref{fig:syntax}.
It has a very conventional sub-language that is just the lambda calculus with
some built-in operations and tuples as data constructors:
\begin{itemize}
\item \emph{Values.}  A value |v| is either a variable |x| or a head-normal form |hnf|.
  In \versecalc{}, a variable counts as a value because in a functional logic language an expression
  may evaluate to an as-yet-unknown logical variable.
  A head-normal form is a conventional value: a built-in constant |k|, an operator |op|, a tuple, or
  a lambda.  Our tiny calculus offers only integer constants |k| and two illustrative integer operators |op|, namely |gt| and |add|.
\item \emph{Expressions} |e| includes values |v|, and applications |apply1N v_1 v_2|; we will
  introduce the other constructs as we go.
For clarity, we often write |apply1 v_1 v_2| rather than |apply1N v_1 v_2| when |v_2| is not a tuple.
\item A term |eq| is either an ordinary expression |e|, or
an \emph{equation} |v==e|; this syntax ensures that equations can only occur to the left of a ``|;|'' (\cref{sec:unify}).
\item \emph{A program}, |p|, is a closed expression from which we extract one result using |oneKW| (see \cref{sec:one}).
%%  \lennart{It's not really supposed top be |OneKW|, but |succeeds|.  But we don't have |succeeds|.}
\end{itemize}
The formal syntax for |e| allows only applications of \emph{values}, |(v_1 v_2)|, but the desugaring
rules in \cref{fig:syntax} show how to desugar more applications |(e_1 e_2)|. This ANF-like
normalization is not fundamental; it simply reduces the number of rewrite rules we need.
The desugaring rules are more suggestive than precise; we aim to be precise about \versecalc{} but
less so about the source language.

Modulo this desugaring, every lambda calculus term is a \versecalc{} term,
and has the same semantics. Just like the lambda calculus, \versecalc{} is untyped;
adding a type system is an excellent goal, but is the subject of another paper.

Expressions also include two other key collections of constructs:
logical variables and unification (\cref{sec:unify}), and choice (\cref{sec:choice}).
The details of choice and unification, and especially their
interaction, are subtle, so this section will do a lot of
arm-waving.  But fear not: \cref{sec:rewrite-rules} will spell out the precise details.
We only have space to describe one incarnation of \versecalc{}; \cref{sec:variations}
explores some possible alternative design choices.

\subsection{Logical variables and equations} \label{sec:unify}

The Verse calculus includes first class \emph{logical variables} and \emph{equations} that constrain their values.
You can bring a fresh logical variable into scope with |defKW|, constrain a value to be equal to
an expression with an equation
|v == e|, and compose expressions in sequence with |eq ; e| (see \cref{fig:syntax}).
As an example, what might be written |llet x e_1 e_2| in a conventional functional language can be written
|def x (x==e_1; e_2)| in \versecalc{}.  The syntax carefully constrains both the form of
equations and where they can appear: an equation |(v==e)| always equates a \emph{value} $v$ to an expression $e$;
and an equation can only appear to the left of a ``|;|'' (see |eq| in \cref{fig:syntax}).
The desugaring rules in \cref{fig:syntax} rewrite a general equation |e_1 == e_2| into this simpler form.

\emph{A program executes by solving its equations}, using the process of unification.
For example,
$$
|def (x ^^ y ^^ z) (x==tup(y,3); x==tup(2,z); y)|
$$
is solved by unifying |x| with |tup(y,3)| and with |tup(2,z)|; that in turn unifies |tup(y,3)| with |tup(2,z)|, which
unifies |y| with |2| and |z| with |3|.  Finally, |2| is returned as the result.
Note carefully that, as in any declarative language,
\emph{logical variables are not mutable}; a logical variable stands for a single, immutable value.
We use ``|defKW|'' to bring a fresh logical variable into scope, because we really
mean ``there exists an |x| such that |xdots|.''
%% Logical variables \emph{are} existential variables.

High-level functional languages usually provide some kind of
pattern matching; in such a language, we might define |first| by
|applyN first (a,b) == a|. Such pattern matching is typically desugared to
more primitive |case| expressions, but in \versecalc{} we do not need |case| expressions:
unification does the job.  For example we can define |first| like this:
$$
|first :== lam p (def (a ^^ b) (p == tup (a,b); a))|
$$
% The lambda unifies its argument (presumably a pair)
%with |tup(a,b)|, where |a,b| are fresh logical
% variables in the body of the lambda; and returns the first of the two, namely |a|.
For convenience, we allow ourselves to write a term like
|applyN first (2,5)|, where we define the library function |first| separately with ``|:==|.'' Formally, you can imagine
each example |e| being wrapped with a binding for |first|, thus |def first (first == ...); e|,
and similarly for other library functions.

This way of desugaring pattern matching means that the input
to |first| is not required to be fully determined when the function is called. For example:
$$
|def (x ^^ y) (x == tup(y,5); 2 == apply1 first x; y)|
$$
Here |apply1 first x| evaluates to |y|, which we then unify with |2|.
Another way to say this is that, as usual in logic programming, we may
constrain the \emph{output} of a function (here |2 == apply1 first x|), and thereby
affect its \emph{input} (here |tup(y,5)|).

Although ``|;|\!'' is called ``sequencing,'' the order of that sequence
is immaterial for equations that do not contain choices (see \Cref{sec:choice} for the latter caveat).
For example,
% foothand not needed, the figure introduces this shorthand
%\footnote{We informally write |3+y| rather than |applyN add (3,y)|, and similarly with other operators.},
consider |(def (x ^^ y) (x == 3+y; y == 7; x))|.  The sub-expression |3+y| is stuck until |y| gets a value.
In \versecalc{}, we can only unify |x| with a \emph{value}---we will see why in \cref{sec:call-time-choice}---and
hence the equation |x==3+y| is stuck.
No matter!  We simply leave it and try some other equation. In this case, we can make progress with |y==7|,
and that in turn unlocks |x == 3+y| because now we know that |y| is |7|, so we can evaluate |3+7| to |10|
and unify |x| with that.  The idea of leaving stuck expressions aside and executing other parts of the
program is called
\emph{residuation}~\cite{hanus:survey}\footnote{Hanus did not invent the terms ``residuation'' and``narrowing,'' but
his survey is an excellent introduction and bibliography.}, and is at the heart of our mantra ``just solve the equations.''

\subsection{Choice} \label{sec:choice} \label{sec:call-time-choice}

In conventional functional programming, an expression evaluates to a single value.
In contrast, a \versecalc{} expression evaluates to zero, one, or many values;
or it can get stuck, which is different from producing zero values.
The expression |fail| yields no values; a value |v| yields one value;
and the choice |e_1 choice e_2| yields all the values
yielded by |e_1| followed by all the values yielded by |e_2|.
Order is maintained and duplicates are not eliminated; we shall see why in \cref{sec:for}.
In short, an expression yields a \emph{sequence} of values, not a bag, and certainly not a set.

The equations we saw in \cref{sec:unify} can fail, if the arguments
are not equal, yielding no results.  Thus |3==3| succeeds,
while |3==4| fails, returning no results.  In general, we use ``fail''
and ``returns no results'' synonymously.

What if the choice was not at the top level of an expression?  For example,
what does |tup (3, (7 choice 5))| mean?
In \versecalc{}, it does \emph{not} mean a pair with some kind of multi-value in its second component.
Indeed, as you can see from \cref{fig:syntax}, this expression is syntactically ill-formed.
We must instead give a name to that choice, and then we can put it in the pair, thus: |def x (x == (7 choice 5); tup (3,x))|.
Now the expression is syntactically legal, but what does it mean?  In \versecalc{}, a variable is never bound to a multi-value.
Instead, |x| is successively bound to |7|, and then to |5|, like this:
$$
|def x (x == (7 choice 5); tup (3,x))| \quad \movesto \quad |(def x (x==7; tup (3,x))) choice (def x (x==5; tup (3,x))))|
$$
We duplicate the context surrounding the choice, and ``float the choice outwards.''
The same thing happens when there are multiple choices.  For example:
$$
|def (x ^^ y) (x == (7 choice 22); y == (31 choice 5); tup (x,y))|
\quad
\text{yields the sequence |tup (7,31), tup(7,5), tup(22,31), tup(22,5)|}
$$
Notice that the order of the two equations now \emph{is} significant:
$$
|def (x ^^ y) y == (31 choice 5); x == (7 choice 22); tup (x,y)|
\quad
\text{yields the sequence |tup (7,31), tup(22,31), tup(7,5), tup(22,5)|}
$$
Readers familiar with list comprehensions in Haskell and other
languages will recognize this nested-loop pattern, but here it
emerges naturally from choice as a deeply built-in primitive,
rather than being a special construct for lists.

Just as we never bind a variable to a multi-value, we never bind it to |fail| either; rather
we iterate over zero values, and that iteration of course returns zero
values.  So:
$$
|def x (x == fail; 33)| \quad \movesto \quad |fail|
$$

\subsection{Mixing choice and equations} \label{sec:mixing}

In the last section, we discussed what happens if there is a choice in the right-hand side (RHS) of
an equation. What if we have equations under choice?  For example:
$$
|def x ((x==3; x+1) choice (x==4; x*2))|
$$
Intuitively, ``either unify |x| with |3| and yield |x+1|, or unify |x| with |4| and yield |x*2|''.
But there is a problem: so far we have said only ``a program executes by solving its equations'' (\cref{sec:unify}).
Here, we can see two equations, |(x==3)| and |(x==4)|, which are mutually contradictory,
so clearly we need to refine our notion of ``solving.''
The answer is pretty clear: in a branch of a choice, solve the equations in that branch
to get the value for some logical variables, \emph{and propagate those values to occurrences in
  that branch (only)}. Occurrences of that variable outside the choice are unaffected.  We call
this \emph{local propagation}.
This local-propagation rule would allow us to reason thus:
$$
|def x ((x==3; x+1) choice (x==4; x*2))| \movesto  |def x ((x==3; 4) choice (x==4; 8))|
$$
Are we stuck now?  No, we can float the choice out as
before\footnote{Indeed, we could have done so first, had we wished.},
$$
|def x ((x==3; 4) choice (x==4; 8))| \movesto |(def x (x==3; 4)) choice (def x (x==4; 8))|
$$
and now it is apparent that the sole occurrence of |x| in each |defKW| is the equation
|(x==3)|, or |(x==4)| respectively; so we can drop the |defKW| and the equation, yielding |(4 choice 8)|.
% To take a celebrated example, suppose
% |square = lam x (x*x)|.  What does |apply1 square (3 choice 4)| mean? If we make the choice \emph{before}
% the call, the call will yield |9 choice 16|.  But if we a more call-by-name semantics, we might imagine
% it yields |(3 choice 4) * (3 choice 4)| which in turn yields |(9 choice 12 choice 12 choice 16)|.
% All functional-logic languages, including

\subsection{Pattern matching and narrowing}

We remarked in \cref{sec:unify} that we can desugar the pattern matching of a high-level
language into equations. But what about multi-equation pattern matching, such as this definition in Haskell:
\begin{code}
  append []       ysl == ysl
  append (x:xsl)  ysl == x : append xsl ysl
\end{code}
If pattern matching on the first equation fails, we want to fall through to the second.
Fortunately, choice allows us to express this idea directly, where
we use the empty tuple |emptytup| to represent the empty list and pairs to represent cons cells
(see \cref{fig:syntax} to desugar the pattern matching lambda):
$$
|append :== lam (tup (xsl,ysl))
  (((xsl == emptytup; ysl) choice
   (def (x ^^ xr) (xsl == tup (x,xr); tup (x, applyN append (xr, ysl))))))|
$$
If |xsl| is |emptytup|, the left-hand choice succeeds, returning |ysl|; and the right-hand choice
fails (by attempting to unify |emptytup| with |tup(x,xr)|).
If |xsl| is of the form |tup (x,xr)|,
the right-hand choice succeeds, and we make a recursive call to |append|.
Finally, if |xsl| is built with head-normal forms other than the empty tuple and pairs, both choices fail, and |append| returns no
results at all.

This approach to pattern matching is akin to \emph{narrowing}~\cite{hanus:survey}.
Suppose |single == tup (1, emptytup)|, a singleton list whose only element is 1.
Consider the call |def zsl (applyN append ( zsl, single) == single; zsl)|.
The call to |append| expands into a choice:
$$
  |(zsl == emptytup; single)| \quad |choice| \quad
  |(def (x ^^ xr) (zsl == tup (x,xr); tup(x, applyN append (xr, single))))|
$$
which amounts to exploring the possibility that |zsl| is headed by |emptytup| or a pair---the
essence of narrowing.  It should not take long to reassure yourself that the program
evaluates to |emptytup|, effectively running |append| backwards in the classic logic-programming manner.

This example also illustrates that \versecalc{} allows an equation (for |append|)
that is recursive. As in any functional language with recursive bindings, you can
go into an infinite loop if you keep fruitlessly inlining the function in its own right-hand side.
It is the business of an \emph{evaluation strategy} to do only rewrites that make
progress toward a solution (\cref{sec:eval-strategy}).

\subsection{Conditionals and \texorpdfstring{|oneKW|}{``one''}} \label{sec:rigid} \label{sec:one} \label{sec:if}

Every source language will provide a conditional, such as |if (x==0) then
e_2 else e_3|.  But what is the equality operator in |(x==0)|?  One
possibility, adopted by Curry, is this: there is one ``='' for
equations (as in \cref{sec:unify}), and another, say ``=='', for
testing equality (returning a boolean with constructors |True| and |False|).  \versecalc{} takes a
different, more minimalist position, following Icon's lead, see \cref{sec:icon}.
In \versecalc{}, \emph{there is just one equality operator}, written ``='' just as in
\cref{sec:unify}. The expression |if (x==0) then e_2 else e_3| tries to unify |x| with |0|.  If that succeeds (yields one or more values),
the |if| returns |e_2|; otherwise it returns |e_3|.  There are no data constructors |True| and |False|; instead
failure (returning zero values) plays the role of falsity.

But something is terribly wrong here.  Consider |def (x ^^ y) (y == (if (x==0) then 3 else 4); x==7)|.  Presumably this is meant
to set |x| to |7|, test whether it is equal to |0| (it is not), and unify |y| with 4.
But what is to stop us instead unifying |x| with |0| (via |(x==0)|), unifying |y| with |3|, and then
failing when we try to unify |x| with |7|?  Not only is that not what we intended, but it also looks very
non-deterministic: the result is affected by the order in which we did unifications.

To address this, we give |if| a special property: in the expression |if e_1 then e_2 else e_3|, equations inside |e_1| (the condition of the |if|)
can only unify variables bound inside |e_1|; variables bound outside |e_1| are called ``rigid.''  So in our example, the |x| in |(x==0)| is
rigid and cannot be unified.  Instead, the |if| is stuck, and we move on to unify |x==7|.  That unblocks
the |if| and all is well.
% This special property is precisely the local propagation rule that we sketched for
% choice (\cref{sec:mixing}).

In fact, \versecalc{} desugars the three-part |if| into something simpler, the unary construct |one(e)|.
Its specification is this:
if |e| fails, |one(e)| fails; otherwise |one(e)| returns the first of the values yielded by |e|.
Now, |if e_1 then e_2 else e_3| can (nearly) be re-expressed like this:
$$
|one( (e_1; e_2) choice e_3 )|
$$
This isn't right yet, but the idea is this: if |e_1| fails, the first branch of the choice fails, so we get |e_3|; if |e_1| succeeds, we
get |e_2|, and the outer |oneKW| will select it from the choice.
But what if |e_2| or |e_3| \emph{themselves} fail or return multiple results?  Here is a better translation,
given in \cref{fig:syntax}\footnote{The translation in the figure also allow variables bound in the condition to scope over the |then| branch.}, which wraps the |then| and |else| branches
in a thunk\footnote{Using thunks for the branches of a conditional is another very old idea;
for example, see \cite[p.\ 54]{steele:rabbit}.}:
$$
|apply0 (one( (e_1; (thunk (e_2))) choice (thunk (e_3))))|
$$
The argument of |oneKW| reduces to either |(thunk e_2) `choice` (thunk e_3)| or |(thunk e_3)| depending on whether |e_1| succeeds or fails,
respectively; |oneKW| then picks the first value, that is |thunk e_2| if |e_1| succeeded, or |thunk e_3| if |e_1| failed, and
applies it to |tup ()|.
As a bonus, provided we do no evaluation under a lambda, then |e_2| and |e_3| will
remain unevaluated until the choice is made, just as we expect from a conditional.

We use the same local-propagation rule for |oneKW| that we do for choice (\cref{sec:mixing}).
This, together with the desugaring for |if| into |oneKW|, gives the ``special property'' of |if| described above.

\subsection{Tuples and \texorpdfstring{|allKW|}{``all''}} \label{sec:tuple}

The main data structure in \versecalc{} is the \emph{tuple}.
A tuple is a finite sequence of values, |tup (v_1,dots,v_n)|, where $n\geq 0$.
A tuple can be used like a function: indexing is simply function application
with the argument being integers from 0 and up.  Indexing out of range is |fail|, as is indexing with
a non-integer value.
For example, |t :== tup (10,27,32); apply1 t 1| reduces to |27| and |apply1 t 3| reduces to |fail|.

What if we apply a tuple to a choice, thus |apply1 (tup (10,27,32))(1 `choice` 0 `choice` 1)|?  First we must
desugar the application to the form |(apply1N v_1 v_2)|, because that is all \versecalc{} permits (\cref{fig:syntax}),
giving |x:==(1 `choice` 0 `choice` 1); apply1(tup (10,27,32)) x|, which readily reduces to |(27 `choice` 10 `choice` 27)|.

Tuples can be constructed by collecting all the results from a multi-valued expression, using the |allKW| construct:
if |e| reduces to |(v_1 `choice` xdots `choice` v_n)|, where $n \geq 2$, then |all e| reduces to
the tuple |tup (v_1, dots, v_n)|; |all v| produces the singleton tuple |tup v|; and |all fail| produces the empty tuple |emptytup|.
Note that |choice| is associative, which means that we
can think of a sequence or tree of binary choices as really being a single
$n$-way choice.
\label{sec:all}

You might think that tuple indexing would be stuck until we know the index,
but in \versecalc{}, the application of a tuple to a value rewrites to a choice of all the possible
values of the index.
For example,
|t :== tup (10,27,32); def i (apply1 t i)|
looks stuck because we have no value for |i|, but actually rewrites to:
$$
|def i ((i==0; 10) choice (i==1; 27) choice (i==2; 32)|
$$
which (as we will see in \cref{sec:rewrite-rules}) simplifies
to just |(10 `choice` 27 `choice` 32)|.
So |allKW| allows a choice to be reified into a tuple; and |(def i (apply1 t i))| allows a
tuple to be turned back into a choice.
The idea of rewriting a call of a function with a finite domain into a finite choice
is called ``narrowing'' in the literature.

Do we even need |oneKW| as a primitive construct, given that we have
|allKW|? Can we not use |apply1 (all e) 0| instead of |one e|? Indeed,
they behave the same if |e| fully reduces to finitely many choices of
values. But |allKW| really requires \emph{every} arm of the choice tree to resolve to a value
before proceeding, while |oneKW| only needs the
\emph{first} choice to be a value. So, supposing that |loop| is a non-terminating function,
|one (1 `choice` apply0 loop)| can reduce to 1, while
|apply1 (all (1 `choice` apply0 loop)) 0| loops.

\subsection{Programming in Verse}  \label{sec:verse}

\versecalc{} is a fairly small language, but it is quite expressive.
For example, we can define the typical list functions one would expect from functional programming
by using the duality between tuples and choices, as seen in \cref{fig:list-functions}.
A tuple can be turned into choices by indexing with a logical variable |i|.
Conversely, choices can be turned into a tuple using |allKW|.
The choice operator, |choice|, serves as both |cons| and |append| for choices;
the corresponding operations for tuples are defined in \cref{fig:list-functions}.
Partial functions, \eg, |head|, will |fail| when the argument is outside of the domain.

% \subsection{Syntactic sugar}
%
% \begin{figure}
% % no space needed around math environments in a figure
% \setlength{\abovedisplayskip}{0pt}%
% \setlength{\belowdisplayskip}{0pt}%
%   $$
%     \begin{array}{r@@{\;}c@@{\;}ll}
%       |x :== e; dots| & \equiv & |def x (x = e; dots)| \\
%       |apply1 f x :== e; dots| & \equiv & |f :== lam x e; dots| \\
%       |applyN f (x,y) :== e; dots| & \equiv & |apply1 f p :== (def (x ^^ y) (p=tup (x,y); e)); dots| & \text{|p| fresh} \\
%       |apply1 e_1 e_2| & \equiv & |def (x ^^ y) (x = e_1;y=e_2;apply1 x y)| & \text{|x,y| fresh} \\
%       |tup (e_1, e_2)| & \equiv & |def (x ^^ y) (x = e_1;y=e_2; tup (x,y))| & \text{|x,y| fresh} \\
% % already in the first figure
% %      |e_1+e_2| & \equiv & |applyN add (e_1,e_2)| \\
% %      |e_1>e_2| & \equiv & |applyN gt (e_1,e_2)| \\
%       |if e_1 then e_2 else e_3| & \equiv & |v :== one( (e_1; thunk e_2) choice (thunk e_3)); apply0 v| & \text{|v,x| fresh}
% % only used in sectino about for, and there the equation is repeated
% %      |for e_1 do e_2| & \equiv & |v :== all (e_1; thunk e_2); applyN map (lam z (apply0 z), v)| & \text{|v,x| fresh} \\
%     \end{array}
%   $$
% \caption{Syntactic sugar}
% \label{fig:sugar}
% \end{figure}
%
% The {\versecalc} grammar can be a bit unwieldy, with explicit |defKW| and ANF-like restrictions allowing only values,
% not arbitrary expressions, in tuple literals and applications \cite{felleisen:anf}.
% To keep the examples readable, we allow ourselves a bit of syntactic sugar, summarized in \cref{fig:sugar}:
% \begin{itemize}
% \item With |:==| we simultaneously introduce and ``assign'' a variable
%   an initial value. (Which later unification and choice may sharpen up
%   or constrain.)
% \item When defining functions, we allow simple patterns on the left of |:==|
% \item We allow the use of |if|, as explained in \cref{sec:if}.
% % \item We allow the use of |for|, as we will explain in \cref{sec:for}.
% \end{itemize}

% \subsection{List functions}

\begin{figure}
  $$
  \begin{array}{l}
    \textbf{Desugaring}\\
    \begin{array}{rcll}
      |apply1 f x :== e| & \text{means} & |f :== lam x e| \\
      |applyN f (x,y) :== e| & \text{means} & |f :== lam (tup (x,y)) e|  \\
     \end{array} \\ \\[-3mm]

    \begin{array}{rcl}
      |apply1 head xsl| & |:==|  & |apply1 xsl 0| \\
      |apply1 tail xsl| & |:==|  & |all (def i (i > 0; apply1 xsl i))| \\
      |applyN cons (x,xsl)| & |:==|  & |all (x `choice` def i (apply1 xsl i))| \\
      |applyN append (xsl, ysl)| & |:==|  & |all ((def i (apply1 xsl i)) `choice` (def i (apply1 ysl i)))| \\
      |applyN flatMap (f,xsl)| & |:==| & |all (def i (apply1 f (apply1 xsl i)))| \\
      |applyN map (f, xsl)| & |:==|  & |if x :== apply1 head xsl then applyN cons (apply1 f x, applyN map (f,apply1 tail xsl)) else emptytup| \\
      |applyN filter (p, xsl)| & |:==| & |all (def i (x :== apply1 xsl i; one (apply1 p x); x))| \\
      |applyN find (p, xsl)| & |:==| & |one (def i (x :== apply1 xsl i; one (apply1 p x); x))| \\
%%      |applyN every (p, xsl)| & |:==| & |applyN map (p,xsl)| \\
      |applyN some (p, xsl)| & |:==| & |one (def i (apply1 p (apply1 xsl i)))|\\
%      |applyN fromTo (lo,hi)| & |:==| & |if lo > hi then fail else (lo `choice` applyN fromTo (lo+1,hi))|
      |applyN zip (xsl,ysl)| & |:==| & |all (def i (tup (apply1 xsl i, apply1 ysl i)))| \\
    \end{array}
  \end{array}
  $$
\caption{Functions on tuples, analogous to list or array functions in some other languages} \label{fig:list-functions}
\Description{Functions on tuples, analogous to list or array functions in some other languages}
\end{figure}

%and |oneKW| is like |head| for choices.
% With that in mind, here is how to understand
% the definitions in \cref{fig:list-functions}:
% \begin{itemize}
% \item |head|.
% To obtain the first element of a tuple, we can simply use array indexing. This fails if the tuple is empty.
%
% \item |tail|.
% To get the tail of a tuple, we apply the tuple to a fresh variable, and restrict that variable with |i > 0|.
% This way, the expression fails at the first value and only the other values are collected by |allKW|.
% It returns the empty tuple if |xs| is empty or not even a tuple.
%
% \item |cons| and |append|.
% We can implement |applyN cons (x,xs)| by collecting with |allKW| the choices we get from choosing |x|
% or all the elements of |xs|, and similarly we can define |append|.
%
% \item |flatMap|.
% For the function |map|, which applies a function to each element of a tuple and collects the results,
% two possible implementation come to mind, and they differ when the function passed to |map| fails or produces more than a single value.
% |flatMap| flattens the choices from iterating through |xs| with
% the choices taken by |f|, so |applyN flatMap ( (lam x ((x choice x+1))), tup (10,20)))| $\equiv$ |tup (10,11,20,21)|.
%
% \item |map|.
% But often this is not what we want: we may want the effects of |f|, such as failure or returning multiple values,
% to affect the whole expression and obtain the cartesian product of these choices.
% The expression |applyN map ((lam x ((x `choice` x+1))), tup (10,20))| should reduce to |tup (10,20)`choice` tup(10,21) `choice` tup (11,20) `choice` tup (11,21)|.
% We get that behavior if we simply define |map| much like in any other functional programming language:
% if taking the head of the tuple fails, we return an empty tuple.
% Else we apply |f| to the head, recur on the |tail| and assemble the result with |cons|.\label{sec:map}
%
% \item |filter|.
% How can we filter a tuple with a predicate |p|?
% We turn the tuple into the sequence of choices, and let those fail where |p| fails before collecting the other values into a tuple.
%
% \item |find|.
% If we use |oneKW| instead of |allKW|, we get the function that checks if |p| holds for some element of the tuple,
% and returns the first matching element, or fails if |p| does not hold for any element.
%
% \item |every|.
% To test if a predicate |p| holds for every element of a tuple |xs|, we can simply use |applyN map (p,xs)|!
% If |p| fails for any value, the whole call to |applyN map (p,xs)| fails.
% \end{itemize}

Pattern matching for function definitions is simply done by unification of ordinary expressions;
see the desugaring of pattern-matching lambda in \cref{fig:syntax}.
This in turn means that we can use ordinary abstraction mechanisms for patterns.
For example, here is a function, |fcn|, that could be called as follows: |applyN fcn (88,1,99,2)|.
$$
\begin{array}{l}
|apply1 fcn t :== def (x ^^ y) (t == tup(x,1,y,2); x+y)|
\end{array}
$$
If we want to give a name to the pattern, it is simple to do so:
$$
|applyN pat (v,w) :== tup(v,1,w,2);| \quad \quad
|apply1 fcn t :== def (x ^^ y) (t == applyN pat (x,y); x+y)|
$$
Patterns are truly first class, going well beyond what can be done with, say, pattern synonyms
in Haskell.  For example, |pat| could be \emph{computed}, like this:
$$
|applyN pat (a,v,w) :== if a==0 then tup (v,1,w,2) else tup (1,1,w,v)|
$$
so that the pattern depends on the value of |a|.

\subsection{\texorpdfstring{|for|}{``for''} loops} \label{sec:for}

The expression |fordo e_1 e_2| will evaluate |e_2| for each of the choices in |e_1|,
rather like a list comprehension in languages like Haskell or Python.
% The |e_1| is called the \emph{domain} of the loop.
The scoping is peculiar\footnote{But similar to C++, Fortress, and Swift, and explained by
  the subsequent desugaring into |allKW|.}
in that variables bound in |e_1| also scope over |e_2|.
So, for example, |fordo(x :== 2 `choice` 3 `choice` 5) (x+1)| will reduce to the tuple |tup (3,4,6)|.
% This example can, with the |:==| shorthand, be written |for^(x :== 2 `choice` 3 `choice` 5) do (x+1)|.

Like list comprehension, |for| supports filtering; in \versecalc{}, this falls out naturally by just using
a possibly failing expression in |e_1|.
So, |fordo(x :== 2 `choice` 3 `choice` 5; x > 2) (x+1)| reduces to |tup (4,6)|.
Nested iteration in a |for| works as expected and requires nothing special.
So, |fordo(def (x^^y) (x==10 `choice` 20; y == 1 `choice` 2 `choice` 3)) (x+y)| reduces to |tup (11, 12, 13, 21, 22, 23)|.

Just as |if| is defined in terms of the primitive |oneKW| (\cref{sec:if}), we can define |for|
in terms of the primitive |allKW|.
Again, we have to be careful when |e_2| itself fails or produces multiple results;
simply writing |all (e_1;e_2)| would give the wrong semantics. So we put |e_2| under a lambda, and
apply each element of the tuple to |emptytup| afterwards, using a |map| function (as defined in \cref{fig:list-functions}):
\[
|fordo(def (x_1^^xdots^^x_n) e_1) e_2| \quad \text{means} \quad
|v :== all (def (x_1^^xdots^^x_n) e_1; thunk e_2); applyN map (lam z (apply0 z), v))|
\]
for a fresh variable |v|.
Note how this achieves that peculiar scoping rule: the initial variables in |def (x_1^^xdots^^x_n) e_1| are in scope in |e_2|.
Moreover, any effects (like being multi-valued) in |e_2| will not affect the choices defined by |e_1| since they are inside that lambda.
So, for example, |fordo(x:==10 `choice` 20) (x `choice` x+1)| will reduce to
|tup (10, 20) `choice` tup (10, 21) `choice` tup (11, 20) `choice` tup (11, 21)|.
At this point, it is crucial that the desugaring of |for| uses |map|, not |flatMap|.

Given that tuple indexing expands into choices, we can iterate over tuple indices and elements using |for|.
For example, |fordo(def (i ^^ x) (x == apply1 t i)) (x+i)| produces a tuple with the elements of |t|, increased by their index in |t|.  Notice the absence of the
fencepost-error-prone iteration of |i| over |(0..size(t)-1)|, common in most languages.


% \subsection{Enumerations}
% Making a choice between a sequence of numbers can be expressed like this
% $$
% \begin{array}{l@@{\;}c@@{\;}l}
%   |applyN fromTo (lo,hi)| & :== & |if lo > hi then fail else (lo `choice` applyN fromTo (lo+1,hi))|
% \end{array}
% $$
% so |all (applyN fromTo (1,5))| will reduce to the tuple |tup (1,2,3,4,5)|.

%% \subsection{Handling out-of-bounds errors}
%% Indexing out-of-bounds of a tuple results in |fail|.
%% This means that we cannot avoid handling this case.
%% $$
%% |if x:==apply1 t 5 then apply1 print x else apply1 print "out of bounds"|
%% $$

%% \subsection{Subset and powerset}
%% The subsets (subtuples) of a tuple can be expressed with this multi-valued function.
%% First compute the subsets of all but the first element, and then return these with and without the first element added.
%% $$
%% |subset^(t) :== if x:==apply1 head t then s:== apply1 subset (apply1 tail t); applyN cons (x, s) `choice` s else emptytup|
%% $$
%% And computing the powerset is then simply a matter of collecting all the subsets into a tuple.
%% $$
%% |powerset^(t) :== for x :== apply1 subset t in x|
%% $$

\section{Rewrite rules} \label{sec:rewrite-rules}

How can we give a precise semantics to a programming language?
Here are some possibilities:
\begin{itemize}
\item \emph{A denotational semantics} is the classical approach,
  but it is tricky to give a (perspicuous) denotational semantics to a functional logic language
  because of the logical variables.  We have such a denotational semantics under development,
  which we offer for completeness in \cref{sec:den-sem}, but that is the subject of another paper.
\item \emph{A big-step operational semantics} typically involves explaining how a (heap, expression) starting point
  evaluates to a (heap, value) pair; Launchbury's natural semantics for lazy evaluation~\cite{launchbury:natural} is the classic paper.  The heap, threaded through the semantics, accounts for updating thunks as they are evaluated.
  Despite its ``operational semantics'' title, the big-step approach
  does not convey accurate operational intuition, because it goes all the way to a value
  in one step.
\item \emph{A small-step operational semantics} addresses this criticism:
  it typically describe how a (heap, expression, stack) configuration evolves, one small
  step at a time ({\eg}, ~\cite{pj:stg-machine}).
  The difficulty is that the description is now so low level that it is again hard to explain to programmers.
\item \emph{A rewrite semantics} steers between these two extremes.
  For example, Ariola {\etal}'s ``A call-by-need lambda calculus''~\cite{ariola:call-by-need} shows how
  to give the semantics of a call-by-need language as a set of rewrite rules. The great
  advantage of this approach is that it is easily explicable to programmers.
  In fact, teachers almost always explain the execution of Haskell or ML programs as a succession of rewrites
  of the program, such as: inline this call, simplify this |case| expression, \etc
\end{itemize}
Up to this point, there has been no satisfying rewrite semantics for functional logic languages
(see \cref{sec:related} for previous work).
Our main technical contribution is to fill this gap with a rewrite semantics for \versecalc{}, one that
has the following properties:
\begin{itemize}
\item The semantics is expressed as a set of rewrite rules (\cref{fig:rules}). These rules apply to the core language of \cref{fig:syntax}, after
  all desugaring.
\item Any rule can be applied, in either direction,
  anywhere in the program term (including under lambdas).
\item The rules are (mostly) oriented, with the intent that using them left to right makes progress.
\item Despite this orientation, the rules do not say which rule should be
  applied where; that is the task of a separate \emph{evaluation strategy} (\cref{sec:eval-strategy}).
\item The rules can be applied by programmers to reason about what their program does,
  and by compilers to transform (and hopefully optimise) the program.
\item There is no ``magical rewriting'' (\cref{sec:magical}):
  all the free variables on the
  right-hand side of a rule are bound on the left.
\end{itemize}

\figureRewriteRules
\figureContexts
%% \figureStructuralRules

% \subsection{Ordering terms}
% The \rulename{var-swap} orders expressions using the ordering |lte|.
% The reason for this is to simplify the confluence proof by having a more canonical representation.
% The ordering is based on using de~Bruijn indices as the variable representation;
%%The ordering is defined for expressions |eq|, \ie, expressions and equalities.
%%The |lte| is defined so that all equalities, |v==e|, are less than all expressions, |e|.
% |e_1 lte e_2| if the natural lexicographical comparison of the terms with de~Bruijn indices is |e_1 < e_2|.
% \Eg, in the expressions |def (y x) (y == x; 0)| the |x| is considered less than the |y| since
% the de~Bruijn index for |x| is 0, and for |y| it is |1|.
% So in this case \rulename{var-swap} will turn it into |def (y x) (x == y; 0)|.
%
% Note that the \rulename{seq-swap} rule does not reorder choices;
% the rule only applies if at least one of the |eq| has no choice.


\subsection{Functions and function application rules} \label{sec:beta}

Looking at \cref{fig:rules}, rule \rulename{app-add}
should be familiar: it simply rewrites an application of |add| to integer
constants.  For example $|applyN add (3,4)| \movesto |7|$.  Rules \rulename{app-gt} and
\rulename{app-gt-fail} are more interesting: |applyN gt (k_1,k_2)| fails if $k_1 \leq k_2$ (rather than returning |False| as is more conventional), and returns |k_1| otherwise (rather than returning |True|).
An amusing consequence is that |(10 > x > 0)| succeeds
iff |x| is between |10| and |0| (comparison is right-associative).

% Rule \rulename{p-map} implements a map operation over tuples: It applies the function |f| to each value
% in turn, and collects the results into a tuple. If the function applications produce a choice
% of values, this will, as it should, produce a choice of tuples:
% \[
% |applyN map ((lam x (x choice  x+1)), (10,20))| \movesto^\ast |(10,20) choice (10,21) choice (11,20) choice (11,21)|
% \]
\textbeta-reduction is performed quite conventionally by \rulename{app-beta}; the
only unusual feature is that on the RHS of the rule, we use an |defKW| to bind |x|,
together with |(x==v)| to equate |x| to the argument.
The rule may appear to use call-by-value, because the argument is a value |v|,
but remember that values include variables, which may be bound to an
as-yet-unevaluated expression.  For example:
$$
|def y (y == 3+4; apply1 (lam x (x+1)) y)|
\; \movesto \;
|def y (y == 3+4; def x (x==y; x+1))|
$$
Finally, the side condition $x \not\in \freevars{v}$ in \rulename{app-beta}
ensures that the $|defKW| |x|$
% hacky typesetting to ge the \exists right next to the x.
does not capture any variables free in |v|.  If |x| appears free in |v|,
\textalpha-conversion may be used on |lam x e| to rename |x| to $y \not\in \freevars{v}$.

In \versecalc{}, tuples behave like (finite) functions in which application is
indexing.  Rule \rulename{app-tup} describes how tuple application works on non-empty
tuples, while \rulename{app-tup-0} deals with empty tuples.
Notice that \rulename{app-tup} does not require the argument to be evaluated to an integer $k$; instead
the rule works by narrowing.  So the expression |def x (apply1 (tup(2,3,2,7,9)) x == 2; x)|
does not suspend awaiting a value for |x|; instead it explores all the
alternatives, returning |(0 choice 2)|.  This is a free design decision:
a suspending semantics would be equally easy to express.

\subsection{Unification rules} \label{sec:unification-rules}

Next we study unification, again in \cref{fig:rules}.
Rules \rulename{u-lit} and \rulename{u-tup}
are the standard rules for unification, going back nearly 60 years \cite{robinson:unification}.
% Note that
% when unification succeeds it yields the common value; hence |s == s| rewrites
% to |s|\footnote{An alternative choice would
% for unification to yield |emptytup| on success.  It does not make much difference
% either way.}.
Rule \rulename{u-fail} makes unification fail on two different head-normal forms (see
\cref{fig:syntax} for the syntax of |hnf|).
Note in particular that unification fails if you attempt to unify a
lambda with any other value, including itself (see \cref{sec:unifying-lambda}).

The standard ``occurs check'' is rule \rulename{u-occurs}, which makes use of a
\emph{context} |V|, whose syntax is given in \cref{fig:contexts}
\cite{felleisen:contexts1,felleisen:contexts2}.
In general, a context is a syntax tree containing a single hole, written |hole|.
The notation |V^[v]| is the term obtained by filling the hole
in |V| with |v|.  For example, \rulename{u-occurs} reduces |x == tup(1,x,3)| to |fail|
using the context |V| = |tup( 1, hole , 3)|.


% Failure, written |fail|, propagates upwards, as \rulename{fail-def} and \rulename{fail} show.
% The latter rule makes use of an \emph{execution context}, |X|,
% whose syntax is given in \cref{fig:contexts}.
% In general, a ``context'' is a syntactic construct (in this case an expression)
% containing a single hole, written |hole| \cite{felleisen:contexts1,felleisen:contexts2}.
% The notation |X^[e]| is the expression obtained by filling the hole
% in |X| with |e|.
% For example |(y == tup (hole, 4))| is a possible execution context, |X|; filling its hole with 99
% gives the expression $|X^[99]| \equiv |(y == tup (99,4))|$.
% So \rulename{fail} says that if |fail| occurs in an |X|-context, thus |X^[fail]|, we can rewrite
% it to |fail|, discarding |X| altogether. For example |(x==y; fail; 3)| rewrites to |fail|.

% \subsection{Unification: the \rulename{subst} rule}

The key innovation in \versecalc{} is the way bindings (that is, just ordinary equalities)
of logical variables are propagated.
The key rule is:
$$
\begin{array}{rrcll}
  \rulename{subst} & |X^[x==v;e]| & \movesto & |(subst X x v)^[x==v; subst e x v]| \hspace*{-0.3em}
    & \text{if $|x| \in \freevars{|X,e|}$, $|x| \notin \freevars{v}$} \\
  &&&& \text{and $|v| = |y| \implies |x| \prec |y|$} \\
\end{array}
$$
The rule says that if we have an equation |(x==v)|,
we can replace the occurrences of |x| by |v| in its surrounding context |X|.
This rule makes use of new context |X| (see \cref{fig:contexts}), and
uses the notation |subst X x v| to mean ``substitute |v| for |x| in |X|'' and
similarly |subst e x v|.
There are several things to notice:
\begin{itemize}
\item \rulename{subst} fires only when the right-hand side of the equation
  is a \emph{value} |v|, so that the substitution does not risk
  duplicating either work or choices. This
restriction is precisely the same as the \rulename{let-v} rule of \citet{ariola:call-by-need}
and, by not duplicating choices, it neatly implements so-called \emph{call-time choice}~\cite{hanus:survey}.
We do not need a heap, or thunks, or updates; the equalities of the program elegantly
suffice to express the necessary sharing.
\item \rulename{subst} replaces all occurrences of |x| in |X| and |e|, but
  \emph{it leaves the original |(x==v)| undisturbed}, because |X| might not be
  big enough to encompass all occurrences of |x|.
  For example, we can rewrite |(y==x+1; (x==3; z==x+3))| to |(y==x+1; (x==3; z==3+3))|, using
  |X == (hole; z==x+3)|, but that leaves an occurrence of |x| in |(y==x+1)|.
  When there are no remaining occurrences of |x| we may eliminate
  the binding: see \cref{sec:gc}.
\item The side condition $x \not\in \freevars{v}$ in \rulename{subst} prevents infinite substitution,
  while $x \in \freevars{X,e}$ ensures that there is at least one occurrence to substitute.
  The other side condition will be explained next, when we discuss \rulename{var-swap}.
\end{itemize}

\subsection{Swapping and binding order} \label{sec:swapping}

Rules \rulename{hnf-swap} helps \rulename{subst} to fire by putting the variable on the left.
Rule \rulename{var-swap} is trickier.  Consider this example where |a| and |b| are
bound further out, perhaps by lambdas.  It can rewrite in two different ways:
$$
\begin{array}{l@@{\,}l@@{\hspace{1cm}}l@@{\,}l}
  \multicolumn{4}{c}{|def x (x == tup a ; x == tup b; x)|} \\
  \movestoR{subst} & |def x (x == tup a ; tup a == tup b; tup a)| &
     \movesto & |def x (tup b == tup a ; x == tup b; tup b)| \\
  \movestoR{u-tup, eqn-elim} & |a == b; tup a| &
     \movesto & |b == a; tup b| \\
  \movestoR{subst} & |a == b; tup b| &
      \movesto & |b == a; tup a| \\

\end{array}
$$
Each column is a reduction sequence starting from the same common term at the top;
the two sequences differ when it comes to which equation for |x| is chosen for \rulename{subst}
in the first step. As you can see, they conclude with two terms that are ``obviously''
the same, but which are syntactically different.  Rule \rulename{var-swap} allows
them to be brought together, so that the unification rules are confluent.
Rule \rulename{seq-swap} is needed for a similar reason. Consider
this example:
% where |a,b,c| are bound in that order; \ie, $|c| \prec |b| \prec |a|$:
$$
\begin{array}{l@@{\,}l@@{\hspace{1cm}}l@@{\,}l}
  \multicolumn{4}{c}{|c==a; c==b; c|} \\
  \movestoR{subst} & |c==a; a==b; a| &
     \movestoR{subst} & |b==a; c==b; b| \\

  \movestoR{var-swap} & |c==a; b==a; a| &
     \movestoR{subst}   & |b==a; c==a; a| \\
\end{array}
$$
Again, the concluding terms of the two columns are ``obviously'' the same,
because they differ only in the order of the equations |(b==a)| and |(c==a)|;
\rulename{seq-swap} allows them to be brought together, and makes explicit
our intuition that order of equations |(x=v)| does not matter.
% Both \rulename{var-swap} and \rulename{seq-swap} are needed only for \emph{confluence},
% not for \emph{progress}. \simon{I think this is true of both, isn't it?}
% \lennart{Without \rulename{var-swap} I think we need another \rulename{eqn-elim} rule for the |y==x| case.}
% \lennart{We also don't need \rulename{exi-swap} if we a stronger \rulename{eqn-elim}.}

Next we study the mysterious $x \prec y$ side condition in \rulename{var-swap},
and similar ones in \rulename{subst} and \rulename{seq-swap}.
In the overall proof of confluence, it turns out to
be very helpful if the unification rules
are \emph{terminating} (see \cref{sec:design-confluence}).
To achieve this, \rulename{var-swap} fires on |y==x| \emph{only if
|x| is bound inside |y|}, written $|x| \prec |y|$, so that the innermost-bound variable
ends up on the left.
Similarly, the side condition on \rulename{seq-swap} prevents it firing infinitely;
and the side condition $(|v| = |y| \implies |x| \prec |y|)$ on \rulename{subst}
prevents the rule from firing until \rulename{var-swap} has done its work.

Other rules, notably \rulename{exi-swap}, may change this binding order
and thereby re-enable \rulename{var-swap} or \rulename{seq-swap},
but the unification rules \emph{considered in isolation} are terminating and confluent,
and that is what we need for the proof.

\subsection{Local substitution} \label{sec:local-subst}

Consider this (extremely) tricky term:
|def x (x = if (x==0; x>1) then 33 else 55)|.
What should this do? At first you might think it was stuck---how can we simplify the |if|
when its condition mentions |x| which is not yet defined?  But in fact, rule \rulename{subst} allows
us to substitute \emph{locally} in any X-context surrounding the equation |(x==0)|
thusly:\footnote{
  Here and elsewhere we rewrite terms that have not
  been fully desugared, but that is just an expository aid;
  formally, the rules apply only to programs in the language of \cref{fig:syntax}.}
$$
\begin{array}{l@@{\;}l}
\multicolumn{2}{l}{|def x (x = if (x==0; x>1) then 33 else 55; x)|} \\
\movestoR{subst}                & |def x (x = if (x==0; 0>1) then 33 else 55; x)| \\
\movestoR{app-gt-fail,fail-elim}     & |def x (x = if fail then 33 else 55; x)| \\
\multicolumn{2}{l}{\movestoC{\text{simplify |if|}} \; |def x (x = 55; x)| \quad
\movestoR{subst} \; |def x (x = 55; 55)| \quad
\movestoR{exi-elim}  \; |55|}
\end{array}
$$
Minor variants of the same example get stuck instead of reducing. For example, if you replace
the |(x==0)| with |(x==100)| then rewriting gets stuck, as the reader may verify; and yet there
is a solution to the equations, namely |x==55|.
And if you replace |(x==0)| with |(x==55)|, then rewriting again
gets stuck, and reasonably so, since in this case there are \emph{no} valid solutions to the equations.
Perhaps this is not surprising: we cannot reasonably expect the program to solve arbitrary equations.
For example, |def x (x*x = x)| has two solutions but discovering that involves solving a quadratic equation.

% It would be nicer if all three variants got stuck. Perhaps we can simplify the system, by only allowing
% substitution over the entire scope of the |defKW|, and combining it with \rulename{def-elim} thus:
% $$
% \rulename{global-subst} \quad |def x (X^[x==v])|  \movesto  |(subst X x v)^[v]|
%         \quad \text{if $|x| \in \freevars{X}$, $|x| \notin \freevars{v}$} \\
% $$
% Now all three variants get stuck.  But alas, we lose confluence.  Consider this term:
% $$
% |def x (x == if (def y (y==x; y==0; y>1)) then 33 else 55)|
% $$
% In the condition of the |if| we have introduced a local variable |y|, and set it equal to |x|.
% Suppose we use \rulename{global-subst} instead of \rulename{subst}.  Then we can rewrite this term in two
% different ways:
% \begin{itemize}
% \item Use \rulename{global-subst} on |(y==x)|; doing so recovers the original tricky term, which is
% stuck, because \rulename{global-subst} cannot rewrite it.
% \item Use \rulename{global-subst} on |(y==0)|; then we can simplify the |if|, and reduce the entire thing to |55|.
% \end{itemize}
% We have lost confluence: two different rewrites starting from the same term lead to non-joinable terms.
% This is what drives us to adopt the tricky local-substitution rule \rulename{subst} rather than
% the simpler \rulename{global-subst}: to establish confluence.

\subsection{Elimination and normalization rules} \label{sec:gc} \label{sec:normalization}

% Indeed, |e| can mention |x| too: consider |(x == ((y==x+1); 3); y)|. Using
% \rulename{eqn-float} gives us |(y==x+1; x==3; y)|; to make further progress we must
% move the |x==3| to the left, using \rulename{seq-assoc-swap2}, giving |(x==3; y==x+1; y)|.
% Now we can substitute for |x|, and then for |y|.

Four \emph{elimination} rules allow dead code to be dropped (\cref{fig:rules}):
\rulename{val-elim} discards a value to the left of a semicolon; \rulename{exi-elim} discards
a dead existential; \rulename{eqn-elim} discards an existential $|defKW| |x|$ whose only
occurrence is an equation |x=v|; and \rulename{fail-elim} discards the context surrounding a |fail|.
Note that none of these rules, except \rulename{fail-elim}, discard an unevaluated expression,
because that expression might fail and we don't want to ``lose'' that failure (c.f. \cref{sec:lenient}).
The exception is \rulename{fail-elim}, which propagates failure.

Four \emph{normalization} rules help to put the expression in a form that
allows other rules to fire (\cref{fig:rules}):  \rulename{exi-float} allows an existential to float outwards;
\rulename{seq-assoc} makes semicolon right-associated; \rulename{eqn-float} moves work out
of the right-hand side of an equation |v==e|.
For example, we cannot use \rulename{subst} to substitute for |x| in |(x == (e; 3); x+2)|,
because the RHS of the |x|-equation is not a value; but we can instead apply \rulename{eqn-float}
to get |(e; x==3); x+2|, and then \rulename{seq-assoc} to get |e; x==3; x+2|; and now we \emph{can}
apply \rulename{subst}.

Rule \rulename{exi-swap} allows you to move an existential inward so that a dead equation can be eliminated
by \rulename{eqn-elim}.
Rule \rulename{exi-swap} is unusual because it can be infinitely applied;
avoiding that eventuality is easily achieved by tweaking the evaluation strategy (\cref{sec:eval-strategy}).

Note that all these swapping and normalization rules \emph{preserve the left-to-right sequencing of expressions},
which matters because choices are made left to right as we saw in \cref{sec:mixing}.
Moreover, the rules do not float equalities or existentials out of choices:
that restriction is the key to localizing unification (\cref{sec:mixing}), and
the flexible/rigid distinction of
\cref{sec:rigid}. For example, consider the expression
|(y == ((x==3; x*2) choice (x==4)); tup (x+1,y))|.
We must not float the binding |(x==3)| to the expression |(x+1)|,
because the latter is outside the choice, and a different branch of the choice binds |x| to |4|.

% If the equality |(x==v)| turns out to be the sole remaining occurrence of |x| in
% the body of an |defKW|, then rule \rulename{def-eliml} can drop both the |defKW|
% and the unification equation, and similarly \rulename{def-elimr} if the equation is ordered the other way.
%
% A number of rules shuffle expressions about in order to make
% the main unification rules apply.
% The rule \rulename{swap} swaps the arguments to unification to put the variable on the left.
% Rule \rulename{def-float} moves an |defKW| ``out of the way.'' For example,
% the expression |(y == (def x (tup (x,2))); dots)| does not bind |y| to a value, but if we float the |defKW| out
% we get |def x (y==tup(x,2); dots)|, and now we may be able to make progress by substituting for |y|.
%
% Similarly, rules \rulename{unify-seql} and \rulename{unify-seqr} move sequencing out of unification,
% again to make it more likely to get the form |(x==v)|.  Notice that |e_1|, |e_2| and |e_3| stay in the
% same left-to-right order, there is no risk of re-ordering choices (c.f. \cref{sec:mixing}).
%
% Rules \rulename{unify-unifyl} and \rulename{unify-unifyr} move a unification between values out of another unification.
% Without these rules |(x==y)==1;y>0| would be stuck; but \rulename{unify-unifyl}
% rewrites it to |((x==y);(y==1);y>0|, thus exposing |y==1| for further use by \rulename{subst}. Furthermore, in order
% to prove confluence, we need to be able to freely move the equation |(v_1==v_2)|,
% as discussed in \cref{sec:conf-move-eqn}.
%% The \label has been commented out.
% These rules could be more general, see \cref{sec:variations-liberate-unify}.


\subsection{Rules for choice} \label{sec:choice-rules}

The rules for choice are given in \cref{fig:rules}:
\begin{itemize}
\item Rules \rulename{one-fail}, \rulename{one-value}, and \rulename{one-choice} describe
  the semantics of |oneKW|, as in \cref{sec:one}.
\item Similarly, \rulename{all-fail}, \rulename{all-value}, and \rulename{all-choice} describe
  the semantics of |allKW| (\cref{sec:all}).
\item Rules \rulename{choose-r} and \rulename{choose-l} eliminate |fail|, which behaves as
  an identity for choice.
\item Rule \rulename{choose-assoc} associates choice to the right, so that \rulename{one-choice}/\rulename{all-choice}
  can fire.  (The dots on the left of \rulename{all-choice} should be read as
  a string of right-associated choices.)
\end{itemize}
The most interesting rule is \rulename{choose}, which, just as described in \cref{sec:choice},
``floats the choice outwards,'' duplicating the surrounding context.
But what ``surrounding context'' precisely? We use two new contexts, |SX| and |CX|, both
defined in \cref{fig:syntax}. A \emph{choice context} |CX| is like an
execution context |X|, \emph{but with no possible
  choices to the left of the hole}:
$$
CX ::= \hole \vb |v == CX| \vb |CX ; e| \vb |ce; CX| \vb |def x CX|
$$
Here, |ce| is a guaranteed-choice-free expression (syntax in \cref{fig:choice-free-syntax}).
This syntactic condition is necessarily conservative; for example, a call |apply1 f x|
is considered not guaranteed-choice-free because it depends on what function |f| does.
We must guarantee not to have choices to the left so that we preserve order---see \cref{sec:mixing}.

The context |SX| (\cref{fig:contexts}) in \rulename{choose} ensures that |CX| is as large as possible.
This is a very subtle point:
without this restriction we lose confluence.  To see this, consider\footnote{Remember, |if| is syntactic sugar for
  a use of |oneKW|, see \cref{sec:one}, but using |if| makes the example easier to understand.}:
$$
\begin{array}{l@@{\;}l}
                                 & |def x ((if (x>0) then 55 else (44 `choice` 2)); x==1; (77 choice 99))|\\
 \movestoR{subst}                & |def x ((if (1>0) then 55 else (44 `choice` 2)); x==1; (77 choice 99))|\\
 \movestoC{\text{simplify |if|}} & |def x (55; x==1; (77 choice 99))|\\
 \movestoR{val-elim, eqn-elim}   & |77 choice 99|\\
\end{array}
$$
But suppose instead we floated the choice out, \emph{partway}, like this:
$$
\begin{array}{ll}
\multicolumn{2}{l}{|def x ((if (x>0) then 55 else (44 `choice` 2)); x==1; (77 choice 99))|}\\
\quad \movestoC{\text{Bogus \rulename{choose}}} & |def x ((if (x>0) then 55 else (44 `choice` 2));
                            ((x==1; 77) choice (x==1; 99)))|\\
\end{array}
$$
Now the |(x==1)| is inside the choice branches, so we cannot use \rulename{subst} to
substitute for |x| in the condition of the |if|.
Nor can we use \rulename{choose} again to float the choice further out
because the |if| is not guaranteed choice-free (in this case the |else| branch has a choice).
So, alas, we are stuck!
Our not-entirely-satisfying solution is to force \rulename{choose} to float the choice all the way
to the top.  The |SX| context (\cref{fig:contexts}) formalizes what we mean by ``the top'':
rule \rulename{choose} can float a choice outward only when it becomes part of
the choice tree (context |SC|) immediately under a |oneKW| or |allKW| construct (context |SX|).

Rule \rulename{choose} moves choices around; only \rulename{one-choice} and \rulename{all-choice}
\emph{decompose} choices.
So choice behaves a bit like a data constructor, or normal form, of the language.
This contrasts with other approaches that eliminate choice by non-deterministically
picking one branch or the other, which immediately gives up confluence.
% For this reason we call
% this approach \emph{spatial choice}, in contrast to approaches that eliminate choice by non-deterministically
% picking one branch or the other, which immediately gives up confluence.

% The rules for |oneKW| and |allKW| expect multiple choices to be
% normalized into a right-associative list of non-failing values, and
% the administrative rules \rulename{assoc-choice}, \rulename{fail-l}
% and \rulename{fail-r} bring nested choices into that form. But why do these rules
% need a |SX| context? Again, they are needed to guarantee confluence. Suppose we had a rule \rulename{fail-l-no-sx}
% that unconditionally rewrites |(fail choice e)| to |e|. Now consider these
% two reduction sequences, starting from the same expression:
% $$
% \begin{array}{rll}
% |apply0 f; (fail choice (3 == (1 choice 3)))| & \movestoR{fail-l-no-sx} & |applyN f (); 3 == (1 choice 3)| \\
% |apply0 f; (fail choice (3 == (1 choice 3)))| & \movestoR{choose} & |applyN f (); (fail  choice  ((3 == 1) choice (3 == 3)))| \\
%  & \movestoRn{ulit}{2}  & |applyN f (); (fail  choice  (fail choice 3))| \\
%  & \movestoRn{fail-l-no-sx}{2} & |applyN f ();3|
% \end{array}
% $$
% The first sequence gets stuck after one step\footnote{The strange |apply0 f| prevents
%   us using \rulename{choose} to float the |(1 choice 3)| upwards.}, while the second makes more progress;
% and the two results are not joinable.

% Consider this expression:
% $$
% |(if (x > 2) then e1 else e2); (x == 1 choice fail)|
% $$
% The |if| is stuck until we have a value for |x|; we cannot float the choice out with \rulename{choose} because
% |if| is not choice-free.  If \rulename{fail-r} lacked a |SX| context we could rewrite
%  $|(x==1 choice fail)| \movesto |(x==1)|$; and now we can propagate the |(x==1)| to the |if| via \rulename{subst}
% and make progress.  But to do this \emph{we must look at the right branch of the choice, before we know whether
%   the left branch will reach a value or fail}, and that requires the implementation to explore
% both branches in parallel, a bit like adding parallel-or to the language.  That is a heavy obligation, and one
% we do not need to incur.
% Adding the |SX| context to \rulename{fail-r} allows an implementation
% to be sequential---without, of course, prohibiting parallel evaluation.  The |SX| is less significant
% for \rulename{fail-l} and \rulename{assoc-choice}, but we add it anyway for uniformity.

\subsection{The Verse calculus is lenient} \label{sec:lenient}

\versecalc{} is \emph{lenient} \cite{lenience},
not lazy (call-by-need), nor strict (call-by-value).
Under lenient
evaluation, everything is eventually evaluated, but functions can run
before their arguments have a value.
Consider a function
call |apply1 f e|, where |e| is not a value.  In \versecalc{}, applications are in administrative normal form (ANF),
so we must actually write |def x (x==e; apply1 f x)|. This expression will not return a value until
|e| reduces to a value: that is, everything is eventually evaluated.  But even so, |apply1 f x| can proceed
to \textbeta-reduce (\cref{sec:beta}), assuming we know the definition of |f|.

Lenience supports \emph{abstraction}.  For example, we can replace an expression |(x==tup(y,3); y>7)| by
$$
|def f (f == (lam (tup (p,q)) (p==tup(q,3); q>7)); applyN f (x,y)|
$$
Here, we abstract over the free variables of the expression, and define a named function |f|.
Calling the function is just the same as writing the original expression.
This transformation would not be valid under call-by-value.

This is not just a way to get \emph{parallelism}, which was the original motivation for introducing
lenience in the data-flow language Id~\cite{lenience}; it affects
\emph{semantics}. Consider:
$$
|def (f ^^ x ^^ y) (f == (lam p (x==7;p)); y == (if (x>0) then 7 else 8); apply1 f y|
$$
Here, |y| does not get a value until |x| is known; but |x| does not get its
value (in this case |7|) until |f| is called.  Without lenience this program would be stuck.

However, moving to \emph{laziness} seems problematic.
For example, consider: |def x (x == apply0 wombat; 3)|.  In a lazy, language this expression would
yield |3|, but in \versecalc{}, everything is evaluated, and
the expression will not return a value until |apply0 wombat| converges.
There is a good reason for this choice: |apply0 wombat| might fail, and we should not return |3| until
we know there is no failure.  With laziness, we could easily lose confluence.

\subsection{Evaluation strategy} \label{sec:eval-strategy}

Any rewrite rule can apply anywhere in the term, at any time.  For example, in the term |(x==3+4; y==3*2; x+y)|
the rewrite rules do not say whether to rewrite |3+4| $\rightarrow$ |7| and then |3*2| $\rightarrow$ |6|,
or the other way around.
The rules do, however, require us to reduce |3+4| $\rightarrow$ |7| before substituting for |x| in |x+y| because
rule \rulename{subst} only fires when the RHS is a value.
The rewrite rules thereby express \emph{semantics}.

For example, in the lambda calculus, by changing the rewrite
rule \textbeta{} to \textbeta V, we change the language from call-by-name to call-by-value; by
adding {\bf let}, plus suitable rewrite rules, we can express call-by-need \cite{ariola:call-by-need}.
In \versecalc{}, the rewrite rules are carefully crafted in a similar way; for example, \rulename{subst} will
only substitute |x==v| when the equation binds a variable to a \emph{value}, rather like \textbeta V in
lambda calculus.  Similarly, the elimination rules never discard a term that could fail.

In any term there may of course be many redexes---that is good.
An \emph{evaluation strategy} answers the question: given a closed term, which unique redex,
out of the many possible redexes, should be rewritten next to make progress toward the result?
Let us call an evaluation strategy \emph{good} if it guarantees to terminate if there
is \emph{any} terminating sequence of reductions; \ie, if any path terminates with a value,
then a good evaluation strategy will terminate with that same value\footnote{
It would be even better if the strategy could (c) guarantee to find the
result in the minimal number of rewrite steps---so called ``optimal reduction'' \cite{levy1978reductions,lamping1989algorithm,asperti1998optimal}---but optimal reduction is typically very hard, even in theory, and
invariably involves reducing under lambdas, so for practical purposes it is well out of reach.}.
For example, in the pure lambda calculus, \emph{normal-order reduction},
sometimes called \emph{leftmost-outermost reduction}, is a \emph{good} evaluation strategy.

We believe that the same leftmost-outermost strategy is close to being \emph{good} for
\versecalc{}\footnote{
  We say ``close to'' being good because we do not yet have a proof; indeed
  rule \rulename{fail-elim} may be a bit too powerful.
}: just repeatedly reduce the leftmost-outermost
redex, with some tweaks to avoid infinite application of \rulename{exi-swap}.
That is easy in theory, but it is tricky in practice.  For
example, consider |(x+y; tup(x,3)==tup(2,y));x|.  The |(x+y)| is not a
redex, but the equation is; we can apply unification to get |(x==2; y==3)|, and then
substitution to rewrite the |(x+y)| to |(2+3)|; and
\emph{now} the |(2+3)| is a redex.  So a reduction may ``unlock'' a
redex far to its left.  A major challenge of an implementation is to
find the next redex efficiently.

We have several prototype implementations of \versecalc{}, each involving an abstract
machine with a stack, a heap, a bunch of blocked computations, and so on. Exploring this
design space is, however, beyond the scope of this paper.

% The trouble is that it is hard to say what the ``leftmost'' redex is.  For example in |(e; tup(x,3)==tup(2,y))|,
% the equation may or may not be the leftmost redex, depending on whether |e| is stuck ({\ie}, contains no
% redexes); and whether or not |e| is stuck is a not syntactic property, and
% (worse) may depend not only on |e| itself, but on its context. Even worse, |e| may subsequently
% become un-stuck when we rewrite the equation.
% Any calculus in which a redex to the ``right'' may unblock one to the
% ``left''---that is, residuation---must grapple with this problem,
% so we leave evaluation strategy and compilation for future work.

% Formalising an evaluation strategy for \versecalc{} is beyond the scope of this paper, but
% we can make some informal comments.  First, in service of (b) we envisage compiling
% lambdas to code, and thus we never rewrite under a lambda \cite{pj-book}.
% Second, it never makes sense to evaluate in the right-hand argument of
% a choice\footnote{Except perhaps in parallel, of course.}, because
% \versecalc{}'s strong-ordering semantics mean that we must first find out what the left-hand argument
% is (especially, whether it fails) before the right-hand one can be used.
% So the basic plan is: rewrite the leftmost-outermost redex, subject to these two constraints.


\subsection{Developing and debugging rules}
The rules we describe here should both be able to transform a program to its value,
and also be confluent.
To aid in the development of the rules, we have used several mechanized tools to automate
reduction, random test-case generation, and confluence checking.
Initially, we used PLT Redex, \cite{pltredex}, which is very easy to use,
but not very efficient.
For better efficiency we switched to a Haskell library for term
rewriting.
The library provides a DSL for writing rules, and provides the infrastructure
to apply the rules everywhere, detect cycles, provide traces, \etc
Some sample rewrite rules can be found in \cref{fig:sample-rules}.

\begin{figure}
{\small
\vspace*{-0.5em}
\begin{code}
rules lhs  =   "APP-ADD"    `name`  (do  Op Add :@: Tup [Int k1, Int k2] <- [lhs]
                                         pure (Int (k1+k2)))

           <>  "EXI-SWAP"   `name`  (do  EXI x (EXI y e) <- [lhs]
                                         pure (EXI y (EXI x e)))

           <>  "EQN-ELIM"   `name`  (do  EXI x a <- [lhs]
                                         (ctx, (Var x' :=: Val v) :>: e) <- execX a
                                         guard (x == x' && x `notElem` free (ctx (v :>: e)))
                                         pure (ctx e))
\end{code}
\vspace*{-1.5em}
}
\caption{Sample Haskell reduction rules} \label{fig:sample-rules}
\Description{Sample Haskell reduction rules}
\end{figure}

We used this infrastructure in two ways.
First, we have a set of examples with known results, against which we can test a
potential rule set.
% We cannot prove formally that the rules evaluates all the programs we
% expect, since we have no specification of what we expect.
Second, before beginning a proof of confluence, we used QuickCheck \cite{quickcheck}
to generate test cases, and check them for confluence.
QuickCheck turned out to be invaluable at finding counterexamples to
otherwise reasonable looking rules;
it has run on the order of 100M tests on the current rule set.

\newcommand{\circlednum}[1]{\raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {#1}}}}
\newcommand{\xmovesto}{\hspace{-0.6em}\movesto\hspace{-0.6em}}


\section{Metatheory} \label{sec:metatheory}

The rules of our rewrite semantics can be applied anywhere,
in any order, and they give meaning to programs without
committing to a particular evaluation strategy.
%
But then it had better be the case that no matter how the
rules are applied, one always obtains the same result!
That is, our rules should be \emph{confluent}.  In this section, we describe
our proof of confluence.  Because the rule set is quite big (compared, say, to the
pure lambda calculus), this proof turns out to be a substantial undertaking.

\mypara{Reductions and Confluence}
%
A \emph{binary relation} is a set of pairs of related items.
%
A \emph{reduction relation} \rr is the \emph{compatible closure}\footnote{
``Compatible closure'' means that, for any context $E$ and any two terms $M$ and $N$,
if $(M,N)\in\rr$ then $(E[M],E[N])\in\rr$.}
of any binary relation on a set of tree-structured  \emph{terms},
such as the terms generated by some BNF grammar.
%
We write $\rr^{*}$ for the reflexive transitive closure of $\rr$.
% We write $\rr^{k}$ for the $k$-step closure of $\rr$ and $\rr^{*}$
% for the reflexive and transitive closure of $\rr$, \ie $\rr^{*} \equiv \cup_{0 \leq k} \rr^k$.
%\lennart{We don't see $\rr^k$ being used anywhere below, so we dropped it here.}\guy{And now dropping Lennart's comment.}
%
We write $\rstep{\rr}{e}{e'}$ ($a$ \emph{steps to} $b$) if $(e, e') \in \rr$ and
$\rsteps{\rr}{e}{e'}$ ($a$ \emph{reduces to} $b$) if $(e, e') \in \rr^{*}$.
%
A reduction relation $\rr$ is \emph{confluent} if whenever $\rsteps{\rr}{e}{e_1}$ and $\rsteps{\rr}{e}{e_2}$,
there exists an $e'$ such that $\rsteps{\rr}{e_1}{e'}$ and $\rsteps{\rr}{e_2}{e'}$.
%
Confluence gives us the assurance that we will not get
different results when choosing different rules, or get
stuck with some rules and not with others.

\mypara{Normal Forms and unicity}
%
A term $e$ is an $\rr$-\emph{Normal Form} if there does
not exist any $e'$ such that $\rstep{\rr}{e}{e'}$.
Confluence implies uniqueness of normal forms (unicity): if $\rsteps{\rr}{e}{e_1}$ and $\rsteps{\rr}{e}{e_2}$,
and $e_1$ and $e_2$ are normal forms, then $e_1 = e_2$
\cite[Corollary 3.1.13(ii)]{Barendregt}.

% \begin{lemma}[Unicity] \label{lem:nf}
% If $\rr$ is confluent then every term reduces to at most one normal form.
% \end{lemma}

% \begin{definition}[Recursive Equations] \label{def:norec}


\subsection{Recursion, and the notorious even/odd problem}  \label{sec:even-odd}
%
It is well known that adding {\bf letrec} to the lambda calculus
makes it non-confluent, in a very tiresome, but hard-to-avoid, way \cite{ariola:skew-confluence}.
In our context, consider the term:
%
$$
\begin{array}{lcll}
 |def (x ^^ y) (x == tup (1, y)); y == lam z x; x)|
 & \rsteps{}{}{} & |def y (y == lam z (tup (1, y)); tup (1,y))| &  \text {(1: substitute for |x| first)} \\[0.5mm]

 |def (x ^^ y) (x == tup (1, y)); y == lam z x; x)|
& \rsteps{}{}{} & |def x (x = tup (1, lam z x); x)| & \text {(2: substitute for |y| first)}
\end{array}
$$
%
The result of (1) and (2) have the same meaning (are indistinguishable by a \versecalc{} context)
but cannot be joined by our rewrite rules. Nor is this easily fixed by adding new
rules, as we did when we added \rulename{var-swap} (\cref{sec:unification-rules})
and \rulename{seq-swap} (\cref{sec:normalization}).  Why not? Because the terms
are equivalent only under some kind of graph isomorphism.

We have tackled this problem in three different ways.  First, we can simply prohibit recursion, and prove
confluence under that restriction (\cref{sec:confluence-sketch}).
This is akin to proving confluence for the lambda calculus with {\bf let} but not {\bf letrec}.
In \versecalc{}, excluding recursion is not so simple because \versecalc{} has no {\bf letrec}; rather,
recursion emerges during execution.  For example, is this recursive: |def (x ^^ y) (x==tup (1,y); applyN f (x,y))|?
It might be if |f == (lam (tup (v, w)) (v==w))|!
For \emph{tuples}, we have a simple sol\-ution: rule \rulename{u-occurs} makes the entire
term fail if we get recursion through a tuple. But we cannot do this for lambdas because it
leads to non-confluence. Consider |f == (lam x (applyN const (x, f)))|, where |const == (lam (tup (p,q)) p)|.
The equation for |f| \emph{looks} recursive because the RHS mentions |f|; but if we \textbeta-reduce the application
of |const|, the occurrence of |f| disappears.

Thus motivated, we restrict our attention to terms that have \emph{\norecursion}:
\begin{itemize}
\item A term $e$ has \emph{\norecursion} if it is not transitively recursive.
\item A term $e$ is \emph{transitively recursive} if $\rsteps{}{e}{e'}$ where $e'$ is recursive.
\item A term $e$ is \emph{recursive} if it contains a recursive equation.
\item A \emph{recursive equation} is an equation of the form
|x == V^[lam y e]|, where $x \in \freevars{e}$,
which equates a variable |x| with a value that contains a lambda in which |x| is free.
\end{itemize}
%

\noindent
The \norecursion condition is not as severe as it might first appear: it only prohibits
recursion through \emph{equations}.
But no expressiveness is lost thereby: in our untyped setting, one can still write recursive (and non-terminating) programs
using one's favorite fixpoint combinator, such as \textbf{Y} or \textbf{Z}.

This approach is not entirely satisfying: it is hard to prove that a term
has no recursion, and it is clumsy to write recursive programs using only \textbf{Y}-combinators.
Our second approach is to adopt the idea of \emph{skew confluence}
\cite{ariola:skew-confluence}, a clever technique developed
specifically to handle the even/odd problem; we give an overview of skew confluence
in \cref{sec:skew-overview-2}, and
provide details of our approach to a proof of skew confluence for \versecalc{} in \cref{sec:skew-confluence},
including several new lemmas, but we emphasize that the proof of skew confluence is not yet complete.

A third approach is simply to abandon confluence as a goal altogether.
Confluence is, after all, purely \emph{syntactic}, and hence
much stronger than what we really need, which is that each of our rules be
\emph{semantics}-preserving.  But, of course, that requires
an independent notion of semantics, a direction we sketch in \cref{sec:den-sem}.

\subsection{Proof of confluence} \label{sec:confluence-sketch}

Our main result is that \versecalc{}'s reduction rules are confluent for terms with \norecursion.
We sketch the proof here, with full details in \cref{sec:confluence} (and relevant preliminaries in \cref{sec:preliminaries}).


\begin{theorem}[Confluence] \label{thm:confluence-sketch}
%
The reduction relation in \cref{fig:rules}
is confluent for terms with \norecursion.
%
%  If $e$ has \norecursion and $\rsteps{\rr}{e}{e_1}$ and $\rsteps{\rr}{e}{e_2}$ then $\join{\rr}{e_1}{e_2}$.
%
\end{theorem}

\noindent
\emph{Proof sketch.}  Our proof strategy is to
%
(1)~divide the rules into groups for application, unification,
    \etc, approximately as in \cref{fig:rules},
(2)~prove confluence for each separately, and then
(3)~prove that their combination is confluent via commutativity.
%
Given two reduction relations $R$ and $S$, we say that $R$ \emph{commutes} with $S$ if for all terms
$e, e_1, e_2$ such that $\rsteps{R}{e}{e_1}$ and $\rsteps{S}{e}{e_2}$
there exists $e'$ such that $\rsteps{S}{e_1}{e'}$ and $\rsteps{R}{e_2}{e'}$.
%
We prove each individual sub-relation is confluent
and that they pairwise commute. Then confluence of
their union follows, using \citet{Huet80}:

\begin{lemma}[Commutativity] \label{lem:comm-summary}
If $R$ and $S$ are confluent and commute, then $R \cup S$ is confluent.
\end{lemma}
\noindent
%
Proving confluence for application, elimination and choice is easy:
%
% \simon{added normalization to this list -- Ranjit check} \ranjit{ACK: please check updated}
%
they all satisfy the \emph{diamond property}---namely, that two
different reduction steps can be joined at a common term
\emph{by a single step}---which suffices to show the
relations are confluent \cite{Barendregt}.
%
The diamond property itself can be verified easily by considering
critical pairs of transitions.
%
The rules for unification and normalization pose two problems.
%

\mypara{The Unification Problem}
%
The first problem is that the unification relation does not satisfy the diamond property
as it may need multiple steps to join the results of two different one-step reductions.
%
For example, consider the term $|(x == arr(1,y); x == arr(z,2); x == arr(1,2); 3)|$.
%
The term can be reduced in one step by substituting |x|
in the final equation by either |arr(1,y)| or |arr(z,2)|.
After this, it will take multiple steps to join the two terms.

Following a well-trodden path in proofs of confluence
for the \textlambda-calculus (\eg, \cite{Barendregt}),
our proof of confluence for the unification rules works in two stages.
%
First, we prove that the reductions are \emph{locally confluent},
meaning if $e$ single-steps to $e_1$ and $e_2$ then $e_1$
and $e_2$ can be joined at some $e'$ in \emph{multiple}
unification rule steps.
%
Second, we prove that the unification reductions are \emph{terminating}, which relies
upon eliminating recursion in tuples via \rulename{u-occurs} and in lambdas
via the \norecursion condition.
%
Newman's Lemma \cite[Lemma 2.4]{Huet80} then implies the
locally confluent, terminating unification relation is
also confluent.

\mypara{The Normalization Problem}
%
The second problem is that the normalization rules do not commute with the unification rules.
%
Recall from \cref{sec:swapping} that the unification rules rely upon
variable ordering to orient equations between variables in a canonical fashion.
%
The normalization rule \rulename{exi-swap} can \emph{change}
the variable order and hence, its behavior is deeply intertwined
with unification and cannot be factored out via a commutativity argument.
%
Instead, we prove that the union of unification and normalization
is confluent by showing that unification \emph{postpones after}
normalization \cite{Hindley}; see \cref{sec:confluence} for the gory details.
%
% , \ie, informally speaking, that a
% mixed sequence of unification-or-normalization steps can be reordered
% into two blocks of normalization-only steps followed by unification-only
% steps. The confluence of the union can then be established by suitably
% composing the confluence properties for the individual relations.

\subsection{Design for confluence} \label{sec:design-confluence}  \label{sec:unifying-lambda}

\versecalc{} is carefully designed to ensure confluence.


% Rule \rulename{deref-h} is particularly important.
% It prevents substituting heap values |h| under lambdas and inside tuples; and
% the |A| context only permits substitution in a place where the value |h| can be \emph{used immediately}, by
% application or unification.  These restrictions matter for at least three different reasons.

\mypara{Ensuring that unification terminates.}
%
Our proof strategy for the confluence of the unification rules requires that they terminate.
%
The side condition $|x| \not\in \freevars{v}$ in \rulename{subst} avoids infinite
substitution.
If instead we dropped that condition, the following sequence of \rulename{subst}
reductions would not terminate:
$$
|def x (x == tup (1,x); x)|
\; \rightarrow \; |def x (x == tup (1,x); tup (1,x))|
\; \rightarrow \; |def x (x == tup (1,x); tup (1,tup (1,x)))|
\; \rightarrow \; \ldots
$$
Here, each step makes one substitution for |x|.
An exactly analogous example can be made for a lambda value.

Similarly, as we discussed in \cref{sec:unification-rules},
rule \rulename{var-swap} uses the variable-ordering side condition $|x| \prec |y|$
to put the equation in a canonical orientation, and thus ensure that the unification rules terminate.

\mypara{Unifying lambdas.} In \versecalc{}, an attempt to unify two
lambdas fails even if the lambdas are semantically identical (rule
\rulename{u-fail}). Why? Because semantic identity of functions is
unimplementable.
%
We cannot instead say that the attempt to unify gets stuck because that leads to non-confluence.
Here is an expression that rewrites in two different ways, depending on which equation we \rulename{subst} first:
\[ |(lam p 1) == (lam q 2); 1| \lgos{} |def x (x == (lam p 1); x == (lam q 2); x emptytup)| \gos{} |(lam q 2) == (lam p 1); 2| \]
% $$
% \begin{array}{lll}
%  \multicolumn{3}{l}{|def x (x == (lam p 1); x == (lam q 2); x emptytup)|} \\
% &    \movesto \{\textit{use \rulename{SUBST} on first equality}\} & |(lam p 1) == (lam q 2); 1| \\
% OR & \movesto \{\textit{use \rulename{SUBST} on second equality}\} & |(lam q 2) == (lam p 1); 2|
% \end{array}
% $$
These two outcomes cannot be joined.  But making unification for lambdas fail, makes both outcomes
lead to |fail|, and confluence is restored.

\mypara{Unifying variables}
%
Note that while \rulename{u-lit} lets us eliminate equalities on the same literal $|k == k|$,
there is no analogous \rulename{u-var} rule to drop equalities on the same variable $|x == x|$.
Perhaps surprisingly, adding that rule would lead to non-confluence.
%
To see why, suppose we had such a \rulename{u-var}, and consider the term |(def x (x==(lam y y); x==x; 0))|.
%
If we first apply \rulename{u-var} to eliminate the equality $|x==x|$, then the remainder
reduces to |0|.
However, if we first \rulename{subst} the equality |x == (lam y y)|,
we get |((lam y y) == (lam y y); 0)| which |fail|s.
%
Thus, there is no \rulename{u-var}: such equalities can only be eliminated
after the value of |x| is substituted in and checked to not be a \textlambda.

% %
% \rulename{var-swap} has a curious side condition: it only rewrites
% $|x == y|$ to $|y == x|$ if $y \prec x$.
% %
% To understand why the \rulename{var-swap} is necessary and why it
% has this side condition, consider the expression $|lam a (lam b (def x (x == a; x == b; x)))|$.
% %
% Depending on which equality we \rulename{subst} first, we get either
% %
% $$\begin{array}{lcll}
% |lam a (lam b (def x (x == a; x == b; x)))|
%   & \movesto
%   & |lam a (lam b (def x (x == a; a == b; a)))|
%   &  \text {(1: substitute |x == a| first)} \\[0.5mm]
%   & \movesto
%   & |lam a (lam b (def x (b == a; x == b; b)))|
%   &  \text {(2: substitute |x == b| first)} \\[0.5mm]
% \end{array}$$
% %
% Now we are in a pickle: the two expressions are not syntactically the same!
% %
% Fortunately, \rulename{var-swap} rule lets us join the two expressions as follows.
% %
% Here, $b \prec a$ as the binder for $|b|$ is ``closer'', so we can
% join the two terms as
% %
% $$\text{(1)}\ |lam a (lam b (def x (x == a; a == b; a)))| \ \movesto\
%     |lam a (lam b (def x (x == a; b == a; a)))|
%     \ \movesfrom \ |lam a (lam b (def x (b == a; x == a; a)))|\ \text{(2)}$$
% %
% where from (1) we \rulename{var-swap} $|a == b| \movesto |b == a|$,
% and from (2) we \rulename{subst} $|b == a|$ and reorder the equations
% with \rulename{seq-swap}.
% %
% Thus, \rulename{var-swap} lets us pick a ``canonical representative''
% for all the variables that are equal to each other (here, we pick $|a|$
% to represent $|a|$, $|b|$ and $|x|$).
% %
% Furthermore, the ordering (based on binding-depth) ensures that the
% \rulename{var-swap} cannot apply forever, \ie\ that unification
 % terminates.


\subsection{Overview of skew confluence} \label{sec:skew-overview-2}

We travel a path very similar to the one blazed by Ariola and her co-authors.
Ariola and Klop studied a form of the lambda calculus with an added $\mathsf{letrec}$ construct and
determined (like us) that their calculus was not confluent; then they added a specific
constraint on recursive substitution
and proved that the modified calculus is confluent \cite{ariola:cyclic-lambda,ariola:lambda-letrec}.
In a later paper, Ariola and Blom proved that their calculus without the constraint, while not confluent,
does obey a weaker related property, which they invented, of \emph{skew confluence} \cite{ariola:skew-confluence}.
We believe, and currently are trying to prove, that \versecalc{} is skew confluent without the
pesky no-recursion side condition of \cref{thm:confluence-sketch}.
\begin{center}
  \begin{tabular}{rl}
    Confluence: & $\forall |e|, |e_1|, |e_2|.\; \rsteps{\rr}{|e|}{|e_1|} \wedge \rsteps{\rr}{|e|}{|e_2|} \implies \exists e'.\; \rsteps{\rr}{|e_1|}{|e'|} \wedge \rsteps{\rr}{|e_2|}{|e'|}$. \\
    Skew confluence: & $\forall |e|, |e_1|, |e_2|.\; \rsteps{\rr}{|e|}{|e_1|} \wedge \rsteps{\rr}{|e|}{|e_2|} \implies \exists e'.\; \rsteps{\rr}{|e_1|}{|e'|} \wedge |e_2| \preceqomegarr |e'|$. \\
  \end{tabular}
\end{center}
These are depicted here as two commutative diagrams, which differ only on the bottom edge:
%
\begin{center}
  \begin{tabular}{c@@{\qquad}c@@{\qquad\quad}c}
  Confluence & Skew confluence & \\
  % https://q.uiver.app/?q=WzAsNCxbMCwwLCJlIl0sWzEsMCwiZV8xIl0sWzAsMSwiZV8yIl0sWzEsMSwiZSciXSxbMCwxLCJcXHJyIiwyLHsic3R5bGUiOnsiaGVhZCI6eyJuYW1lIjoiZXBpIn19fV0sWzAsMiwiXFxyciIsMix7InN0eWxlIjp7ImhlYWQiOnsibmFtZSI6ImVwaSJ9fX1dLFsxLDMsIlxccnIiLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJkb3R0ZWQifSwiaGVhZCI6eyJuYW1lIjoiZXBpIn19fV0sWzIsMywiXFxyciIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImRvdHRlZCJ9LCJoZWFkIjp7Im5hbWUiOiJlcGkifX19XV0=
\begin{tikzcd}
	e & {e_1} \\
	{e_2} & {e'}
	\arrow["\rr"', two heads, from=1-1, to=1-2]
	\arrow["\rr"', two heads, from=1-1, to=2-1]
	\arrow["\rr", dotted, two heads, from=1-2, to=2-2]
	\arrow["\rr"', dotted, two heads, from=2-1, to=2-2]
      \end{tikzcd}
&
% https://q.uiver.app/?q=WzAsNCxbMCwwLCJlIl0sWzEsMCwiZV8xIl0sWzEsMSwiZSciXSxbMCwxLCJlXzIiXSxbMCwxLCJcXHJyIiwyLHsic3R5bGUiOnsiaGVhZCI6eyJuYW1lIjoiZXBpIn19fV0sWzEsMiwiXFxyciIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImRvdHRlZCJ9LCJoZWFkIjp7Im5hbWUiOiJlcGkifX19XSxbMCwzLCJcXHJyIiwyLHsic3R5bGUiOnsiaGVhZCI6eyJuYW1lIjoiZXBpIn19fV0sWzIsMywiXFxwcmVjZXFvbWVnYXJyIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiZG90dGVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XV0=
\begin{tikzcd}
	e & {e_1} \\
	{e_2} & {e'}
	\arrow["\rr"', two heads, from=1-1, to=1-2]
	\arrow["\rr", dotted, two heads, from=1-2, to=2-2]
	\arrow["\rr"', two heads, from=1-1, to=2-1]
	\arrow["\preceqomegarr", dotted, no head, from=2-2, to=2-1]
\end{tikzcd}
&
\raise0.5ex\hbox{\smash{\fbox{\parbox{18em}{For each diagram, given |e|, |e_1|, |e_2| that obey the relationships
               indicated by all the solid lines, there exists |e'| such that
               all relationships indicated by dotted lines are also satisfied.}}}}
\end{tabular}
\end{center}
%
% \simon{why dotted arrows?}
% The meaning of $|e_2| \inccon{\preceq}{\rr} |e'|$ is that |e_2| is \emph{convertible} to |e'|, but only through a sequence of reductions
% and converse reductions that maintain a certain partial order $\preceqomegarr$ (see below).
% \simon{don't understand this}
You can understand skew confluence as follows: if two different reduction paths from |e| produce
terms |e_1|, |e_2|, then |e_1| can be further reduced to |e'| so that all of |e_2|'s permanent structure
is present in |e'|, written $|e_2| \preceqomegarr |e'|$.  By ``permanent structure'' we mean
an outer shell of tuples, lambdas, and constants, that will never change no matter how much
further reduction takes place.  For example, however far we reduce the term |tup (1, lam z e)|, the
result will always look like |tup (1, lam z e')|, where $\rsteps{\rr}{e}{e'}$.  We can formalize the notion of
permanent structure by defining an
\emph{erasure function} $\omegarr(|e|)$ which replaces all the impermanent bits of |e|
with a new dummy term |Omega|.  Thus $\omegarr(|tup (1, lam z x)|) = |tup (1, lam z Omega)|$.
Then $|e_2| \preceqomegarr |e'|$ if $\omegarr(|e_2|)$ can be made equal to |e'| by
replacing each occurrence of |Omega| in $\omegarr(|e_2|)$ with an (individually-chosen) term.

Consider the even-odd problem discussed in \cref{sec:even-odd}.
$$
\begin{array}{l@@{\,}l@@{\hspace{1cm}}l@@{\,}l}
\multicolumn{4}{c}{|def (x ^^ y) (x == tup (1, y)); y == lam z x; x)|} \\
    \rsteps{}{}{} & |def y (y == lam z (tup (1, y)); tup (1,y))|
  & \rsteps{}{}{} & |def x (x = tup (1, lam z x); x)| \\
    \rstep{}{}{} & |def y (y == lam z (tup (1, y)); tup (1, lam z (tup (1,y))))|
  & \rstep{}{}{} & |def x (x = tup (1, lam z x); tup (1, lam z x)))| \\
    \rstep{}{}{} & |def y (y == lam z (tup (1, y)); tup (1, lam z (tup (1,lam z (tup (1,y))))))|
  & \rstep{}{}{} & |def x (x = tup (1, lam z x); tup (1, lam z (tup (1,lam z x))))| \\
    \rstep{}{}{} & \cdots & \rstep{}{}{} & \cdots
\end{array}
$$
The two columns can never join up, but if you pick any term, there is a term in the other
column that has a greater amount of permanent structure.  That in turn means that the terms in the left-hand
column are contextually equivalent to those in the right-hand column, because the context can
inspect only the permanent structure. This contextual equivalence is the real reason
for seeking confluence in the first place.

In \cref{sec:skew-confluence} we show how to adapt the proof strategy
of \cref{sec:confluence-sketch} and \cref{sec:confluence} for skew confluence.
To do this we need a new result: if two relations are skew confluent
with respect to the same information content function and commute, then their union is also skew confluent.
(In fact, it is not required that the two relations fully commute: a slightly weaker precondition suffices.)
Using this result, our plan
is to (i) define an appropriate information content function for \versecalc{} expressions;
(ii) prove that all the rewrite rules for \versecalc{} are monotonic in this information content function;
(iii) prove that the Unification rules (modified to permit recursive substitution) together with the Normalization rules are skew confluent;
(iv) prove that this combined set of rules commutes in the necessary way with the rules for Application, Elimination, and Choice (which taken together are already known to be confluent);
and (v) then apply our new result to show that the entire set of rewrite rules is skew confluent.
At this time steps (iii) and (iv) are incomplete, so we emphasize that we do not yet have a complete proof of skew confluence for \versecalc{}.


% ----------------------- Variations

\section{Variations and choices} \label{sec:variations}

%% \lennart{Mention binary {\vs} n-ary for choice and tuples?}

In a calculus like \versecalc{}, there is room for many design variations.
We discuss some of them here.


% JB: Removed this section now that fail-l and fail-r are wrapped into SX; this also fixes this problem I think.

% \subsection{Choice and failure} \label{sec:choice-and-failure}
%
% In \versecalc{} an expression delivers a \emph{sequence} of results, not a bag,
% and certainly not a set (\cref{sec:choice}).  That suggests that a good evaluation strategy might be to
% only explore the left branch of a choice first, and try the right branch only if former fails.  But consider
% $$
% |(if (x > 2) then e1 else e2); (x == 1 choice  fail)|
% $$
% The |if| is stuck until we have a value for |x|; we cannot float the choice out with \rulename{choose} because
% |if| is not choice-free.  With the rules in \cref{fig:rules2} we \emph{can} use \rulename{fail-r} to rewrite
% $|(x==1 choice fail)| \movesto |(x==1)|$; and now we can propagate the |(x==1)| to the |if| via \rulename{subst}
% and make progress.  But to do this we must look at the right branch of the choice, before we know whether
% the left branch will reach a value or fail.  There is nothing wrong with this, but it is arguably a bit peculiar that
% |(x==1 choice e)| can substitute for |x| if and only if |e| fails.  Even |(x==1) choice (x==1)| cannot substitute
% for |x|.
%
% An alternative would be to drop \rulename{fail-l} and \rulename{fail-r}, and drop the |fail| terms only in
% the rules for |oneKW| and |allKW|.  Both alternatives seem reasonable.


% \subsection{Administrative normal form}
%
% We have chosen to restrict tuple formation and function application to
% \emph{values} and not \emph{expressions}; thus we allow |apply1 v_1 v_2|, but not |apply1 e_1 e_2| unless
% |e_1| and |e_2| are values; and similarly |tup(v_1,v_2)| but not |tup(e_1,e_2)|.
% It would be entirely straightforward to allow expressions rather than values, but doing so
% would require many more rules, to describe how to distribute, {\eg}, application over sequencing.
% Restricting to values involves no loss of expressive power:
% it is simply a form of administrative normal form (ANF) \cite{felleisen:anf}.

% \subsection{Dead existentials}
%
% \simon{I don't think this section is persuasive; let's drop it.}
% \lennart{Drop it!}
%
% Consider the term |(def x 99)|. This rewrites to |99| by \rulename{def-elim}, but you could
% argue that it should instead be stuck.
% For example, the term
% |(def x (x==(1 choice 2); 99))| rewrites to |(99 choice 99)|, producing two results, \emph{one for each
%   solution for |x|}.  So, if |x| is entirely unconstrained, maybe we should return an infinite number of results?
% It would be easy to change this decision, by adjusting the rules in \cref{fig:wf} for well-formed results.

\subsection{Ordering and choices} \label{sec:ord-choice}

As we discussed in \cref{sec:choice-rules}, rule \rulename{choose} is less than satisfying for two reasons.
First, the |CX| context uses a conservative, syntactic analysis for choice-free expressions; and second,
the |SX| context is needed to force |CX| to be maximal.  A rule like this would be more satisfying:
$$
\rulename{simpler-choose} \quad   |CX^[ e_1 choice e_2 ]| \movesto |CX^[e_1] choice CX^[e_2]|
$$
The trouble with this is that it may change the order of the results (\cref{sec:mixing}).
Another possibility would be to accept that results may come out in the ``wrong'' order, but have
some kind of sorting mechanism to put them back into the ``right'' order.  Something like this:
$$
\rulename{labeled-choose} \quad   |CX^[ e_1 choice e_2 ]| \movesto |CX^[left e_1] choice CX^[right e_2]|
$$
Here, the two branches are labeled with |L| and |R|.  We can add new rules to reorder such
labeled expressions, something in the spirit of:
$$
\rulename{sort} \quad   |(right e_1) choice (left e_2)| \movesto |(left e_2) choice (right e_1)|
$$
We believe this can be made to work, and it would allow more programs to evaluate, but it adds unwelcome clutter to program terms, and the cure may be worse than the disease. However, the idea directly inspired our denotational
semantics (\cref{sec:order-sensitive}), where it seems to work rather beautifully.

\subsection{Generalizing \texorpdfstring{|oneKW|}{``one''} and \texorpdfstring{|allKW|}{``all''}} \label{sec:split}

In \versecalc{}, we introduced |oneKW| and |allKW| as the primitive choice-consuming operators, and neither is more general than the other, as discussed in \cref{sec:all}.
We could have introduced a more general operator |splitKW|\footnote{The name inspired by \citet{oleglogict}.} as $e ::=  |dots| \vb |split e v_1 v_2|$ and rules:
\[\begin{array}{lrcll}
    \rulename{split-fail} & |split fail f g| & \movesto & |apply0 f| \\
    \rulename{split-choice} & |split (v `choice` e) f g| & \movesto & |applyN g (v, thunk e)|  \\
    \rulename{split-value} & |split v f g| & \movesto & |applyN g (v, thunk fail)|  \\
\end{array}\]
The intuition behind |splitKW| is that it distinguishes a failing computation from one that returns at least one value. If |e| fails, it calls |f|, and if |e| returns at least one value, it passes that to |g| together with the remaining computation, safely tucked away behind a lambda.
When adding more effects to \versecalc{}, see \cref{sec:store}, it is in fact crucial to use $|splitKW|$ to exactly control the order of effects.

Indeed, this is more general, as we can implement |oneKW| and |allKW| with |splitKW|:
\[\begin{array}{r@@{\;}l@@{\;}l@@{\;}l}
    |one e| & \equiv |apply1 f x :== fail; | & |applyN g (x,y):== x;| & |split e f g| \\
    |all e| & \equiv |apply1 f x :== emptytup;|    & |applyN g (x,y):== applyN cons (x, split (apply0 y) f g);| & |split e f g|
\end{array}\]
For this paper, we stuck to the arguably simpler |oneKW| and |allKW|,
to avoid confusing the presentation with these higher-order encodings, but there are no complications using |splitKW| instead.

% \subsection{Laziness} \label{sec:laziness}
%
% As \cref{sec:lenient} discussed, \versecalc{} is lenient. Unlike Curry however, \versecalc{} is not \emph{lazy}.
% For example, consider: |def x (x == apply0 loop; 3)|.  In a lazy language this expression would
% yield |3|, but in \versecalc{} everything is evaluated, and the infinite computation |apply0 loop| will
% prevent the expression from returning a value.
% There a good reason for this choice: the call to |apply0 loop| might fail, and we should not return |3| until
% we know there is no failure.  With laziness we could easily lose confluence.
% % $$
% % |def (ones ^^ y ^^ z) (ones == tup (1,ones); tup (y,z) == ones; y)|
% % $$
% % Here, |ones| denotes an infinite list of |1|'s, and we might hope that the expression returns
% % |1|; but in \versecalc{} it won't because rule \rulename{subst} declines to substitute for the equality
% % |ones == tup(1,ones)|, which has an ``occurs check.''
%
% Another place that laziness could play a role is this.
% Remembering the duality between values and choices,
% one might also want |allKW| to return a lazy stream
% of results, one by one, rather than waiting for them all to terminate.  For example, one might hope that
% this program would converge:
% $$
% |def (y ^^ z) (tup (y, z) == all (def onec (onec == (1 choice onec))); y|
% $$
% Here we suppose that |allKW| returns a lazy stream of values (represented as nested
% pairs), from which we may pick the first and discard the rest.
%
% In short, there are good reasons for lenience, but
% a lazy variant of \versecalc{} could be worth exploring.

\section{\texorpdfstring{\versecalc{}}{The Verse calculus} in context: reflections and related work} \label{sec:related}

Functional logic programming has a rich literature; an excellent
starting point is Hanus's survey~\cite{hanus:survey}.  Now that we
know what \versecalc{} is, we can identify its distinctive features,
and compare them to other approaches.

\subsection{Choice and non-determinism} \label{sec:deterministic}

A significant difference between our presentation and earlier works is our
treatment of choice. Consider an expression like |(3 + (20 choice 30))|.
This is typically handled by a pair of non-deterministic rewrite rules:
$$
|e_1 choice e_2| \movesto |e_1| \hspace{1cm} |e_1 choice e_2| \movesto |e_2|
$$
So our expression rewrites (non-deterministically) to either |(3+20)| or |(3+30)|,
and that in turn allows the addition to make progress.
Of course, including non-deterministic choice means the rules are non-confluent by
construction.  Instead, one must generalize to say that a reduction does not change the
\emph{set} of results; in the context of lambda calculi, see
for example \citet{schmidt-schauss:non-det1,schmidt-schauss:non-det2}.

In contrast, our rules never pick one side or the other of a choice.
And yet, |(3 + (20 choice 30))| can still make progress by floating out the
choice (rule \rulename{choose} in \cref{fig:rules}), thus
|(3+20) choice (3+30)|.
In effect, \emph{choices are laid out in space} (in
the syntax of the term), rather than being explored by
non-deterministic selection.  Rule \rulename{choose} is not a new idea:
it is common in calculi with choice,
%%see {\eg}, ~\citetext{\citealp[Section 6.1]{liguoro:non-det}, \citealp[Section 3]{lago:probabilistic}}
see {\eg}, ~\citet[Section 6.1]{liguoro:non-det} and \citet[Section 3]{lago:probabilistic},
and, more recently, has been used to describe functional logic languages where it is variously
called \emph{bubbling} \cite{antoy:bubbling} or \emph{pull-tabbing} \cite{antoy:pull-tabbing}.
However, our formulation appears simpler because we avoid the need for attaching
an identifier to each choice with its attendant complications.

\subsection{One and all}

Logical variables, choice, and equalities are present in many functional logic
languages.  However, |oneKW| and |allKW| are distinctive features of \versecalc{},
with the notable exception of Fresh, a very interesting design introduced in a
technical report nearly 40 years ago~\cite{smolka:fresh} that also aims
to unify functional and logical constructs.
%
Fresh reifies choice into data via \emph{confinement} (corresponding to |oneKW|)
and \emph{collection} (corresponding to |allKW|).
%
However, Fresh differs from \versecalc{} in crucial ways.
%
First, it solves equations in a strictly left-to-right fashion, which means that
it is not lenient in the sense discussed in \cref{sec:lenient}.
%
Second, its semantics are presented in an operational fashion with explicit stacks and heaps,
in contrast to our focus on developing an equational account of functional logic programming.
%
Finally, Fresh appears not to have been implemented.
%
% It is a very interesting design, but one does not appear to have been implemented,

Several aspects of |allKW| and |oneKW| are worth noting.
First, |allKW| \emph{reifies} choice (a control operator) into a tuple (a data structure);
for example, |all (1 choice 7 choice 2)| returns the tuple |tup (1,7,2)|.
In the other direction, indexing turns a tuple into choice (for example, |def i (apply1 (tup (1,7,2)) i)| yields |(1 choice 7 choice 2)|).
Other languages can reify choices into a (non-deterministic)
list, via an operator called \verb!bagof!, or a mechanism called
\emph{set-functions} in an extension of Curry \cite[Section 4.2.7]{curry:tutorial},
implemented in the Kiel Curry System interpreter \cite{brassel:encapsulated,kics,hanus:set-functions}.
But in Curry, this is regarded as a somewhat sophisticated feature,
whereas it is part of the foundational fabric of \versecalc{}.
Curry's set-functions need careful explanation about sharing across non-deterministic choices,
or what is ``inside'' and ``outside'' the set function, something that appears as a straightforward
consequence of \versecalc{}'s single rule \rulename{choose}.

Second, even under the reification of |allKW|,
\versecalc{} is \emph{deterministic}.
\versecalc{} takes pains to maintain order, so that
when reifying choice into a tuple, the order of elements in that
tuple is completely determined.
This determinism has a price: we have to take care to maintain the
left-to-right order of choices (see \cref{sec:mixing} and \cref{sec:choice-rules}, for example).
However, maintaining that order has other payoffs. For example, it is relatively easy to add
effects other than choice, including mutable variables and input/output, to \versecalc{}.
To substantiate this claim, \cref{sec:store} gives the additional syntax and rewrite rules
for mutable variables.

Thirdly, |oneKW| allows us to reify failure; to try something and take different actions
depending on whether or not it succeeds. Prolog's ``cut'' operator has a similar flavor,
and Curry's set-functions allow one to do the same thing.

Finally, |oneKW| and |allKW| neatly encapsulate the idea of ``flexible'' {\vs} ``rigid'' logical variables.
As we saw in \cref{sec:rigid}, logical variables bound outside |oneKW|/|allKW| cannot be unified inside it;
they are ``rigid.''  This notion is nicely captured by the fact that equalities
cannot float outside |oneKW| and |allKW| (\cref{sec:normalization}).

\subsection{The semantics of logical variables} \label{sec:magical}

Our logical variables, introduced by |defKW|, are often called
\emph{extra variables} in the literature, because they
are typically introduced as variables that appear on the right-hand side of a
function definition, but are not bound on the left.  For example, in Curry we can write:
\begin{verbatim}
     first x | x =:= (a,b) = a where a,b free
\end{verbatim}
Here, \verb!a! and \verb!b! are logical variables, not bound on the left; they
get their values through unification (written ``\verb!=:=!'').   In Curry, they are explicitly introduced
by the ``\verb!where a,b free!'' clause, while in many other papers their
introduction is implicit in the top-level rules, simply by not being bound
on the left.  These extra variables (our logical variables) are
at the heart of the ``logic'' part of functional logic programming.

Constructor-based ReWrite Logic (CRWL) \cite{crwl} is the brand leader for
high-level semantics for non-strict, non-deterministic functional logic languages.
%
CRLW is a ``big-step'' rewrite semantics that rewrites a term to a value in a single
step.
%
\Citet{fraguas:rewriting} make a powerful case for instead giving the semantics
of a functional logic language using ``small-step'' rewrite rules, more like those
of the lambda calculus, that successively rewrite the term, one step at
a time, until it reaches a normal form.
Their paper does exactly this, and proves equivalence to the
CRWL framework.  Their key insight (like us, inspired by
\citet{ariola:call-by-need}'s formalization of the call-by-need lambda calculus)
is to use |let| to make sharing explicit.

However, both CRWL and Fraguas {\etal} suffer from a major problem:
they require something we call \emph{magical rewriting}. A key rewrite rule is this:
$$
\begin{array}{l}
f( \theta(e_1),\ldots, \theta(e_n) ) \movesto \theta( rhs ) \\
\text{if}
\begin{array}[t]{l}
  (e_1, \ldots, e_n) \movesto rhs
   \; \text{is a top-level function binding, and} \\
\text{$\theta$ is a substitution mapping variables to closed values, s.t.
$dom(\theta) = \mathit{fvs}(e_1, \ldots, e_n, rhs)$}
\end{array}
\end{array}
$$
The substitution for the free variables of the left-hand-side can
readily be chosen by matching the left-hand-side against the call.
But the substitution for the extra variables must be chosen ``magically''
\cite[Section 7]{fraguas:rewriting} or clairvoyantly, so as to make the future execution
work out. This is admirably high level because it hides everything
about unification, but it is not much help to a programmer trying to
understand a program, nor is it directly executable.
In a subsequent journal paper, they refine CRWL to avoid magical rewriting using
``let-narrowing'' \cite[Section 6]{fraguas:let-narrowing}; this system
looks rather different to ours, especially in its treatment of choice,
but is rather close in spirit.

To explain actual execution, the state of the art is described by
\citet{hanus:semantics}.  They give both a big-step operational
semantics (in the style of \citet{launchbury:natural}), and
a small-step operational semantics.  These two approaches both thread a
\emph{heap} through the execution, which holds the unification variables and their unification
state; the small-step semantics also has a \emph{stack}, to specify
the focus of execution.
%
The trouble is that heaps and stacks are difficult to explain to a programmer,
and do not make it easy to reason about program equivalence.
%
In addition to this machinery, the model is further complicated with
concurrency to account for residuation.

In contrast, our rewrite rules give a complete, executable (\ie, no ``magic'')
account of logical variables and choice, directly as small-step
rewrites on the original program, rather than as the evolution of a
(heap, control, stack) configuration. Moreover, we have no problem with
residuation.

\subsection{Flat {\vs} higher order}

When giving the semantics
of functional logic languages, a first-order presentation is almost universal.
User-defined functions can be defined at top level only;
and function symbols (the names of such functions) are syntactically
distinguished from ordinary variables.
% As Hanus describes, it is possible
% to translate a higher-order program into a first-order
% form\footnote{
%   Hanus does not mention this, but for a language with arbitrarily nested lambdas
%   one would need to do lambda-lifting as well, but that is perhaps a minor point.}
% using defunctionalization~\cite[Section 3.3]{hanus:survey}, and a built-in
% |apply| function.
As Hanus describes, it is possible
to translate a higher-order program into a first-order
form using defunctionalization~\cite[Section 3.3]{hanus:survey}, and a built-in
|apply| function.
(Hanus does not mention this, but for a language with arbitrarily nested lambdas,
one would need to do lambda-lifting \cite{johnsson:lambdalift} as well; this is perhaps a minor point.)
Sadly, this encoding is hardly a natural rendition of the lambda calculus, and
it obstructs the goal of using rewrite rules to explain to programmers how
their program might execute. In contrast, a strength of our \versecalc{}
presentation is that it deals natively with the full lambda calculus.

\subsection{Intermediate language}

Hanus's \emph{Flat Language}~\cite[Fig 1]{hanus:semantics}, FLC, plays the same role as \versecalc{}:
it is a small core language
into which a larger surface language can be desugared.
There are some common features: variables, literals, constructor applications,
and sequencing (written \verb!hnf! in FLC).
However, it seems that \versecalc{} has a greater economy of concepts.
In particular, FLC
has two forms of equality \verb!(==)! and \verb!(=:=)!, and two forms
of case-expression, \verb!case! and \verb!fcase!.  In each pair, the former suspends
if it encounters a logical variable; the latter unifies or narrows
respectively. In contrast, \versecalc{}
has a single equality |(=)|, and the orthogonal |oneKW| construct, to deal with all four concepts.

FLC has let-expressions \verb!(let x=e in b)!, where \versecalc{} uses
|defKW| and (again) unification.
FLC also uses the same construct for a different purpose, to
bring a logical variable into scope, using the strange binding \verb!x=x!, thus \verb!(let x=x in e)!.
In contrast, |def x e| seems more direct.

\subsection{Comparison with Icon} \label{sec:icon}

There are many obvious similarities between Verse and the Icon programming language
\cite{griswold:icon-users-manual,griswold:generators-in-icon,griswold:icon-book-1983,griswold:icon-overview,griswold:icon-book-2002}:
\begin{itemize}
\item An expression can (successively) produce any number of values.
\item An expression that produces zero values is said to \emph{fail} \cite[\S 3.1]{griswold:generators-in-icon};
  an expression that produces at least one value is said to \emph{succeed}.
\item The expression |e_1 choice e_2| produces all the values of |e_1| followed by all the values of |e_2|.
\item There is a way to turn an array (or tuple) $a$ into a sequence of produced values.
  In Icon, this is written \texttt{!|a|} \cite[\S 3]{griswold:icon-overview}; in Verse, \texttt{|a|?}; in \versecalc{}, |def (i) (apply1 a i)|.
\item Most ``scalar'' operations (such as addition and comparisons) run through
  all possible combinations of values of their operand expressions,
  using a specific left-to-right evaluation order and automatic chronological backtracking.
\item Success and failure are used in place of boolean values for control-structure purposes.
  Some operations, especially comparisons, can fail as part of their defined semantics.
  The expression |if e_1 then e_2 else e_3| checks to see whether |e_1| succeeds;
  it then produces the values of |e_2| (if |e_1| succeeded) \emph{or} produces the values of |e_3| (if |e_1| failed).
  If |e_1| succeeds and then |e_2| fails, backtracking does \emph{not} attempt to examine further values from |e_1|.
\item The ``|`choice`|'' construct is idiomatically used as a \emph{logical or} operation \cite[\S 3]{griswold:icon-overview}.
\item There is a control structure that executes a specified expression once for every value produced by another expression.
  In Icon, this is \texttt{every}~|e_1| \texttt{do}~|e_2| and in Verse, it is written \texttt{for(}|e_1|\texttt{)} \texttt{do}~|e_2|;
  in each, the ``\texttt{do}~|e_2|'' may be omitted to simply evaluate |e_1| repeatedly until it has yielded all its values.
  (In Icon, \texttt{every} |e| may also be written as \texttt{repeat} |e|.)
\item It is impossible to name a generator (Icon) or choice (Verse); if |e| produces multiple values,
  \texttt{|x| := |e|} will provide one value at a time from |e| to be named by variable |x|.
\end{itemize}

But there are also major differences between Verse and Icon.
Icon was designed primarily to use expressions as generators to automatically
explore a combinatorial space of possibilities (``goal-directed evaluation''), and secondarily to use
success/failure rather than booleans to drive control structure.
But in other respects, \emph{Icon is a fairly conventional imperative language}, relying
on side effects (assignments) to process the generated combinations.
The designers judged that the interactions of such side effects with completely
unrestrained control backtracking would be difficult for programmers to
understand \cite[\S 3.1]{griswold:generators-in-icon}; therefore, the design of Icon emphasizes limited scopes for control
backtracking and tools for controlling the backtracking process \cite[\S 3.3]{griswold:generators-in-icon}.

In contrast, \emph{Verse is a declarative language} and
avoids these difficulties by using a functional logic approach
rather than an imperative approach to processing generated combinations:
\begin{itemize}
\item While Icon typically processes multiple values from an expression by using \emph{assignment},
  Verse typically processes multiple values by using \emph{equations}
  (which are then \emph{solved}).
\item
  Verse also has a concise way to turn a finite sequence of multiple
  values into an array directly.
  For example, to make variable |a| refer to an array containing all values generated by expression |e|,
  code such as the following (using a \texttt{repeat} loop containing an assignment) is idiomatic in Icon \cite[\S 8]{griswold:icon-overview}:
  \begin{center}
  \texttt{a := array 0 string; i := 0; repeat a[i+] := e; close(a)}
  \end{center}
  (In Icon, \emph{postfix} \texttt{+} is an autoincrement operator \cite[\S 3.1.2]{griswold:icon-users-manual}
  with the same behavior as \emph{prefix} \texttt{++} in~C \cite[\S 2.8]{kr-c-book}.)
  In Verse, \texttt{a~=~for(x :== e) do~x} or even just
  \texttt{a~=~for\{e\}} does the job; in \versecalc{}, |a == all(e)| is all it takes.
\item Backtracking in Icon is ``only control backtracking;'' side effects, such as assignments, are not undone \cite[\S 3.1]{griswold:generators-in-icon}.
\item Both languages have an implicit ``cut'' (permanent acceptance of the first produced value) after the predicate part of an |if|-|then|-|else|,
  but Icon furthermore has an implicit cut at each statement end (semicolon or end of line) \cite[\S 3.1]{griswold:generators-in-icon}, each
  closing brace ``\texttt{\}}'', and most keywords \cite{icon-programming-corner-4}.
\end{itemize}



\section{Looking back, looking forward}

We believe that this is the first presentation of a functional logic
language as a determinstic rewrite system.  A rewrite system has the
advantage (compared to more denotational, or more operational,
methods) that it is is sufficiently low-level to capture the
\emph{computational model} of the language; and yet sufficiently
high-level to be \emph{illuminating} to a programmer or compiler
writer.  Our focus on rewriting as a way to define the
semantics has forced us to focus on confluence, a rather syntactic
property that is stronger (and hence more delicate and harder to prove) than the
contextual equivalence that we really need.  That in turn led us to
study the elegant and ingenious notion of skew confluence, which has been
barely revisited in 25 years, but which we believe deserves a wider
audience.

% Defining a language by giving its low-level semantics is precise but not necessarily illuminating;
% for example, giving a reference compiler that compiles
% the program to x86 instructions is precise, but is not helpful
% to a human who is trying to understand exactly what the original program
%  meant. Likewise, a high-level (\eg, denotational) semantics simply provides the eventual answer
% produced by the program without insight into the computational steps
% that got us from program start to program completion.

%%\simon{dropped a short para here that interrupted the flow}
% The moral here is that \emph{formal specifications can be obfuscatory---or
% illuminating.} The latter kind shed light because they are defined \emph{in
% terms of the intended mechanisms of the language}. \versecalc{} does this; it
% respects the conceptual structures of the Verse language.

% Note that when we say ``illuminating'' we mean that in multiple ways.
% A semantics can be illuminating for humans who are trying to understand
% what a particular program does, or how a proposed change to the language
% will affect the language. It can drive analyses that help the compiler
% optimize programs. It underlies the use of formal methods and verification
% to provide machine-derived and -checkable proofs of correctness. All of
% these applications depend on the semantics being defined at the appropriate
% level: the level of the computational model that underlies the language.
% This has been our goal in this work.

We have much left to do.  The full Verse language has
statically checked types.  In the dynamic semantics, the types can be
represented by partial identity functions---identity for the values of
the type, and |fail| otherwise.  This gives a distinctive new
perspective on type systems, one that we intend to develop in future
work.  The full Verse language also has a statically checked effect system,
including both mutable references and input/output.  All these effects
must be \emph{transactional}, {\eg}, when the condition of an |if|
fails, any store effects in the condition must be rolled back.
We have preliminary reduction rules for updateable references,
see \cref{sec:store}.

% -------------------- Acknowledgements ----------------
\begin{acks}                            %% acks environment is optional
                                        %% contents suppressed with 'anonymous'
  We thank our colleagues for their helpful and specific
  feedback on earlier drafts of this paper,
  including
  Jessica Augustsson,
  Francisco L\'opez-Fraguas,
  Andy Gordon,
  Michael Hanus,
  David Holz,
  Juan Rodr\'iguez Hortal\'a,
  John Launchbury,
  Dale Miller,
  Andy Pitts,
  Niklas R\"ojemo,
  Jaime S\'anches-Hern\'andez,
  Andrew Scheidecker,
  and Stephanie Weirich.
\end{acks}

% -------------------- Bibliography ----------------
\bibliography{../verse}

\clearpage

\appendix

\clearpage
% ----------------------- Example

\section{Example}
A complete reduction sequence for a small example can be found in \cref{fig:sample-reductions}.
This example shows how constraining the output of a function call can constrain the argument.
While most of the reductions are administrative in nature, these are the highlights:
At \circlednum{1} the |swap| function is inlined so that at \circlednum{2} a \textbeta-reduction can happen.
Step \circlednum{3} inlines the argument, and \circlednum{4} does the matching of the tuple.
At \circlednum{5} and \circlednum{6} the actual numbers are inlined.
\begin{figure}[h]
\begin{footnotesize}
% no space needed around math environments in a figure
\setlength{\abovedisplayskip}{0pt}%
\setlength{\belowdisplayskip}{0pt}%
$$
\begin{array}{@@{}l@@{\;}l@@{\;}l@@{}}
                &                    & |applyN swap (x,y) :== tup(y, x); def (p) (apply1 swap p = tup(2,3); p)| \\
                & \movestoR{desugar} & |def (swap) (swap == (lam xy (def (x ^^ y) (tup (x,y) == xy ; tup (y,x)))) ; def (p ^^ t) (t == apply1 swap(p) ; t == tup (2,3) ; p))| \\


\circlednum{1}  & \movestoR{subst,eqn-elim}   & |def (p ^^ t) (t == apply1 (lam xy (def (x ^^ y) (tup (x,y) == xy ; tup (y,x))))(p) ; t == tup (2,3) ; p)| \\
                & \movestoR{subst,eqn-elim}   & |def p ((2,3) == apply1 (lam xy (def (x ^^ y) (tup (x,y) == xy ; tup (y,x))))(p) ; p)| \\
\circlednum{2}  & \movestoR{app-beta}          & |def p ((2,3) == (def xy (xy == p; def (x ^^ y) (tup (x,y) == xy ; tup (y,x)))) ; p)| \\
                & \movestoR{exi-float}          & |def (p ^^ xy) ((2,3) == ((xy == p; def (x ^^ y) (tup (x,y) == xy ; tup (y,x)))) ; p)| \\
\circlednum{3}  & \movestoR{subst,eqn-elim}   & |def p ((2,3) == (def (x ^^ y) (tup (x,y) == p ; tup (y,x))) ; p)| \\
                & \movestoR{exi-float,exi-float} & |def (p ^^ x ^^ y) ((2,3) == (tup (x,y) == p ; tup (y,x)) ; p)| \\
                & \movestoR{eqn-float,seq-assoc}& |def (p ^^ x ^^ y) (tup (x,y) == p; (2,3) == tup (y,x) ; p)| \\
                & \movestoR{hnf-swap}          & |def (p ^^ x ^^ y) (p == tup (x,y); (2,3) == tup (y,x) ; p)| \\
                & \movestoR{subst,eqn-elim}   & |def (x ^^ y) ((2,3) == tup (y,x) ; tup(x,y))| \\
\circlednum{4}  & \movestoR{u-tup,seq-assoc}    & |def (x ^^ y) (2==y; 3==x; tup(x,y))| \\
                & \movestoR{hnf-swap}          & |def (x ^^ y) (y==2; 3==x; tup(x,y))| \\
\circlednum{5}  & \movestoR{subst,eqn-elim}   & |def x (3==x; tup(x,2))| \\
                & \movestoR{hnf-swap}          & |def x (x==3; tup(x,2))| \\
\circlednum{6}  & \movestoR{subst,eqn-elim}   & |tup(3,2)| \\
\end{array}
$$
\end{footnotesize}
\caption{A sample reduction sequence} \label{fig:sample-reductions}
\Description{A sample reduction sequence}
\end{figure}


\input{../confluence/03-preliminaries}
\input{../confluence/04-confluence}

\clearpage
\input{../confluence/08-skew}


\clearpage
% ----------------------- Denotational semantics  -----
\input{densem-icfp23}

\clearpage
\input{store-icfp23}


\end{document}

% LocalWords:  Lennart Augustsson nnnn Peyton Joachim Breitner ccs ce
% LocalWords:  desc tup optimise defunctionalization oneKW allKW lll
% LocalWords:  distils metaverse Ariola subst utup elim Hanus lrcl sx
% LocalWords:  residuation Primops isInt exprs CX SX rcl applyN defKW
% LocalWords:  modelled hnf dlet ys xs emptytup lamhdr xr zs unary
% LocalWords:  boolean un rcll ANF flatMap cartesian behavior fromTo
% LocalWords:  featureful multivalued mapFlat powerset denotational
% LocalWords:  Launchbury ulit ux localising uc seql seqr unifyl ary
% LocalWords:  unifyr neighbourhood ariola deterministically bagof xy
% LocalWords:  ReWrite CRWL fraguas formalization rhs dom fvs hanus
% LocalWords:  launchbury fcase labelled splitKW lrcll SML emp Niklas
% LocalWords:  updateable alloc storeTwoLines canCopy storeKW Koen ss
% LocalWords:  subexpression Claessen equationally eliml fcn elimr
% LocalWords:  rll redex Formalising equational stepX compositionally
% LocalWords:  iff cuten stepXs stepXrs Rosser QuickCheck quickcheck
% LocalWords:  joinability metatheory fixpoint liguoro det lago CRLW
% LocalWords:  Smolka flavor FLC urefl onec obfuscatory opez Rodr
% LocalWords:  transactional iguez Hortal Ranjit Jhala ojemo anches
% LocalWords:  Hern andez Scheidecker stepA stepB stepAe stepArs cutr
% LocalWords:  stepAs stepBr stepAr stepAB stepABrs stepABs stepUn
% LocalWords:  Eqns ohlebusch
