%include lhs2TeX.fmt
%include lhs2TeX.sty
%include polycode.fmt
%include ../calculus.fmt

\section{Updateable references} \label{sec:store}

\begin{figure}
  $$
  \begin{array}{l}
    \textbf{Syntax extension} \\
    \begin{array}{lrcl}
      \text{References}         & r \\
      \text{Expressions}        & e   & ::= & \cdots \vb |store h e| \\
      \text{Primops}            & op  & ::= & \cdots \vb |alloc| \vb |read| \vb |write| \\
      \text{Head values}        & hnf & ::= & \cdots \vb r \\
      \text{Execution contexts} & X   & ::= & \cdots \vb |store h X| \\
      \text{Scope contexts}     & SC  & ::= & \cdots \vb |store h SC| \\
      \text{Heap}               & h   & ::= & |emp| \vb |cell r v, h| \\
      \text{Heap context}       & H   & ::= & \hole|,h| \vb |cell r v, H| \\
      \text{Store contexts}     & S   & ::= & \hole \vb |v == S| \vb |S ; e| \vb |sf; S| \vb |def x S| \\
      \text{Store-op free exprs}&|sf| & ::= & |v| \vb |sf_1 = sf_2| \vb |sf_1; sf_2| \vb |def x sf|
                                              \vb |apply1 sp v|
                                              %%% need sv \vb |split sf sv_1 sv_2|
                                              \\
      \text{Results}            & w   & ::= & |v| \vb |v `choice` e| \\
      \text{Non-store primops}  & sp  & ::= & \text{any, except } |alloc|, |read|, |write| \\
      \text{Non-store expression} & oe  & ::= & \text{like |e|, but not |w|, |storeKW|, or |fail|} \\
    \end{array} \\
\\

  \textbf{Axiom extensions} \\
  \textit{Normalization change} \\
  \begin{array}{l r @@{\;}c@@{\;} l l}
%  \begin{array}{lrcll}
    \rulename{exi-float} & |X^[def x e]| & \movesto & |def x (X^[e])| & \text{if $|X| \ne \hole,\,  x \notin \freevars{|X|}$, \alphamark}\\
    &&&& \text{if there is |storeKW| in |X| then |e| $\in$ |ce|} \\
  \end{array} \\

  \textit{Reference ops} \\
  \begin{array}{l r @@{\;}c@@{\;} l l}
%  \begin{array}{lrcll}
    \rulename{ref-alloc} & |store h (S^[apply1 alloc v])| & \movesto & |store (cell r v, h) (S^[r])| \\
    &&& \quad \freevars{|v|} \# \boundvars{|S|}, \text{|r| fresh}  \\
    \rulename{ref-read} & |store (H^[cell r v]) (S^[apply1 read r])| & \movesto & |store (H^[cell r v]) (S^[v])| \\
    &&& \quad \freevars{|v|} \# \boundvars{|S|}, \text{\alphamark}  \\
    \rulename{ref-write} &
        | store (H^[cell r v_1]) (S^[applyN write (r,v_2)])|
        & \movesto & |store (H^[cell r v_2]) (S^[emptytup])| \\
    &&& \quad \freevars{|v_2|} \# \boundvars{|S|}  \\
  \end{array} \\
    \\

  \textit{Store duplication} \\
  \begin{array}{l r @@{\;}c@@{\;} l l}
%  \begin{array}{lrcll}
    \rulename{st-split-dup} & \hspace{-1em} |store h (S^[split oe f g])| & \movesto & |store h (S^[split (store h oe) f g])| \\
    &&& \quad \freevars{|h|} \# \boundvars{|S|}, \text{\alphamark}  \\
    \rulename{st-choice-dup} & |store h (oe `choice` e)| & \movesto & |store h (store h oe `choice` e)| \\
  \end{array} \\

    \\
  \textit{Store commit} \\
  \begin{array}{l r @@{\;}c@@{\;} l l}
%  \begin{array}{lrcll}
     \rulename{st-split} & |store h_1 (S^[split (store h_2 w) f g])| & \movesto &
         |store h_2 (S^[split w f g])| \\
    &&& \quad \freevars{|h_2|} \# \boundvars{|S|}  \\
     \rulename{st-choice} & |store h_1 (S^[(store h_2 w) `choice` e]| & \movesto & |store h_2 (S^[w `choice` e])| \\
    &&& \quad \freevars{|h_2|} \# \boundvars{|S|}  \\
  \end{array} \\

    \\
    \textit{Unification} \\
    \begin{array}{lrcll}
      \text{Extension with the obvious axioms making equal references unify,
        and anything else fail.}
    \end{array} \\
  \textit{Top level} \\
  \begin{array}{lrcll}
    \text{Start top level reduction of |e| with |store emp e|.}
  \end{array} \\

  \end{array}
   $$
  \caption{The Verse calculus: store axioms} \label{fig:store-axioms}
\end{figure}
The full Verse language has updatable references (à la ML).
There are three new primitive operations, |alloc|, |read|, and |write|.
The |alloc| creates a new reference with an initial value,
|read| extracts the value from a reference,
and |write| sets the value of a referene.

Modifying these references is transactional in the sense that if a computation fails,
then any updates will not be visible outside the construct that handles the failures.
\Eg,\\
|r:=apply1 alloc 0; (if (applyN write (r, 1); fail) then 1 else 2); apply1 read r|\\
will have the value |0|,
because the |write| is part of an expression that fails, and so its effect is not visible.


To add updateable references we extend the system with syntax and rules from figure~\ref{fig:store-axioms}.
The |store h e| indicates that |e| should be reduced using the heap |h|.
A heap is simply a mapping from references to values (one mapping being |cell r v|).
A reference is some opaque type that supports equality (unification) and creation of a new reference.

The interaction of the new primitives with the store can be seen from the axioms.
The |apply1 alloc v| operation creates a new reference and adds a binding with |v| to the store.
The |apply1 read r| operation retrieves the value for reference |r| from the store,
and |applyN write (r, v)| updates the reference |r| with |v| in the store.
All of these operations use the context |S| which ensures that there are no store operations
to the left of the hole, \ie, a store operation in the hole is the next one that should execute.

The interesting rules involve choice and |splitKW| because store operations are transactional in the sense
that when an expressions fails, none of its store operations will happen.

When reducing |split e f g| in an |S| hole, rule \rulename{st-split-dup}, the store is duplicated.
Any store operations inside the |splitKW| will happen in this local copy of the store.
Note the two occurrences of |h| in the right hand side of \rulename{st-split-dup}.
If the reduction of |e| results in |fail| then rule \rulename{fail-elim} is used,
and the store from the failing computation is simply thrown away.
If the reduction of |e| results in a value (with or without more alternatives) then rule
\rulename{st-split} is used.  This rule replaces the outer store with the inner store,
since we know the inner computation has succeeded.

Similarely, the reduction of |e_1 `choice` e_2| will duplicate the store into the first branch, \rulename{st-choice-dup}.
Here |e_1| must not contain any store operation nor be a value.
And again, similarely, \rulename{st-choice} commits the new store and throws away the old.

The use of |oe| in the rules is to ensure that the rules cannot get stuck in a loop.
Using |e| instead of |oe| would mean that failing or committing would make the expression match the duplication rule again.
It also prevents the duplication rule from repeatedly duplicating the |storeKW|.

Note that |storeKW| is part of the |X| context, which means that the |storeKW| can float inside
existentials.
This is necessary for the store rules to fire since the |S| context does not allow
going under existentials.
%If |S| contained |storeKW| then an |alloc| or |write| could update the store with
%a variable that is not in scope.

The semantics of |for(d) do e| with respect to store effects is somewhat intricate.
The expression |d| is possibly multi-valued; any effects that happens when computing the
first value of |d| will be visible the first time |e| is computed.
Both these effects are then visible when computing the second value of |d|, and so on.
If any iteration of |d| fails, then the effects of that computation are not visible outside |d|.
This means that the desugaring of |for| into |split| needs to be more elaborate.

|for(def (x_1 xdots x_n) d) do e|\\
means\\
\hspace*{5mm}|apply1N f emptytup := emptytup;|\\
\hspace*{5mm}|g^(v)^(r) := (v==tup(x_1,xdots,x_n); applyN cons (e, split (apply1N r emptytup) f g);|\\
\hspace*{5mm}|split (def (x_1 xdots x_n) (d; tup(x_1,xdots,x_n))) f g|

To support limited store operations ({\eg}, |read|,
but not |write|) we can equip the store with a set of currently allowed operations.
We also need some extra primitives that modify this set.

\subsection{Examples}
\lennart{Not yet}

