%include lhs2TeX.fmt
%include lhs2TeX.sty
%include polycode.fmt
%include ../calculus.fmt
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Overview}\label{sec:overview}

The \verifier takes as input a surface term $t$ in the source verse
language and \emph{desugars} it into a verifier term $e$ in the core
language that is \vcshort (the ICFP23 calculus) extended with
%
(1) \emph{syntax} to record \emph{assumptions} and \emph{obligations},
%
(2) \emph{rewrite rules} that use the assumptions to discharge obligations.
%
Next, the \verifier repeatedly applies \emph{all}
the rewrite rules, that is, the ICFP23 rules \emph{and}
the new verification rules, such that either
%
(a)~we obtain a term |e'| such that $\isdone{e'}$, \ie which
contains no obligations, in a covariant position, in which case
|t| is \emph{accepted}, or
%
(b)~we obtain a normal form |e'| that still contains obligations
in a covariant position, in which case |t| is \emph{rejected}, or
%
(c)~the rewriting times out, in which case |t| is \emph{rejected}.

Next, we describe the extensions to the syntax \S~\ref{sec:syntax},
explain the desugaring by means of examples \S~\ref{sec:desugar},
present the verification rewrite rules \S~\ref{sec:rules}, and
finally show a few examples that show how the verifier works from
end-to-end \S~\ref{sec:examples}.


\subsection{Why do we need |verify|?}
%
(In addition to |succeedKW| and |decideKW|?)
%
The TRS-verifier converts a function definition |f^(x:int):int :== x| into two things:
%
(1) An \emph{obligation} that in fact, the body of |f| returns an |int| when the input
    parameter |x| is an |int|, and
%
(2) A \emph{definition} for the name |f| that can then be used by other code in the program.
%
This is done by desugaring the definition of |f| into
%
\begin{align*}
%
& |def f (verify (lam y (def x (assume (apply1 int y; x = y; x); succeed (apply1 int x; arr ())))); f = v_f)| \\
%
\intertext{where |v| is the desugared body of |f|, namely}
|v_f| \doteq\ & |lam x (apply1 int x; assume (apply1 int x))| \\
%
\intertext{We use the |verify| to record the obligation.
If we omitted it to desugar the above to}
& |def f (verify (lam y (def x (assume (apply1 int y; x = y; x); succeed (apply1 int x; arr ())))); f = v_f)| \\
\intertext{then the \rulename{val-elim} rewrite rule would let us \emph{unsoundly erase}
the proof obligation without first in fact checking that it holds.
%
However, one might wonder: why not record obligations using a
$\exists$ binder instead of $\lambda$?
%
In other words, why not instead desugar the verification of
|f^(x:int):int :== x| as}
%
& |def f ((def y (def x (assume (apply1 int y; x = y; x); succeed (apply1 int x; arr ())))); f = v)|
\end{align*}
%
where |v_f| is the desugared body of |f|, as shown above?
%
The reason is we want to distinguish between $\lambda$ and $\exists$
binders as explained next.

\subsection{Why distinguish between $\lambda$ and $\exists$ binders?}
%
Consider the following examples:
%
$$\begin{array}{ll}
    |f^(x:int) := {y:any; if (x > 1) then { y = 1 } else { y = 2 }; y}| & \text{\# succeeds}\\
    |g^(x:int) := {y:any; if (y > 1) then { y = 1 } else { y = 2 }; y}| & \text{\# stuck}\\
\end{array}$$
%
The function |f| should be \emph{accepted} by the verifier as indeed,
|f| succeeds producing either |1| or |2| depending on the value of |x|.
%
However |g| should be \emph{rejected} by the verifier as the guard depends
upon |y| whose value cannot be determined without first executing the branches!
%
The key difference between |f| and |g| is that |f| uses the $\lambda$-bound |x|
in the guard, while |g| uses the $\exists$-bound |y|.
%
Hence, the TRS verifier handles the former specially, to determine that
a term can make progress (via the $\mustDecide{}{}$ predicate) when its
free variables are $\lambda$-bound.


\subsection{Why not push |succeedKW| inside?}
%
Currently, we verify a term |succeed e| (resp. |decide e|) by
%
(1)~using \rulename{asm-intro} to annotate sub-terms of |e| with |assumeKW|, and then,
%
(2)~using \rulename{suc-elim} (resp. \rulename{dec-elim}) to discharge the |succeedKW| (resp. |decideKW|).
%
An alternative approach might be to \emph{push} the |succeed| inside the term |e|, \eg via a rule like
%
$$\begin{array}{ll}
    \rulename{suc-exi}   & |succeed (def x e)| \movesto |def x (succeed e)| \\
    \rulename{suc-seq}   & |succeed (e_1; e_2)| \movesto |succeed e_1; succeed e_2| \\
    \rulename{suc-elim'} & |P^[succeed e]| \movesto |P^[e]| \quad \text{if}\ \proves{P}{e} \\
\end{array}$$
%
Unfortunately, this doesn't work as shown by the following example:
%
$$\begin{array}{ll}
                   & |succeed (def x (x = 1; x))| \\
\movestoR{suc-exi} & |def x (succeed (x = 1; x))| \\
\movestoR{suc-seq} & |def x (succeed (x = 1); succeed x)| \\
\end{array}$$
%
Now, we are stuck: we cannot prove that the unification |x = 1| succeeds.
%
In contrast, with our current strategy, we get the following rewrite sequence that discharges the |succeedKW|:
%
$$\begin{array}{ll}
                    & |succeed (def x (x = 1; x))| \\
\movestoR{subst}    & |succeed (def x (x = 1; 1))| \\
\movestoR{eqn-elim} & |succeed (1)| \\
\movestoR{suc-elim} & |1|
\end{array}$$
%