\documentclass[acmsmall,screen,review,9pt]{acmart}\settopmatter{}

% Put \everythingfalse in a file local.tex to build a streamlined
% version with just DPS semantics
\newif\ifeverything \everythingtrue
\IfFileExists{local.tex}{\input{local.tex}}{}

\let\Bbbk\relax      %% avoid LaTeX Error: Command `\Bbbk' already defined.
%include lhs2TeX.fmt
%include lhs2TeX.sty
%include polycode.fmt
%include forall.fmt
%include ../LaTex/calculus.fmt
% Boxed figures
\floatstyle{boxed}
\restylefloat{figure}
\usepackage{../LaTex/quiver}
\usepackage{framed}
\usepackage{amssymb}
\usepackage{cleveref}   % \Cref etc
\usepackage{mathpartir}
\usepackage{stmaryrd}
\usepackage{conteq}
\usepackage{textgreek}
\usepackage{xspace}
\usepackage{mathtools}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{multirow}
\usepackage{makecell}
\input{../LaTex/verse-macros}

\begin{document}

\title{The Specification of Verse}
\author{Tim, Jan, Andy S, Simon, Lennart, Koen, Ranjit, Stephanie, Benjamin, \today}
\ifeverything
\maketitle
\fi

\ifeverything
\section{Overview}

% Updated July 15 2025 by BCP from Simon's description in MaxVerse doc

\begin{itemize}
\item {\bf Source Verse} (\Cref{fig:source-verse-expr,fig:source-verse-expr2}) is an approximation of the full source language.
\item The \emph{superficial-desugar} pass desugars
(\Cref{fig:sv-to-sv-1,fig:sv-to-sv-2}) Source Verse into
  {\bf Essential Verse} (\Cref{fig:small-source-verse}), whose distinctive feature is TimLambdas.
  Essential Verse is a subset of Source Verse.
\item A \emph{wrapping transformation} (\Cref{fig:mini-verse}) further
desugars Essential Verse into {\bf Mini Verse}
(\Cref{fig:mini-verse}).
\item A \emph{translation for verification and execution}
(\Cref{fig:ds12})
takes Mini Verse to {\bf Core Verse} (\Cref{fig:core-verse}).  The
operational semantics of Core Verse in terms of rewrite and
verification rules is defined in \Crefrange{fig:vc-syntax}{fig:verifier-rules-arrays}.
(As of July 2025, \Cref{fig:vc-contexts} is out of date.  Not clear
how \Cref{fig:ds12} fits in.)
\item We are currently experimenting with three different approaches
to the {\bf denotational semantics} of Mini Verse. (Tim and Koen prefer DPS;
Simon finds expression style easier to grok.)
\Cref{fig:densem-notation} defines some common notations.
\begin{enumerate}
\item \textbf{(E-LV)}: expression-style, list of values.
  \begin{itemize}
  \item $\semappsub{E}{e}{|e|} : \Env \rightarrow \seq{ \VAL }$
  \item \Cref{fig:e-lv1} and \Cref{fig:e-lv2}
    \end{itemize}
\item \textbf{(D-LS)}: dps-style, list of sets of environments.
  \begin{itemize}
  \item $\semapsub{E}{d}{|e|} r : \seq{ \ENV }$
  \item \Cref{fig:d-ls1} and \Cref{fig:d-ls2}
    \end{itemize}
\item \textbf{(D-SLS)}: dps-style, sets of lists of sets of environments.
  \begin{itemize}
  \item $\semapsub{E}{d}{|e|} r : \powersetnp{\seq{ \ENV }}$
  \item \Cref{fig:d-sls1} and \Cref{fig:d-sls2}
    \end{itemize}
\end{enumerate}
\item \Cref{fig:jan-style-dps-1,fig:jan-style-dps-2}
are work in progress on a ``Jan-style'' variant of the DPS semantics.
\item Lennart is working on reviving a ``direct denotational
semantics'' for Essential Verse  (\Crefrange{fig:den-sem1}{fig:den-sem-dest1}).
\end{itemize}

% -------------------------------------------------------------
%               Source Verse
% -------------------------------------------------------------

\clearpage \newpage
\section{Source Verse, Essential Verse, and superficial desugaring}
\fi % \ifeverything
\input{verse-source}


% -------------------------------------------------------------
%               Denotational semantics: leading version
% -------------------------------------------------------------

\section{Denotational semantics: Essential Verse, using SLS}
\input{densem-notation}
\input{densem-ess-sls}

% -----------> LS: commented out for now <------------
% \input{densem-ess-ls}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%     Rest of doc is commented out for now
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{comment}

% -------------------------------------------------------------
%               Mini to Essential: wrapping
% -------------------------------------------------------------

\clearpage \newpage
\section{Desugaring Essential Verse to Mini Verse: wrapping}
\input{ess-to-mini}

% -------------------------------------------------------------
%               Desugar to Core
% -------------------------------------------------------------

\clearpage \newpage
\section{Desugar Mini Verse to Core}
\input{mini-to-core}

% -------------------------------------------------------------
%               Rewrite and verification rules
% -------------------------------------------------------------

\clearpage \newpage
\section{Rewrite and verification rules: runtime and verification}
\input{verse-rules}

% -------------------------------------------------------------
%               Denotational semantics: notation
% -------------------------------------------------------------

\clearpage \newpage
\section{Denotational semantics: overview}

Design variants (all of them use notation in \Cref{fig:densem-sets-and-seqs} and \Cref{fig:densem-notation}):
\begin{itemize}
\item \textbf{(S-LS)}: semantics of Essential Verse, lists of sets of environments.
  \begin{itemize}
  \item $\semess{S}{u}{v}{|t|} : \seq{ \ENV }$
  \item \Cref{fig:s-ls}
  \end{itemize}

\item \textbf{(E-LS)}: expression-style, list of sets of values.
  \begin{itemize}
  \item $\semappsub{E}{e}{|e|} : \Env \rightarrow \seq{ \VAL }$
  \item \Cref{fig:e-lv1} and \Cref{fig:e-lv2}
  \end{itemize}

\item \textbf{(D-LS)}: dps-style, list of sets of environments.
  \begin{itemize}
  \item $\semapsub{E}{d}{|e|} r : \seq{ \ENV }$
  \item \Cref{fig:d-ls1} and \Cref{fig:d-ls2}
  \end{itemize}

\item \textbf{(D-SLS)}: dps-style, sets of lists of sets of environments.
  \begin{itemize}
  \item $\semapsub{E}{d}{|e|} r : \powersetnp{\seq{ \ENV }}$
  \item \Cref{fig:d-sls1} and \Cref{fig:d-sls2}
  \end{itemize}
\end{itemize}

% -------------------------------------------------------------
%               E-LS: expression-style sequences of sets
% -------------------------------------------------------------

\clearpage \newpage
\section{Denotational semantics: expression style}
\input{densem-expr-style}

% -------------------------------------------------------------
%               D-LS: DPS-style: sequences of sets
% -------------------------------------------------------------

\clearpage \newpage
\section{Denotational semantics: destination passing style}
\input{densem-koen}

% -------------------------------------------------------------
%               D-SLS: DPS-style: sets of sequences of sets
% -------------------------------------------------------------
% \clearpage \newpage
% \ifeverything
% \section{Denotational semantics: DPS (sets of sequences of sets)}
% \fi % \ifeverything
% \input{densem-set-seq-set}

\ifeverything
% -------------------------------------------------------------
%               Denotational semantics: destination style
% -------------------------------------------------------------
\clearpage \newpage
\section{Denotational semantics: DPS (Jan style)}
\input{densem-dps-style}

% -------------------------------------------------------------
%               Denotational semantics: plan A
% -------------------------------------------------------------

\clearpage \newpage
\section{Denotational semantics: Plan A}
\input{verse-densem-planA}
\fi % \ifeverything

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%     End of commented out for now
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\end{comment}

\end{document}
